### 2019-04-20 manuscript/main.tex
commenceth the TeXage

\documentclass[11pt]{article}

\usepackage{fullpage}
\usepackage[round,semicolon]{natbib}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bbm}

\usepackage{hyperref}

\DeclareMathOperator{\erf}{erf}

% NOTE
% one sentence per line for nice git diffs
% WD: I'm a fan of initialed source comments like this

\title{Demographic inference under the coalescent as an ill-posed nonlinear inverse problem}

\author{
W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}\\
% \small{$^\ast$ Equal contribution}
}

\begin{document}

\maketitle

\begin{abstract}
Abstraction
\end{abstract}

\section*{Introduction}\label{sec:intro}

\subsection*{The tale of two bananas}\label{sec:intro:bananas}

What do we need to say about \cite{Rosen2018-bb}? See section~\ref{sec:model:stopping}

\section*{Model}\label{sec:model}

\subsection*{P and K}\label{sec:model:PandK}

\cite{Polanski2003-ll}

\subsection*{Constant MLE}\label{sec:model:constant}

We initialize by fitting a constant population size.
According to WSD's scribbles, the MLE assuming $\eta(t) = \eta_0$ (constant) is $\hat \eta_0 = \frac{S}{2 r H_{n-1}}$, where $S$ is the number of segregating sites (the sum of the observed SFS vector), $r$ is the mutation rate per genome per generation, and $H_{n-1}$ is the $n$th harmonic number.
This was derived by using the well-known result (cited in Rosen et al.) that the expected SFS for a constant population is given by $\xi_i = \frac{2\eta_0}{i}$.
Then the likelihood for $\eta_0$ is a Poisson random field parameterized by the $\xi_i$.

\subsection*{Regularized loss as a penalized log-likelihood}\label{sec:model:penalization}

\subsection*{The coalescent horizon: tempora incognita (see what I did there? ;))}\label{sec:model:horizon}

The coalescent tree height determines the time scale within which the SFS contains demographic information.
The expected height for n samples and a constant history $\eta(t)=\eta_0$ is the sum of the independent exponentially-distributed intercoalescent expectations:
\[
\mathbb{E}[T_n] = \sum_{k=2}^n \frac{\eta_0}{\binom{k}{2}} = 2\eta_0\left(1-\frac{1}{n}\right),
\]
where the last equality follows by recognizing the telescoping series.
Likewise for the variance:
\[
\text{Var}[T_n] = \sum_{k=2}^n \left(\frac{\eta_0}{\binom{k}{2}}\right)^2 = \frac{4 \eta _0^2 \left(-6 n^2 \psi ^{(1)}(n)+\left(\left(\pi ^2-9\right) n+6\right) n+3\right)}{3 n^2},
\]
where $\psi ^{(1)}(n)$ is the derivative of the digamma function \url{https://en.wikipedia.org/wiki/Trigamma_function}.

\subsection*{Asymptotically constant boundary condition}\label{sec:model:boundary}

We must deal with the asymptotically constant boundary condition.
Standard regularizers blow up on the infinite epoch.
Let's use half Gaussian instead of Lebesgue measure on time to induce integrability: $\mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t,$
where $\tau$ is the characteristic time to asmptopia (the boundary of our time grid).
For example, a modified $L2$ would be
\[
R\left[\eta(t)\right] = \int_0^\infty \eta(t)^2 \mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\int_0^\infty \eta(t)^2 \exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t.
\]
So the discretized problem is expressed in terms of the error function $\erf(\cdot)$ if $\eta(t)$ is piecewise constant.
This will give extra (but finite) weight / penalty to the infinite epoch based on the survival function of $\mathrm{d}\mu(t)$.
We can tune how much weight the boundary epoch gets by tuning the width of the Gaussian.

\subsection*{Stopping criteria based on Rosen et al.\ Thm (8)}\label{sec:model:stopping}

Thm (8) of Rosen et al. \cite{Rosen2018-bb} tells us that we can always find a piecewise MLE.
But we're interested in finding smooth solutions that have the same likelihood as that piecewise MLE.
Perhaps this provides a stopping criteria in our penalized likelihood iterates.

\section*{Results}\label{sec:results}

Do we need results?

\section*{Discussion}\label{sec:discussion}

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}

### 2019-04-20 dement.py
refactoring optimization from notebook to `dement.py`, and much tidying

@@ -4,27 +4,27 @@
 import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson
-import matplotlib
-from matplotlib import pyplot as plt, rc_context
-from matplotlib import patheffects
-# plt.xkcd()
-# matplotlib.rcParams['path.effects'] = [patheffects.withStroke(linewidth=0, foreground="w")]
+from matplotlib import pyplot as plt
+from scipy.special import polygamma
+from scipy.optimize import minimize
+from typing import Tuple
 
-# matplotlib.rc('text', usetex=True)
-# matplotlib.rcParams['text.latex.preamble']=[r"\usepackage{amsmath}"]
 
 class DemEnt():
     '''
-    A class that implements the model of Rosen et al., but adds a Poisson random
-    field for generating the SFS from ξ
+    A class that implements the model of Rosen et al., but adds a Poisson
+    random field for generating the SFS from ξ
     '''
-    def __init__(self, n: int, t, y, mutation_rate=1, infinite=False):
+
+    def __init__(self, n: int, t: np.ndarray, y: np.ndarray, r: float = 1,
+                 infinite: bool = True):
         '''
-        n is number of sampled haplotypes
-        t is the time axis
-        The last epoch in t extends to infinity in Rosen, but we truncate if infinite=False
-        y is the vector of eta pieces
-        mutation_rate is per genome per generation
+        n: number of sampled haplotypes
+        t: The time axis. The last epoch in t extends to infinity in Rosen, but
+           we truncate if infinite=False
+        y: vector of eta pieces
+        r: mutation rate is per genome per generation (controls SFS noise)
+        infinite: extend to infinity with constant size (otherwise truncate)
         '''
         assert t[0] == 0
         assert not np.isinf(t[-1])
@@ -34,16 +34,20 @@ class DemEnt():
         self.infinite = infinite
         self.n = n
         self.t = t
-        self.y = y
-        self._binom_array = binom(np.arange(2, n + 1), 2) # 2 choose 2 to (n+1) choose 2
-        self.A = self._init_A(n, mutation_rate=mutation_rate)
-        self.s = np.diff(self.t)
-        self.simulate_sfs()
+        self.y_true = y
+        self.r = r
+        self._binom_array = binom(np.arange(2, n + 1), 2)
[…]

### 2019-04-20 dement.py
simplify vectorization

@@ -36,7 +36,7 @@ class DemEnt():
         self.t = t
         self.y_true = y
         self.r = r
-        self._binom_array = binom(np.arange(2, n + 1), 2)
+        self._binom_array = binom(np.arange(2, n + 1), 2)[:, np.newaxis]
         self.A = DemEnt.A(n)
         self._s = np.diff(self.t)
         self.simulate_sfs()  # this sets self.sfs
@@ -77,19 +77,19 @@ class DemEnt():
             # when using infinite domain, extend last point to infty
             x = np.concatenate((x, [0]))  # final exponential is zero
             y_diff = np.concatenate((y_diff, [0]))  # final diff is zero
-        k = len(y_diff)
-        M2 = np.tile(np.array([1 / self._binom_array]).transpose(), (1, k)) * \
-             np.cumprod((x[np.newaxis, :-1] ** self._binom_array[:, np.newaxis]),
-                        axis=1)
+        M2 = np.cumprod(x[np.newaxis, :-1] ** self._binom_array, axis=1) \
+            / self._binom_array
         c = M2.dot(y_diff)
         if not jacobian:
             return c
         raise NotImplementedError('Jacobian not implemented yet')
         # dM2dy = np.zeros((M2.shape[0], M2.shape[1], k))
         # for depth in range(k):
-        #     dM2dy[:, (depth + 1):, depth] = binom_array[:, np.newaxis] \
-        #       * (self.t[depth + 1] - self.t[depth]) / (y[depth] ** 2) * M2[:, (depth + 1):]
-        # J = np.tensordot(dM2dy, y_diff, ([1], [0])) + M2 @ (np.eye(k) - np.eye(k, k=-1))
+        #     dM2dy[:, (depth + 1):, depth] = binom_array \
+        #       * (self.t[depth + 1] - self.t[depth]) / (y[depth] ** 2) \
+        #       * M2[:, (depth + 1):]
+        # J = np.tensordot(dM2dy, y_diff, ([1], [0])) \
+        #     + M2 @ (np.eye(k) - np.eye(k, k=-1))
         # return c, J
 
     def xi(self, y: np.ndarray = None) -> np.ndarray:
@@ -196,13 +196,15 @@ class DemEnt():
             T_sigma = np.sqrt(T_var)
             # time indices where we're within +/- 2 sigma of the expectation
             horizon_zone = np.flatnonzero((T_exp - 2 * T_sigma < self.t[:-1])
[…]

### 2019-04-21 dement.py
parameterize penalty ramp, and some more tidying

@@ -11,8 +11,7 @@ from typing import Tuple
 
 
 class DemEnt():
-    '''
-    A class that implements the model of Rosen et al., but adds a Poisson
+    '''A class that implements the model of Rosen et al., but adds a Poisson
     random field for generating the SFS from ξ
     '''
 
@@ -43,8 +42,7 @@ class DemEnt():
 
     @staticmethod
     def A(n: int) -> np.ndarray:
-        '''
-        The A_n matrix of Polanski and Kimmel (2003) (equations 13–15)
+        '''The A_n matrix of Polanski and Kimmel (2003) (equations 13–15)
         Using notation of Rosen et al. (2018)
 
         n: number of sampled haplotypes
@@ -61,11 +59,11 @@ class DemEnt():
         return A
 
     def c(self, y: np.ndarray = None, jacobian: bool = False) -> np.ndarray:
-        '''
-        coalescence vector computed by eqn (3) of Rosen et al. (2018), and it's
-        Jacobian
+        '''coalescence vector computed by eqn (3) of Rosen et al. (2018), and
+        it's Jacobian
 
         y: η(t) values
+        jacobian: compute jacobian if True
         '''
         if y is None:
             y = self.y_true
@@ -93,8 +91,7 @@ class DemEnt():
         # return c, J
 
     def xi(self, y: np.ndarray = None) -> np.ndarray:
-        '''
-        The expected SFS vector as in Rosen et al., but with an explicit
+        '''The expected SFS vector as in Rosen et al., but with an explicit
         mutation rate that controls noise
 
         y: optional η(t) values, uses self.y_true if None
@@ -102,14 +99,11 @@ class DemEnt():
         return self.r * self.A.dot(self.c(y))
 
     def simulate_sfs(self) -> None:
-        '''
-        simulate a SFS under the Poisson random field model
-        '''
+        '''simulate a SFS under the Poisson random field model'''
         self.sfs = poisson.rvs(self.xi())
 
     def ell(self, y: np.ndarray = None) -> float:
-        '''
-        Poisson random field log-likelihood
+        '''Poisson random field log-likelihood
 
         y: η(t) values, use self.y_true if None
         '''
@@ -121,8 +115,7 @@ class DemEnt():
 
     @staticmethod
     def constant_MLE(n: int, S: int, r: float) -> float:
-        '''
-        MLE for constant demography (see TeX)
+        '''MLE for constant demography
 
[…]

### 2019-04-22 dement.py
ramp based on TMRCA CDF with constant approximation, cc #8

@@ -5,7 +5,6 @@ import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
-from scipy.special import polygamma
 from scipy.optimize import minimize
 from typing import Tuple
 
@@ -113,108 +112,92 @@ class DemEnt():
         ell = - xi.sum() + self.sfs.dot(np.log(xi))
         return ell
 
-    @staticmethod
-    def constant_MLE(n: int, S: int, r: float) -> float:
-        '''MLE for constant demography
-
-        n: number of sampled haplotypes
-        S: number of segregating sites (sum of SFS entries)
-        r: mutation rate per genome per generation
+    def constant_MLE(self) -> float:
+        '''MLE for constant demography. Updates the instance's self.y_inferred
+        based on the constant MLE
         '''
-        H = (1 / np.arange(1, n - 1)).sum()
-        return (S / 2 / H / r)
+        # number of segregating sites
+        S = self.sfs.sum()
+        # Harmonic number
+        H = (1 / np.arange(1, self.n - 1)).sum()
+        # constant MLE
+        self.y_inferred = (S / 2 / H / self.r) * np.ones(len(self.t) - 1)
 
-    @staticmethod
-    def coalescent_horizon(n: int, eta_0: float) -> Tuple[float, float]:
-        '''return the expectation and variance of the TMRCA of n samples from
-        a population of size eta_0
-
-        n: number of sampled haplotypes
-        eta_0: constant population size
+    def tmrca_cdf(self) -> np.ndarray:
+        '''return the CDF of the TMRCA of n samples from a population of size
+        eta_0, using eqn (3.39) from Wakeley, at each time in self.t.
         '''
-        T_exp = 2 * eta_0 * (1 - 1 / n)
-        T_var = 4 * (3 + n * (6 + n * (-9 + np.pi**2))
-                     - 6 * n ** 2 * polygamma(1, n)) \
-                  * eta_0 ** 2 / (3 * n ** 2)
-        return T_exp, T_var
+        if (np.diff(self.y_inferred) != 0).sum() > 0:
+            raise NotImplementedError('tmrca_cdf() only available for constant'
+                                      'demographies')
+        eta_0 = self.y_inferred[0]
+        with np.errstate(divide='ignore'):
+            binom_factors = self._binom_array / (self._binom_array - self._binom_array.T)
[…]

### 2019-04-22 dement.py
exception handling

@@ -24,11 +24,14 @@ class DemEnt():
         r: mutation rate is per genome per generation (controls SFS noise)
         infinite: extend to infinity with constant size (otherwise truncate)
         '''
-        assert t[0] == 0
-        assert not np.isinf(t[-1])
-        assert all(np.diff(t) > 0)
-        assert all(y > 0)
-        assert n > 1
+        if t[0] != 0:
+            raise ValueError('the first element of t must be 0')
+        if any(np.diff(t) <= 0) or np.isinf(t).sum():
+            raise ValueError('t must be monotonically increasing and finite')
+        if any(y <= 0) or np.isinf(t).sum():
+            raise ValueError('elements of y must be finite and positive')
+        if n < 2:
+            raise ValueError('n must be larger than 1')
         self.infinite = infinite
         self.n = n
         self.t = t
@@ -106,8 +109,8 @@ class DemEnt():
 
         y: η(t) values, use self.y_true if None
         '''
-        # we must have previously run the simulate_sfs method
-        assert hasattr(self, 'sfs')
+        if not hasattr(self, 'sfs'):
+            raise ValueError('use simulate_sfs() to generate data first')
         xi = self.xi(y)
         ell = - xi.sum() + self.sfs.dot(np.log(xi))
         return ell
@@ -132,7 +135,8 @@ class DemEnt():
                                       'demographies')
         eta_0 = self.y_inferred[0]
         with np.errstate(divide='ignore'):
-            binom_factors = self._binom_array / (self._binom_array - self._binom_array.T)
+            binom_factors = self._binom_array / (self._binom_array
+                                                 - self._binom_array.T)
         binom_factors[np.isinf(binom_factors)] = 1
         binom_prod = np.prod(binom_factors, axis=0)
 

### 2019-04-22 dement.py
adding KL model fit option, see #10

@@ -115,6 +115,20 @@ class DemEnt():
         ell = - xi.sum() + self.sfs.dot(np.log(xi))
         return ell
 
+    def d_kl(self, y: np.ndarray = None) -> float:
+        '''Kullback-Liebler divergence between normalized SFS and its
+        expectation
+
+        y: η(t) values, use self.y_true if None
+        '''
+        if not hasattr(self, 'sfs'):
+            raise ValueError('use simulate_sfs() to generate data first')
+        xi = self.xi(y)
+        xi_hat = xi / xi.sum()
+        sfs_hat = self.sfs / self.sfs.sum()
+        d = - sfs_hat.dot(np.log(xi_hat / sfs_hat))
+        return d
+
     def constant_MLE(self) -> float:
         '''MLE for constant demography. Updates the instance's self.y_inferred
         based on the constant MLE
@@ -143,7 +157,7 @@ class DemEnt():
         return 1 - np.exp(-self._binom_array * self.t / eta_0).T @ binom_prod
 
     def loss(self, y, y_prior, lambda_prior: float = 0,
-             lambda_diff: np.ndarray = None) -> float:
+             lambda_diff: np.ndarray = None, fit_func: str = 'prf') -> float:
         '''negative log likelihood (Poisson random field) and regularizations
         on divergence from a prior and on the derivative
 
@@ -151,6 +165,7 @@ class DemEnt():
         y_prior: list of η(t) values
         lambda_prior: regularization strength on Bregman divergence from prior
         lambda_diff: regularization strength on derivative at each time
+        fit_func: 'prf' for Poisson random field, or 'kl' for Kullback-Liebler
         '''
         # generalized KL divergence (a Bregman divergence) from y_prior
         # we need the factor of self._s to encode the Lebesgue measure,
@@ -165,10 +180,18 @@ class DemEnt():
             lambda_diff = np.zeros(len(self.t))
         R_diff = (lambda_diff[:-2] * np.diff(y)**2).sum()
 
-        return - self.ell(y) + R_prior + R_diff
+        if fit_func == 'prf':
+            misfit = - self.ell(y)
+        elif fit_func == 'kl':
+            misfit = self.d_kl(y)
+        else:
+            raise ValueError('fit_func {} not recognized'.format(fit_func))
+
+        return misfit + R_prior + R_diff
 
     def invert(self, iterates: int = 1, lambda_prior: float = 1,
-               lambda_diff_min: float = 1, lambda_diff_max: float = None):
[…]

### 2019-04-23 dement.py
hyperparameter sweeps

@@ -191,7 +191,8 @@ class DemEnt():
 
     def invert(self, iterates: int = 1, lambda_prior: float = 1,
                lambda_diff_min: float = 1, lambda_diff_max: float = None,
-               fit_func: str = 'prf'):
+               fit_func: str = 'prf',
+               plot_initial: bool = False):
         '''infer the demography given the simulated sfs
 
         iterates: number of outer iterates
@@ -200,11 +201,13 @@ class DemEnt():
         lambda_diff_max: ramp derivative penalty to this value as we approach
                           coalescent horizon (no ramp if None)
         fit_func: 'prf' for Poisson random field, or 'kl' for Kullback-Liebler
+        plot_initial: if True show the initial constant demography MLE
         '''
         # Initialize with a MLE constant demography
         self.constant_MLE()
-        print('constant MLE initialization')
-        self.plot()
+        if plot_initial:
+            print('constant MLE initialization')
+            self.plot()
 
         # derivative penalty ramp based on TMRCA CDF
         if lambda_diff_max is None:
@@ -264,9 +267,9 @@ class DemEnt():
         axes[0].set_xlabel('$t$')
         axes[0].set_ylabel('$\\eta(t)$')
         axes[0].legend()
-        axes[0].legend(loc='upper center')
+        axes[0].legend(loc=0)
         axes[0].set_ylim([0, None])
-        axes[0].set_xscale('log')
+        axes[0].set_xscale('symlog')
         # axes[0].set_yscale('log')
 
         if hasattr(self, 'y_inferred'):
@@ -285,7 +288,7 @@ class DemEnt():
         axes[1].set_xscale('log')
         axes[1].set_yscale('symlog')
         axes[1].legend()
-        axes[1].legend(loc='upper right')
+        axes[1].legend(loc=0)
 
         plt.tight_layout()
         plt.show()

### 2019-04-23 manuscript/main.tex
ramp based on TMRCA CDF with constant approximation, cc #8
Kelley's equation

@@
 W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
-\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}\\
+\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}
 % \small{$^\ast$ Equal contribution}
@@
 
-We must deal with the asymptotically constant boundary condition.
-Standard regularizers blow up on the infinite epoch.
-Let's use half Gaussian instead of Lebesgue measure on time to induce integrability: $\mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t,$
-where $\tau$ is the characteristic time to asmptopia (the boundary of our time grid).
-For example, a modified $L2$ would be
+% Kelley's eqn
+In real data (or realistic simulations) the time region where the SFS has lost information is determined by the expected number of coalescent trees $\rho\ell$, where $\rho$ is the recombination rate and $\ell$ is the genome size.
+For $t$ such that the expected number of trees that have $T_{\text{MRCA}} > t$ becomes less that one, the SFS does not contain information beyond $t$.
+To wit,
 \[
-R\left[\eta(t)\right] = \int_0^\infty \eta(t)^2 \mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\int_0^\infty \eta(t)^2 \exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t.
+\mathbb{P}\left(T_{\text{MRCA}}>t\right) < \frac{1}{\rho \ell}.
 \]
-So the discretized problem is expressed in terms of the error function $\erf(\cdot)$ if $\eta(t)$ is piecewise constant.
-This will give extra (but finite) weight / penalty to the infinite epoch based on the survival function of $\mathrm{d}\mu(t)$.
-We can tune how much weight the boundary epoch gets by tuning the width of the Gaussian.
+The P\&K model doesn't model finite genome size, so it makes sense to use the CDF of $T_{\text{MRCA}}$ in that setting.
+
+% We must deal with the asymptotically constant boundary condition.
+% Standard regularizers blow up on the infinite epoch.
+% Let's use half Gaussian instead of Lebesgue measure on time to induce integrability: $\mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t,$
+% where $\tau$ is the characteristic time to asmptopia (the boundary of our time grid).
+% For example, a modified $L2$ would be
+% \[
+% R\left[\eta(t)\right] = \int_0^\infty \eta(t)^2 \mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\int_0^\infty \eta(t)^2 \exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t.
+% \]
+% So the discretized problem is expressed in terms of the error function $\erf(\cdot)$ if $\eta(t)$ is piecewise constant.
+% This will give extra (but finite) weight / penalty to the infinite epoch based on the survival function of $\mathrm{d}\mu(t)$.
+% We can tune how much weight the boundary epoch gets by tuning the width of the Gaussian.
 

### 2019-08-13 manuscript/main.tex
beginning our pivot to mu(t) inference by writing up some theory

--- 2019-04-20 manuscript/main.tex
@@
 
-\title{Demographic inference under the coalescent as an ill-posed nonlinear inverse problem}
+\title{Coalescent inference of mutation spectrum histories from nucleotide context-resolved sample frequency spectra}
 
 \author{
-W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
-\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}\\
+W. DeWitt$^{1}$, K. Harris (if he's into it?)$^{2}$, and K. Harris$^{1}$\\
+\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}
 % \small{$^\ast$ Equal contribution}
@@
 \begin{abstract}
-Abstraction
+
+Single nucleotide variant (SNV) frequencies partitioned according to triplet nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
+The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
+We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the triplet-resolved sample frequency spectrum (3-SFS).
+We formulate inference of mutation spectrum histories from the 3-SFS as a (linear?) inverse problem.
+
 \end{abstract}
@@
 
-\subsection*{The tale of two bananas}\label{sec:intro:bananas}
+Kelley's paper \cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and also used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
+We want to do some coalescent theory to formulate how the expected 3-SFS depends on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation rates as a function of time into the past).
+This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{?}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
 
-What do we need to say about \cite{Rosen2018-bb}? See section~\ref{sec:model:stopping}
+The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
+We will retrace the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
+We will make use of important results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} which facilitate numerical computation under this model, and formulate the expected 3-SFS under given demographic and mutation spectrum histories following the notation of \cite{Rosen2018-bb} for the expected SFS with a constant mutation process.
 
@@
 
-\subsection*{P and K}\label{sec:model:PandK}
+Suppose $n$ haplotypes are sampled in the present, and let rvs $T = T_2,\dots,T_n$ denote the coalescent times measured retrospectively from the present (i.e.
+$T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
+As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
+This is tantamount to an (uncountably) infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
+Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) > 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present).
+A given line in the coalescent tree then acquires mutations on a genomic region $M\subset(0,1)$ at rate $|M|\mu(t)$, where $|M|$ is the Lebesgue measure of $M$ (e.g.\ if $M$ is a subinterval, $|M|$ is the length of $M$).
 
-\cite{Polanski2003-ll}
+Let $C_h$ denote the event that there is a mutation with label $U\in(x,x+h)\subseteq(0,1)$ that subtends $b$ haplotypes in the sample, and let $I_k$ denote the event that this mutation arose at a time when the sample had $k$ ancestral lines.
+For small $h$ we then have
+\begin{align*}
+\mathbbm{P}(C_h\mid T) &= \sum_{k=2}^n \mathbbm{P}(C_h, I_k\mid T) = \sum_{k=2}^n p_{n,k}(b)\mathbbm{P}(I_k, U\in(x,x+h)\mid T)\\
+&= \sum_{k=2}^n p_{n,k}(b)\left(kh\int_{T_{k-1}}^{T_k}\mu(t)dt + O(h)\right),
+\end{align*}
+where $p_{n,k}(b) = \binom{n-b-1}{k-2}/\binom{n-1}{k-1}$ is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9).
+Marginalizing $T$ gives
+\begin{align*}
+\mathbbm{P}(C_h) &\simeq h\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right] = h\sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right),
+\end{align*}
+where $\nu(t)\equiv\int\mu(t)dt$.
+% Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
+% With $\pi_k(t_k)$ denoting the marginal density of $T_k$, define rv $X_k \equiv \nu(T_k)$, which has density $f_{X_k}(x_k) = \pi_k(\nu^{-1}(x_k))$.
+Using eqns 5-8 from \cite{Polanski2003-kg} we can write
+\begin{align*}
+\mathbbm{E}\left[\nu(T_{k})\right] &= \sum_{j=k}^n A_j^k\int_0^\infty\nu(t)q_j(t)dt
+\end{align*}
+where
+\begin{align*}
+A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
+A_n^n &\equiv 1,\\
+q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
+\end{align*}
+and $\eta(t)$ is the haploid effective population size history.
+Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present.
+This can be made explicit by defining the inhomogeneous Poisson process intensity function $\lambda_j(t) = \binom{j}{2}/\eta(t)$.
 
-\subsection*{Constant MLE}\label{sec:model:constant}
+The $b$-th component of the expected SFS is then
 
-We initialize by fitting a constant population size.
-According to WSD's scribbles, the MLE assuming $\eta(t) = \eta_0$ (constant) is $\hat \eta_0 = \frac{S}{2 r H_{n-1}}$, where $S$ is the number of segregating sites (the sum of the observed SFS vector), $r$ is the mutation rate per genome per generation, and $H_{n-1}$ is the $n$th harmonic number.
-This was derived by using the well-known result (cited in Rosen et al.) that the expected SFS for a constant population is given by $\xi_i = \frac{2\eta_0}{i}$.
-Then the likelihood for $\eta_0$ is a Poisson random field parameterized by the $\xi_i$.
+\begin{align*}
+\xi_b &= \sum_{k=2}^n p_{n,k}(b) k \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right)\\
+&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^{n}B_j^k\int_0^\infty\nu(t)q_j(t)dt
+\end{align*}
+where
+\[
+B_j^k \equiv
+\begin{cases}
+A_{k-1}^{k-1}, j=k-1\\
+A_j^k - A_j^{k-1}, k\le j\le n\\
+\end{cases}
+\]
+Exchanging the order of integration and summation yields a manifestly linear expression for the expected SFS as a function of the integrated mutation intensity $\nu(t)$:
+\[
+\xi_b = \left<\psi_b, \nu\right>,
+\]
+where
+\begin{align*}
+\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k q_j(t).
+\end{align*}
 
-\subsection*{Regularized loss as a penalized log-likelihood}\label{sec:model:penalization}
-
-\subsection*{The coalescent horizon: tempora incognita (see what I did there? ;))}\label{sec:model:horizon}
-
-The coalescent tree height determines the time scale within which the SFS contains demographic information.
-The expected height for n samples and a constant history $\eta(t)=\eta_0$ is the sum of the independent exponentially-distributed intercoalescent expectations:
-\[
-\mathbb{E}[T_n] = \sum_{k=2}^n \frac{\eta_0}{\binom{k}{2}} = 2\eta_0\left(1-\frac{1}{n}\right),
-\]
-where the last equality follows by recognizing the telescoping series.
-Likewise for the variance:
-\[
-\text{Var}[T_n] = \sum_{k=2}^n \left(\frac{\eta_0}{\binom{k}{2}}\right)^2 = \frac{4 \eta _0^2 \left(-6 n^2 \psi ^{(1)}(n)+\left(\left(\pi ^2-9\right) n+6\right) n+3\right)}{3 n^2},
-\]
-where $\psi ^{(1)}(n)$ is the derivative of the digamma function \url{https://en.wikipedia.org/wiki/Trigamma_function}.
-
-\subsection*{Asymptotically constant boundary condition}\label{sec:model:boundary}
-
-We must deal with the asymptotically constant boundary condition.
-Standard regularizers blow up on the infinite epoch.
-Let's use half Gaussian instead of Lebesgue measure on time to induce integrability: $\mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t,$
-where $\tau$ is the characteristic time to asmptopia (the boundary of our time grid).
-For example, a modified $L2$ would be
-\[
-R\left[\eta(t)\right] = \int_0^\infty \eta(t)^2 \mathrm{d}\mu(t) = \frac{\sqrt{2}}{\sqrt{\pi}\tau}\int_0^\infty \eta(t)^2 \exp\left(-\frac{1}{2}\left(\frac{t}{\tau}\right)^2\right)\mathrm{d}t.
-\]
-So the discretized problem is expressed in terms of the error function $\erf(\cdot)$ if $\eta(t)$ is piecewise constant.
-This will give extra (but finite) weight / penalty to the infinite epoch based on the survival function of $\mathrm{d}\mu(t)$.
-We can tune how much weight the boundary epoch gets by tuning the width of the Gaussian.
-
-\subsection*{Stopping criteria based on Rosen et al.\ Thm (8)}\label{sec:model:stopping}
-
-Thm (8) of Rosen et al. \cite{Rosen2018-bb} tells us that we can always find a piecewise MLE.
-But we're interested in finding smooth solutions that have the same likelihood as that piecewise MLE.
-Perhaps this provides a stopping criteria in our penalized likelihood iterates.
+% DEPRECATED:
+%
+% \subsection*{Constant MLE}\label{sec:model:constant}
+%
+% We initialize by fitting a constant population size.
+% According to WSD's scribbles, the MLE assuming $\eta(t) = \eta_0$ (constant) is $\hat \eta_0 = \frac{S}{2 r H_{n-1}}$, where $S$ is the number of segregating sites (the sum of the observed SFS vector), $r$ is the mutation rate per genome per generation, and $H_{n-1}$ is the $n$th harmonic number.
+% This was derived by using the well-known result (cited in Rosen et al.) that the expected SFS for a constant population is given by $\xi_i = \frac{2\eta_0}{i}$.
+% Then the likelihood for $\eta_0$ is a Poisson random field parameterized by the $\xi_i$.
+%
+%
+% \subsection*{The coalescent horizon: tempora incognita (see what I did there? ;))}\label{sec:model:horizon}
+%
+% The coalescent tree height determines the time scale within which the SFS contains demographic information.
+% The expected height for n samples and a constant history $\eta(t)=\eta_0$ is the sum of the independent exponentially-distributed intercoalescent expectations:
+% \[
+% \mathbb{E}[T_n] = \sum_{k=2}^n \frac{\eta_0}{\binom{k}{2}} = 2\eta_0\left(1-\frac{1}{n}\right),
+% \]
+% where the last equality follows by recognizing the telescoping series.
+% Likewise for the variance:
+% \[
+% \text{Var}[T_n] = \sum_{k=2}^n \left(\frac{\eta_0}{\binom{k}{2}}\right)^2 = \frac{4 \eta _0^2 \left(-6 n^2 \psi ^{(1)}(n)+\left(\left(\pi ^2-9\right) n+6\right) n+3\right)}{3 n^2},
+% \]
+% where $\psi ^{(1)}(n)$ is the derivative of the digamma function \url{https://en.wikipedia.org/wiki/Trigamma_function}.
+%
+%
+% \subsection*{Asymptotically constant boundary condition}\label{sec:model:boundary}
+%
+% % Kelley's eqn
+% In real data (or realistic simulations) the time region where the SFS has lost information is determined by the expected number of coalescent trees $\rho\ell$, where $\rho$ is the recombination rate and $\ell$ is the genome size.
+% For $t$ such that the expected number of trees that have $T_{\text{MRCA}} > t$ becomes less that one, the SFS does not contain information beyond $t$.
+% To wit,
+% \[
+% \mathbb{P}\left(T_{\text{MRCA}}>t\right) < \frac{1}{\rho \ell}.
+% \]
+% The P\&K model doesn't model finite genome size, so it makes sense to use the CDF of $T_{\text{MRCA}}$ in that setting.
 
@@
 
-Do we need results?
+Todo:
+\begin{itemize}
+\item repeat 1KG analysis like in \cite{Harris2017-fw}, see if we recapitulate Kelley's simulation-based pulse results
+\item cluster triplet time series to see if we pull in minor components. Dynamic time warping?
+\end{itemize}
 

### 2019-08-14 manuscript/main.tex
linear indeed
cite Tennessen

@@
 We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the triplet-resolved sample frequency spectrum (3-SFS).
-We formulate inference of mutation spectrum histories from the 3-SFS as a (linear?) inverse problem.
+We formulate inference of mutation spectrum histories from the 3-SFS as a linear inverse problem.
 
@@
 We want to do some coalescent theory to formulate how the expected 3-SFS depends on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation rates as a function of time into the past).
-This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{?}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
+This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
 

### 2019-08-15 manuscript/main.tex
less wordy title
linear model in mu (not nu)

@@
 
-\title{Coalescent inference of mutation spectrum histories from nucleotide context-resolved sample frequency spectra}
+\title{Coalescent inference of mutation spectrum histories from sample frequency spectra}
 
@@
 \end{align*}
-where $\nu(t)\equiv\int\mu(t)dt$.
+where $\nu(t)\equiv\int_0^t\mu(t)dt$.
 % Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
@@
 \begin{cases}
-A_{k-1}^{k-1}, j=k-1\\
-A_j^k - A_j^{k-1}, k\le j\le n\\
+A_{k-1}^{k-1},\ j=k-1\\
+A_j^k - A_j^{k-1},\ k\le j\le n\\
 \end{cases}
 \]
-Exchanging the order of integration and summation yields a manifestly linear expression for the expected SFS as a function of the integrated mutation intensity $\nu(t)$:
+Using the definition of $\nu$ and exchanging integration order gives
+\begin{align*}
+\int_0^\infty\nu(t)q_j(t)dt &= \int_0^\infty \left(\int_0^t\mu(x)dx\right) q_j(t)dt\\
+&= \int_0^\infty \int_0^\infty 1_{[0<x<t]}(x)\mu(x)dx\ q_j(t)dt\\
+&= \int_0^\infty \left(\int_0^\infty 1_{[0<x<t]}(x)q_j(t)dt\right)\mu(x) dx\\
+&= \int_0^\infty \left(1-\int_0^x q_j(t)dt\right)\mu(x) dx\\
+&= \int_0^\infty \left(1-Q_j(t)\right)\mu(x) dx,
+\end{align*}
+where $Q_j$ is the distribution function corresponding to density $q_j$.
+
+We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
 \[
-\xi_b = \left<\psi_b, \nu\right>,
+\xi_b = \left<\psi_b, \mu\right>,
 \]
@@
 \begin{align*}
-\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k q_j(t).
+\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k \left(1-Q_j(t)\right).
 \end{align*}

### 2019-08-16 manuscript/main.tex
much math tidying

@@
 This is tantamount to an (uncountably) infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
-Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) > 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present).
-A given line in the coalescent tree then acquires mutations on a genomic region $M\subset(0,1)$ at rate $|M|\mu(t)$, where $|M|$ is the Lebesgue measure of $M$ (e.g.\ if $M$ is a subinterval, $|M|$ is the length of $M$).
+Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) \ge 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
+A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
+The following argument can be generalized to a nonuniform distribution of mutation intensity over the genome without changing the results.
 
-Let $C_h$ denote the event that there is a mutation with label $U\in(x,x+h)\subseteq(0,1)$ that subtends $b$ haplotypes in the sample, and let $I_k$ denote the event that this mutation arose at a time when the sample had $k$ ancestral lines.
-For small $h$ we then have
+Let rv $M_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
+Let rv $M_{\delta x, b, k}$ denote the event that this mutation occurred at a time when the sample had $k$ ancestral lines (during the $k$th intercoalescent time interval).
+For small $\delta x$ and finite $\mu(t)$ we have
 \begin{align*}
-\mathbbm{P}(C_h\mid T) &= \sum_{k=2}^n \mathbbm{P}(C_h, I_k\mid T) = \sum_{k=2}^n p_{n,k}(b)\mathbbm{P}(I_k, U\in(x,x+h)\mid T)\\
-&= \sum_{k=2}^n p_{n,k}(b)\left(kh\int_{T_{k-1}}^{T_k}\mu(t)dt + O(h)\right),
+\mathbbm{P}(M_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(M_{\delta x, b, k}\mid T)\\
+&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{T_{k-1}}^{T_k}\mu(t)dt + O(\delta x)\right),
 \end{align*}
-where $p_{n,k}(b) = \binom{n-b-1}{k-2}/\binom{n-1}{k-1}$ is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9).
+where
+\begin{equation}
+\label{eqn:p}
+p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
+\end{equation}
+is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parenthesis is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
 Marginalizing $T$ gives
 \begin{align*}
-\mathbbm{P}(C_h) &\simeq h\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right] = h\sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right),
+\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right].
 \end{align*}
-where $\nu(t)\equiv\int_0^t\mu(t)dt$.
-% Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
-% With $\pi_k(t_k)$ denoting the marginal density of $T_k$, define rv $X_k \equiv \nu(T_k)$, which has density $f_{X_k}(x_k) = \pi_k(\nu^{-1}(x_k))$.
-Using eqns 5-8 from \cite{Polanski2003-kg} we can write
-\begin{align*}
-\mathbbm{E}\left[\nu(T_{k})\right] &= \sum_{j=k}^n A_j^k\int_0^\infty\nu(t)q_j(t)dt
-\end{align*}
+For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
+\begin{align}
+\label{eqn:xi}
+\xi_b &= \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right]\nonumber\\
+&= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \mathbbm{E}_{T_{k-1}}\left[\int_0^{T_{k-1}}\mu(t)dt\right]\right)
+\end{align}
+
+\cite{Polanski2003-kg} (eqns 5-8) give the marginal density for the coalescent time $T_k$ as
+\begin{equation}
+\label{eqn:pi}
+\pi_k(t_k) = \sum_{j=k}^n A_j^k q_j(t_k)
+\end{equation}
 where
@@
 and $\eta(t)$ is the haploid effective population size history.
-Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present.
-This can be made explicit by defining the inhomogeneous Poisson process intensity function $\lambda_j(t) = \binom{j}{2}/\eta(t)$.
+Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present, with inhomogeneous Poisson intensity function $\binom{j}{2}/\eta(t)$.
 
-The $b$-th component of the expected SFS is then
-
-\begin{align*}
-\xi_b &= \sum_{k=2}^n p_{n,k}(b) k \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right)\\
-&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^{n}B_j^k\int_0^\infty\nu(t)q_j(t)dt
-\end{align*}
+The expectations in \eqref{eqn:xi} can be expressed using \eqref{eqn:pi} as
+\begin{align}
+\label{eqn:exp}
+\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] &= \int_0^\infty\pi_k(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
+&= \sum_{j=k}^n A_j^k\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
+&= \sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt
+\end{align}
+where in the last line we've exchanged integration order and defined the inhomogeneous Poisson survival function
+\begin{equation}
+\label{eqn:r}
+r_j(t) \equiv \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]
+\end{equation}
+corresponding to density $q_j(t)$.
+Using \eqref{eqn:exp} in \eqref{eqn:xi} gives
+\begin{align}
+\label{eqn:xi2}
+\xi_b &= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt - \sum_{j=k-1}^n A_j^{k-1}\int_0^\infty r_j(t)\mu(t)dt\right)\nonumber\\
+&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^n B_j^k\int_0^\infty r_j(t)\mu(t)dt
+\end{align}
 where
@@
 \]
-Using the definition of $\nu$ and exchanging integration order gives
-\begin{align*}
-\int_0^\infty\nu(t)q_j(t)dt &= \int_0^\infty \left(\int_0^t\mu(x)dx\right) q_j(t)dt\\
-&= \int_0^\infty \int_0^\infty 1_{[0<x<t]}(x)\mu(x)dx\ q_j(t)dt\\
-&= \int_0^\infty \left(\int_0^\infty 1_{[0<x<t]}(x)q_j(t)dt\right)\mu(x) dx\\
-&= \int_0^\infty \left(1-\int_0^x q_j(t)dt\right)\mu(x) dx\\
-&= \int_0^\infty \left(1-Q_j(t)\right)\mu(x) dx,
-\end{align*}
-where $Q_j$ is the distribution function corresponding to density $q_j$.
 
@@
 \begin{align*}
-\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k \left(1-Q_j(t)\right).
+\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k r_j(t).
 \end{align*}

### 2019-08-16 manuscript/main.tex
fixing some integration limit bugs

@@
 
-Kelley's paper \cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and also used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
-We want to do some coalescent theory to formulate how the expected 3-SFS depends on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation rates as a function of time into the past).
+\cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
+We develop coalescent theory for the dependence of the expected 3-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation rates as a function of time into the past).
 This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
@@
 Let rv $M_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
-Let rv $M_{\delta x, b, k}$ denote the event that this mutation occurred at a time when the sample had $k$ ancestral lines (during the $k$th intercoalescent time interval).
+Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
+Let rv $M_{\delta x, b, k}$ denote the event that the mutation $M_{\delta x, b}$ occurred during $I_k$.
 For small $\delta x$ and finite $\mu(t)$ we have
@@
 \mathbbm{P}(M_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(M_{\delta x, b, k}\mid T)\\
-&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{T_{k-1}}^{T_k}\mu(t)dt + O(\delta x)\right),
+&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O(\delta x)\right),
 \end{align*}
@@
 \begin{align*}
-\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right].
+\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
 \end{align*}
 For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
+\[
+\xi_b = \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
+\]
+Using the explicit bounds of the coalescent intervals gives
 \begin{align}
 \label{eqn:xi}
-\xi_b &= \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right]\nonumber\\
-&= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \mathbbm{E}_{T_{k-1}}\left[\int_0^{T_{k-1}}\mu(t)dt\right]\right)
+\xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
+&= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=3}^{n} (k-1) p_{n,k-1}(b) \mathbbm{E}_{T_{k}}\left[\int_0^{T_k}\mu(t)dt\right]\nonumber\\
+&= \sum_{k=2}^n B_{b,k} \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right],
 \end{align}
+where
+\[
+B_{b,k}\equiv
+\begin{cases}
+k p_{n,k}(b),& k=2\\
+k p_{n,k}(b) - (k-1) p_{n,k-1}(b),& k > 2
+\end{cases}
+\]
 
@@
 \label{eqn:pi}
-\pi_k(t_k) = \sum_{j=k}^n A_j^k q_j(t_k)
+\pi_k(t_k) = \sum_{j=k}^n A_{k,j} q_j(t_k)
 \end{equation}
@@
 \begin{align*}
-A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
-A_n^n &\equiv 1,\\
+A_{k,j} &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
+A_{n,n} &\equiv 1,\\
 q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
@@
 \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] &= \int_0^\infty\pi_k(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
-&= \sum_{j=k}^n A_j^k\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
-&= \sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt
+&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
+&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^\infty 1_{[0<t<t_k]}(t)\mu(t)dt dt_k\nonumber\\
+&= \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt
 \end{align}
@@
 corresponding to density $q_j(t)$.
+
 Using \eqref{eqn:exp} in \eqref{eqn:xi} gives
@@
 \label{eqn:xi2}
-\xi_b &= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt - \sum_{j=k-1}^n A_j^{k-1}\int_0^\infty r_j(t)\mu(t)dt\right)\nonumber\\
-&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^n B_j^k\int_0^\infty r_j(t)\mu(t)dt
+\xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt.
 \end{align}
-where
-\[
-B_j^k \equiv
-\begin{cases}
-A_{k-1}^{k-1},\ j=k-1\\
-A_j^k - A_j^{k-1},\ k\le j\le n\\
-\end{cases}
-\]
 
@@
 \begin{align*}
-\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k r_j(t).
+\psi_b(t) &\equiv \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}r_j(t).
 \end{align*}

### 2019-08-16 manuscript/main.tex
writing up recursions for hypergeometric sums needed for our constant C

@@
 where
-\[
+\begin{equation}
+\label{eqn:B}
 B_{b,k}\equiv
@@
 \end{cases}
-\]
+\end{equation}
 
@@
 \label{eqn:xi2}
-\xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt.
+\xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt\nonumber\\
+&= \sum_{j=2}^n \left(\sum_{k=2}^j B_{b,k} A_{k,j}\right) \int_0^\infty r_j(t)\mu(t)dt.
 \end{align}
@@
 We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
+\begin{equation}
+\label{eqn:xivec}
+\boldsymbol\xi = C \int_0^\infty \boldsymbol r(t)\mu(t)dt,
+\end{equation}
+where the $(n-1)$ by $(n-1)$ matrix
 \[
-\xi_b = \left<\psi_b, \mu\right>,
+C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
 \]
-where
+is constant wrt $\mu$ \emph{and} $\eta$.
+
+We next develop an efficient recursive procedure for computing $C$.
+Using \ref{eqn:B}
 \begin{align*}
-\psi_b(t) &\equiv \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}r_j(t).
+C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
+&= W_{b,j} - W'_{b,j}
 \end{align*}
+Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqs 13-15), and we define
+\begin{equation}
+\label{eqn:W'}
+W'_{b,j} = \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
+\end{equation}
+\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
+\[
+A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
+\]
+so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W'} is a definite sum over geometric terms.
+Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following second-order recursion in $j$:
+\begin{align*}
+&W'_{b,2} = 0\\
+&W'_{b,3} = \frac{(-1)^{j-3} (2 j-1) (b-1)! (j+1)! n! (-b+n-1)! \binom{n-2}{b-1}}{(j-1) j (j-3)! (n-j)! (j+n-1)!}\\
+&W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
+\end{align*}
+
+This completes the discrete mathematics needed to frame mutation rate inference as a linear inverse problem.
+We next consider eqn \ref{eqn:xivec} as a problem in numerical analysis.
 

### 2019-08-17 manuscript/main.tex
fix base case for W'
piecewise constant discretization

@@
 
-\section*{Introduction}\label{sec:intro}
+\section{Introduction}\label{sec:intro}
 
@@
 
-\section*{Model}\label{sec:model}
+\section{Model}\label{sec:model}
+
+\subsection{The expected SFS given demographic and mutation rate histories}\label{sec:model:xi}
 
@@
 
-Let rv $M_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
+Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
 Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
-Let rv $M_{\delta x, b, k}$ denote the event that the mutation $M_{\delta x, b}$ occurred during $I_k$.
+Let rv $\mathcal{E}_{\delta x, b, k}$ denote the event that the mutation $\mathcal{E}_{\delta x, b}$ occurred during $I_k$.
 For small $\delta x$ and finite $\mu(t)$ we have
 \begin{align*}
-\mathbbm{P}(M_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(M_{\delta x, b, k}\mid T)\\
+\mathbbm{P}(\mathcal{E}_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(\mathcal{E}_{\delta x, b, k}\mid T)\\
 &= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O(\delta x)\right),
@@
 \begin{align*}
-\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
+\mathbbm{P}(\mathcal{E}_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
 \end{align*}
@@
 \[
-\xi_b = \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
+\xi_b = \int d\mathbbm{P}(\mathcal{E}_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
 \]
@@
 \label{eqn:xivec}
-\boldsymbol\xi = C \int_0^\infty \boldsymbol r(t)\mu(t)dt,
+\boldsymbol\xi = C \boldsymbol d,
 \end{equation}
@@
 \]
-is constant wrt $\mu$ \emph{and} $\eta$.
+is constant wrt $\mu$ \emph{and} $\eta$, and
+\begin{equation}
+\label{eqn:d}
+d_j \equiv \int_0^\infty r_j(t)\mu(t)dt
+\end{equation}
+has linear (nonlinear) dependence on $\mu$ ($\eta$).
+
+\subsection{Computing the elements of $C$}\label{sec:model:C}
 
@@
 \end{align*}
-Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqs 13-15), and we define
+Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqns 13-15), and we define $W'$ by
 \begin{equation}
@@
 &W'_{b,2} = 0\\
-&W'_{b,3} = \frac{(-1)^{j-3} (2 j-1) (b-1)! (j+1)! n! (-b+n-1)! \binom{n-2}{b-1}}{(j-1) j (j-3)! (n-j)! (j+n-1)!}\\
+&W'_{b,3} = \frac{20 (n-2)}{(n+1)(n+2)}\\
 &W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
 \end{align*}
+This completes the framing of mutation rate inference as a linear inverse problem.
 
-This completes the discrete mathematics needed to frame mutation rate inference as a linear inverse problem.
-We next consider eqn \ref{eqn:xivec} as a problem in numerical analysis.
+\subsection{Finite dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:model:pcsws}
+
+Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
+\begin{equation}
+\label{eqn:d2}
+d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
+\end{equation}
+where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
+We consider piecewise constant $\eta$ and $\mu$ on $\ell$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{\ell-1}, t_\ell)$ where $0=t_0 < t_1 < \dots < t_{\ell-1} < t_\ell=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
+Let $(y_1,\dots,y_\ell)$ denote the constant population size $eta(t)$ during each epoch, and let $(z_1,\dots,z_\ell)$ denote the constant mutation rate $mu(t)$ during each epoch.
+Let $x_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_\ell$, and $x_0\equiv 1$.
+With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
+\begin{equation}
+\label{eqn:d3}
+\boldsymbol d = M(\boldsymbol y) \boldsymbol z
+\end{equation}
+where
+\begin{equation}
+\label{eqn:M}
+M(\boldsymbol y) \equiv
+\begin{bmatrix}
+1 &             &        &                       \\
+  & \frac{1}{3} &        &                       \\
+  &             & \ddots &                       \\
+  &             &        & \frac{1}{\binom{n}{2}}
+\end{bmatrix}
+\begin{bmatrix}
+1       & x_1                & \hdots & \prod_{i=1}^{\ell-1}x_i               \\
+1       & x_1^3              & \hdots & \prod_{i=1}^{\ell-1}x_i^3             \\
+\vdots  & \vdots             & \ddots & \vdots                                \\
+1       & x_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{\ell-1}x_i^{\binom{n}{2}}
+\end{bmatrix}
+\begin{bmatrix}
+y_1  &      &        &             &       \\
+-y_1 & y_2  &        &             &       \\
+     & -y_2 & y_3    &             &       \\
+     &      & \ddots & \ddots      &       \\
+     &      &        & -y_{\ell-1} & y_\ell
+\end{bmatrix}.
+\end{equation}
+Note that the $(n-1)$ by $\ell$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $x_l$ are nonlinear functions of $\boldsymbol y$.
+
+Combining \ref{eqn:d3} with \ref{eqn:xivec} gives the discretized inverse problem
+\begin{equation}
+\label{eqn:xivec2}
+\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z.
+\end{equation}
 
@@
 
-\section*{Results}\label{sec:results}
+\section{Results}\label{sec:results}
 
@@
 
-\section*{Discussion}\label{sec:discussion}
+\section{Discussion}\label{sec:discussion}
 

### 2019-08-19 manuscript/main.tex
fairly complete writeup of methods, including regularization

@@
 \DeclareMathOperator{\erf}{erf}
+\DeclareMathOperator{\lngamma}{\ln\!\Gamma}
 
@@
 \end{equation}
-where the $(n-1)$ by $(n-1)$ matrix
+where the $(n-1)\times(n-1)$ matrix
 \[
@@
 C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
-&= W_{b,j} - W'_{b,j}
+&= W_{b,j}^{(1)} - W_{b,j}^{(2)},
 \end{align*}
-Where $W$ is defined recursively in \cite{Polanski2003-ll} (eqns 13-15), and we define $W'$ by
-\begin{equation}
-\label{eqn:W'}
-W'_{b,j} = \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
-\end{equation}
+where
+\begin{align}
+\label{eqn:W1}
+W_{b,j}^{(1)} &\equiv \sum_{k=2}^j k p_{n,k}(b) A_{k,j}\\
+\label{eqn:W2}
+W_{b,j}^{(2)} &\equiv \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
+\end{align}
 \cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
@@
 \]
-so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W'} is a definite sum over geometric terms.
-Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following second-order recursion in $j$:
+so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W1} and \ref{eqn:W2} are definite sums over geometric terms.
+Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following procedurally generated second-order recursions in $j$:
 \begin{align*}
-&W'_{b,2} = 0\\
-&W'_{b,3} = \frac{20 (n-2)}{(n+1)(n+2)}\\
-&W'_{b,j+2} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W'_{b,j}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W'_{b,j+1}\right)
+&W_{b,2}^{(1)} = \frac{6}{(n+1)}\\
+&W_{b,3}^{(1)} = \frac{10(5n-6b-4)}{(n+2)(n+1)}\\
+&W_{b,j+2}^{(1)} = -\frac{(2 j+3) \left(-(2 j-1) W_{b,j+1}^{(1)}  \left(2 j (j+1) \left(b^2 \left(j^2+j-2\right)-6 b-j (j+1)-2\right)-j (j+1) n \left(3 b \left(j^2+j+2\right)+j^2+j-2\right)+\left(j (j+1) \left(j^2+j+6\right)+4\right) n^2+4 n\right)-(j-1) (j+1)^2 (j-n) W_{b,j}^{(1)}  (4 (n+1)-j (j+2) (b-n-1))\right)}{j^2 (j+2) (2 j-1) (j+n+1) \left(-b j^2+b+\left(j^2+3\right) (n+1)\right)}
+\end{align*}
+and
+\begin{align*}
+&W_{b,2}^{(2)} = 0\\
+&W_{b,3}^{(2)} = \frac{20 (n-2)}{(n+1)(n+2)}\\
+&W_{b,j+2}^{(2)} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W_{b,j}^{(2)}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W_{b,j+1}^{(2)}\right)
 \end{align*}
@@
 where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
-We consider piecewise constant $\eta$ and $\mu$ on $\ell$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{\ell-1}, t_\ell)$ where $0=t_0 < t_1 < \dots < t_{\ell-1} < t_\ell=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
-Let $(y_1,\dots,y_\ell)$ denote the constant population size $eta(t)$ during each epoch, and let $(z_1,\dots,z_\ell)$ denote the constant mutation rate $mu(t)$ during each epoch.
-Let $x_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_\ell$, and $x_0\equiv 1$.
+We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
+We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite dimensional histories.
+Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
+Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_m$, and $u_0\equiv 1$.
 With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
@@
 \begin{bmatrix}
-1       & x_1                & \hdots & \prod_{i=1}^{\ell-1}x_i               \\
-1       & x_1^3              & \hdots & \prod_{i=1}^{\ell-1}x_i^3             \\
+1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
+1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
 \vdots  & \vdots             & \ddots & \vdots                                \\
-1       & x_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{\ell-1}x_i^{\binom{n}{2}}
+1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
 \end{bmatrix}
@@
      &      & \ddots & \ddots      &       \\
-     &      &        & -y_{\ell-1} & y_\ell
+     &      &        & -y_{m-1} & y_m
 \end{bmatrix}.
 \end{equation}
-Note that the $(n-1)$ by $\ell$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $x_l$ are nonlinear functions of $\boldsymbol y$.
+Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.
 
@@
 
-% DEPRECATED:
-%
-% \subsection*{Constant MLE}\label{sec:model:constant}
-%
-% We initialize by fitting a constant population size.
-% According to WSD's scribbles, the MLE assuming $\eta(t) = \eta_0$ (constant) is $\hat \eta_0 = \frac{S}{2 r H_{n-1}}$, where $S$ is the number of segregating sites (the sum of the observed SFS vector), $r$ is the mutation rate per genome per generation, and $H_{n-1}$ is the $n$th harmonic number.
-% This was derived by using the well-known result (cited in Rosen et al.) that the expected SFS for a constant population is given by $\xi_i = \frac{2\eta_0}{i}$.
-% Then the likelihood for $\eta_0$ is a Poisson random field parameterized by the $\xi_i$.
-%
-%
-% \subsection*{The coalescent horizon: tempora incognita (see what I did there? ;))}\label{sec:model:horizon}
-%
-% The coalescent tree height determines the time scale within which the SFS contains demographic information.
-% The expected height for n samples and a constant history $\eta(t)=\eta_0$ is the sum of the independent exponentially-distributed intercoalescent expectations:
-% \[
-% \mathbb{E}[T_n] = \sum_{k=2}^n \frac{\eta_0}{\binom{k}{2}} = 2\eta_0\left(1-\frac{1}{n}\right),
-% \]
-% where the last equality follows by recognizing the telescoping series.
-% Likewise for the variance:
-% \[
-% \text{Var}[T_n] = \sum_{k=2}^n \left(\frac{\eta_0}{\binom{k}{2}}\right)^2 = \frac{4 \eta _0^2 \left(-6 n^2 \psi ^{(1)}(n)+\left(\left(\pi ^2-9\right) n+6\right) n+3\right)}{3 n^2},
-% \]
-% where $\psi ^{(1)}(n)$ is the derivative of the digamma function \url{https://en.wikipedia.org/wiki/Trigamma_function}.
-%
-%
-% \subsection*{Asymptotically constant boundary condition}\label{sec:model:boundary}
-%
-% % Kelley's eqn
-% In real data (or realistic simulations) the time region where the SFS has lost information is determined by the expected number of coalescent trees $\rho\ell$, where $\rho$ is the recombination rate and $\ell$ is the genome size.
-% For $t$ such that the expected number of trees that have $T_{\text{MRCA}} > t$ becomes less that one, the SFS does not contain information beyond $t$.
-% To wit,
-% \[
-% \mathbb{P}\left(T_{\text{MRCA}}>t\right) < \frac{1}{\rho \ell}.
-% \]
-% The P\&K model doesn't model finite genome size, so it makes sense to use the CDF of $T_{\text{MRCA}}$ in that setting.
+\subsection{Tempora incognita: regularization and the coalescent horizon}\label{sec:model:loss}
+
+In linkage equilibrium the log likelihood function of the history $(\boldsymbol y, \boldsymbol z)$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ is the Poisson random field \citep{}
+\begin{align}
+\ell(\boldsymbol y, \boldsymbol z) &= \sum_{b=1}^{n-1}\left(x_b\log(\xi_b) - \xi_b - \log(x_b!)\right)\\
+&= \boldsymbol x^\intercal\log(\boldsymbol\xi) - \boldsymbol 1^\intercal\left(\boldsymbol\xi + \lngamma(\boldsymbol x + \boldsymbol 1)\right)\\
+&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal\left(C M(\boldsymbol y) \boldsymbol z + \lngamma(\boldsymbol x + \boldsymbol 1)\right),
+\end{align}
+where $\log(\cdot)$ and the log-gamma function $\lngamma(\cdot)$ operate on vectors elementwise.
+
+The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
+To penalize oscillatory solutions we introduce regularizers on the history derivatives.
+Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
+Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
+To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
+At times for which the CDF is near unity, the TMRCA has probably occurred, so we should penalize heavily because the data are not informative about the history at this time.
+At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
+From \eqref{eqn:r} the CDF $1 - r_2(t)$, so we define the regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ as
+\begin{align}
+\rho_\eta[\eta(t)] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)\left(1-r_2(t)\right)dt,\\
+\rho_\mu[\eta(t), \mu(t)] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)\left(1-r_2(t)\right)dt.
+\end{align}
+With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and it is straightforward to show (using notation as in section \ref{sec:model:pcsws}) that these become
+\begin{align}
+\rho_\eta(\boldsymbol y) &= \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2,\\
+\rho_\mu(\boldsymbol y, \boldsymbol z) &= \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2.
+\end{align}
+where
+\begin{equation}
+\label{eqn:Gamma}
+\Gamma(\boldsymbol y) \equiv
+\begin{bmatrix}
+\sqrt{1-u_0} &     &        &                       \\
+  & \sqrt{1-u_0 u_1} &        &                       \\
+  &     & \ddots &                       \\
+  &     &        & \sqrt{1-\prod_{i=1}^{m-1}u_i}
+\end{bmatrix}
+\begin{bmatrix}
+1  &      &        &             &       \\
+-1 & 1  &        &             &       \\
+     & -1 & 1    &             &       \\
+     &      & \ddots & \ddots      &       \\
+     &      &        & -1 & 1
+\end{bmatrix}.
+\end{equation}
+Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.
+
+Our regularized loss function is then
+\begin{align}
+L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z)\\
+&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2
+\end{align}
+where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.
 

### 2019-08-22 manuscript/main.tex
ms edits

@@
 \]
-Using the explicit bounds of the coalescent intervals gives
+Using the explicit bounds of the intercoalescent intervals gives
 \begin{align}
@@
 Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
-Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=2,\dots,t_m$, and $u_0\equiv 1$.
+Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$.%, and $u_0\equiv 1$.
 With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
@@
 \begin{bmatrix}
+1  &      &        &             &       \\
+-1 & 1  &        &             &       \\
+     & -1 & 1    &             &       \\
+     &      & \ddots & \ddots      &       \\
+     &      &        & -1 & 1
+\end{bmatrix}
+\begin{bmatrix}
 y_1 &     &      &             &       \\
--y_1 & y_2  &        &             &       \\
-     & -y_2 & y_3    &             &       \\
-     &      & \ddots & \ddots      &       \\
-     &      &        & -y_{m-1} & y_m
+    & y_2 &      &             &       \\
+     &     & y_3 &             &       \\
+     &      &    & \ddots      &       \\
+     &      &        &  & y_m
 \end{bmatrix}.
@@
 
-\subsection{Tempora incognita: regularization and the coalescent horizon}\label{sec:model:loss}
+\subsection{Tempora incognita: regularization toward the coalescent horizon}\label{sec:model:loss}
 
@@
 At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
-From \eqref{eqn:r} the CDF $1 - r_2(t)$, so we define the regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ as
+From \eqref{eqn:pi} and \eqref{eqn:r} the CDF is
 \begin{align}
-\rho_\eta[\eta(t)] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)\left(1-r_2(t)\right)dt,\\
-\rho_\mu[\eta(t), \mu(t)] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)\left(1-r_2(t)\right)dt.
+F_2(t) &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
+&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
+\end{align}
+We define $L_2$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
+\begin{align}
+\rho_\eta[\eta(t); h] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt,\\
+\rho_\mu[\eta(t), \mu(t); h] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt.
 \end{align}
@@
 \begin{align}
-\rho_\eta(\boldsymbol y) &= \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2,\\
-\rho_\mu(\boldsymbol y, \boldsymbol z) &= \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2.
+\rho_\eta(\boldsymbol y, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2,\\
+\rho_\mu(\boldsymbol y, \boldsymbol z, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2.
 \end{align}
 where
-\begin{equation}
+
+\emph{this is wrong, see notes!}
+
+\begin{align}
 \label{eqn:Gamma}
-\Gamma(\boldsymbol y) \equiv
+\Gamma(\boldsymbol y, h) \equiv&
 \begin{bmatrix}
-\sqrt{1-u_0} &     &        &                       \\
-  & \sqrt{1-u_0 u_1} &        &                       \\
+\sqrt{h+(1-h)\sum_2^n A_{2,j}u_0^{\binom{j}{2}}} &     &        &                       \\
+  & \sqrt{h+(1-h)\sum_2^n A_{2,j}(u_0u_1)^{\binom{j}{2}}} &        &                       \\
   &     & \ddots &                       \\
-  &     &        & \sqrt{1-\prod_{i=1}^{m-1}u_i}
-\end{bmatrix}
+  &     &        & \sqrt{h+(1-h)\sum_2^n A_{2,j}\left(\prod_{i=1}^{m-1}u_i\right)^{\binom{j}{2}}}
+\end{bmatrix}\\
+&\times
 \begin{bmatrix}
@@
 \end{bmatrix}.
-\end{equation}
+\end{align}
 Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.
@@
 L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z)\\
-&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y) \boldsymbol z\|_2^2
+&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2
 \end{align}

### 2019-08-22 dement.py
mushi code!

@@ -6,7 +6,6 @@ from scipy.special import binom
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
 from scipy.optimize import minimize
-from typing import Tuple
 
 
 class DemEnt():
@@ -129,7 +128,7 @@ class DemEnt():
         d = - sfs_hat.dot(np.log(xi_hat / sfs_hat))
         return d
 
-    def constant_MLE(self) -> float:
+    def constant_MLE(self) -> None:
         '''MLE for constant demography. Updates the instance's self.y_inferred
         based on the constant MLE
         '''

### 2019-08-22 mushi.py
mushi code!

@@ -1,0 +1,260 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import numpy as np
+from scipy.special import binom
+from functools import lru_cache
+from scipy.stats import poisson
+from scipy.special import gammaln
+from matplotlib import pyplot as plt
+from scipy.optimize import minimize
+
+
+class History():
+    '''piecewise constant history of population size η and mutation rate μ
+    '''
+
+    def __init__(self, t: np.array,
+                 y: np.ndarray = None, z: np.ndarray = None):
+        '''
+        t: The time axis, given as epoch change points. The first epoch starts
+           at zero, and the last epoch extends to infinity
+        y: vector of constant η(t) pieces
+        z: vector of constant μ(t) pieces
+        '''
+        if any(np.diff(t) <= 0) or np.isinf(t).sum() or any(t <= 0):
+            raise ValueError('t must be increasing, finite, and positive')
+        self.t = np.concatenate(([0], t, [np.inf]))
+        if y is not None:
+            if len(y) != len(t) + 1:
+                raise ValueError(f'len(t) = {len(t)} implies {len(t) + 1} '
+                                 f'epochs, but len(y) = {len(y)}')
+            if any(y <= 0) or np.isinf(y).sum():
+                raise ValueError('elements of y must be finite and positive')
+            self.y = y
+        else:
+            self.y = None
+        if z is not None:
+            if len(z) != len(t) + 1:
+                raise ValueError(f'len(t) = {len(t)} implies {len(t) + 1} '
+                                 f'epochs, but len(z) = {len(z)}')
+            if any(z <= 0) or np.isinf(z).sum():
+                raise ValueError('elements of z must be finite and positive')
+            self.z = z
+        else:
+            self.z = None
+
+    def η(self):
+        return History(self.t[1:-1], y=self.y)
+
+    def μ(self):
+        return History(self.t[1:-1], z=self.z)
+
+
+class SFS():
+    '''The SFS model described in the text
+    '''
+
+    def __init__(self, n: int = None, sfs: np.ndarray = None):
+        '''
+        pass one of these arguments
[…]

### 2019-08-23 manuscript/main.tex
correcting horizon-aware Tinkhonov matrix (tex and python)

@@
 Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
-Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$.%, and $u_0\equiv 1$.
+Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
 With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
@@
 
-\subsection{Tempora incognita: regularization toward the coalescent horizon}\label{sec:model:loss}
+\subsection{Poisson random field log likelihood}\label{sec:model:ell}
 
@@
 where $\log(\cdot)$ and the log-gamma function $\lngamma(\cdot)$ operate on vectors elementwise.
+
+\subsection{Tempora incognita: regularization toward the coalescent horizon}\label{sec:model:loss}
 
@@
 \begin{align}
-F_2(t) &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
+F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
 &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
 \end{align}
-We define $L_2$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
+
+We define $L_p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
 \begin{align}
-\rho_\eta[\eta(t); h] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt,\\
-\rho_\mu[\eta(t), \mu(t); h] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt.
+\rho_\eta[\eta(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\eta(t)}{dt}\right|^p dt,\\
+\rho_\mu[\eta(t), \mu(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\mu(t)}{dt}\right|^p dt.
 \end{align}
@@
 \begin{align}
-\rho_\eta(\boldsymbol y, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2,\\
-\rho_\mu(\boldsymbol y, \boldsymbol z, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2.
+\rho_\eta(\boldsymbol y, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p,\\
+\rho_\mu(\boldsymbol y, \boldsymbol z, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p.
 \end{align}
 where
-
-\emph{this is wrong, see notes!}
-
 \begin{align}
 \label{eqn:Gamma}
-\Gamma(\boldsymbol y, h) \equiv&
+\Gamma(\boldsymbol y, h) \equiv &\left(
 \begin{bmatrix}
-\sqrt{h+(1-h)\sum_2^n A_{2,j}u_0^{\binom{j}{2}}} &     &        &                       \\
-  & \sqrt{h+(1-h)\sum_2^n A_{2,j}(u_0u_1)^{\binom{j}{2}}} &        &                       \\
-  &     & \ddots &                       \\
-  &     &        & \sqrt{h+(1-h)\sum_2^n A_{2,j}\left(\prod_{i=1}^{m-1}u_i\right)^{\binom{j}{2}}}
-\end{bmatrix}\\
+h-1 &   &   &        & \\
+    & h &   &        & \\
+    &   & h &        & \\
+    &   &   & \ddots & \\
+    &   &   &        & h
+\end{bmatrix}
++ (1-h)\times
+\text{diag}\left(
+\begin{bmatrix}
+A_{2,2} & A_{2,3} & \hdots & A_{2,n}
+\end{bmatrix}
+\times
+\begin{bmatrix}
+1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
+1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
+\vdots  & \vdots             & \ddots & \vdots                                \\
+1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
+\end{bmatrix}\right)\right)\\
 &\times
@@
 \end{align}
-Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.
+% Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.
 
@@
 \begin{align}
-L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z)\\
-&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2
+L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y, h, p) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z, h, p)\\
+&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p
 \end{align}
 where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.
+
+\subsection{$p=1$}
+
+To optimize with $p=1$ we use the alternating direction method of multipliers (ADMM) of \cite{Boyd2011-dm}, which allows us to handle the non-differentiable penalty term.
 
@@
 
+\section{Acknowledgements}\label{sec:ack}
+Jean Feng suggested ADMM (if we end up using that).
+
 \bibliographystyle{plainnat}

### 2019-08-23 mushi.py
correcting horizon-aware Tinkhonov matrix (tex and python)

@@ -25,6 +25,7 @@ class History():
         if any(np.diff(t) <= 0) or np.isinf(t).sum() or any(t <= 0):
             raise ValueError('t must be increasing, finite, and positive')
         self.t = np.concatenate(([0], t, [np.inf]))
+        self.m = len(self.t) - 1
         if y is not None:
             if len(y) != len(t) + 1:
                 raise ValueError(f'len(t) = {len(t)} implies {len(t) + 1} '
@@ -132,12 +133,11 @@ class SFS():
 
         return self._binom_array_recip \
                @ np.cumprod(u[np.newaxis, :-1], axis=1) ** self._binom_vec[:, np.newaxis] \
-               @ (np.eye(len(history.y), k=0) - np.eye(len(history.y), k=-1)) \
+               @ (np.eye(history.m, k=0) - np.eye(history.m, k=-1)) \
                @ np.diag(history.y)
 
-    @staticmethod
     @lru_cache(maxsize=1)
-    def Γ(history: History, h: np.float = 1) -> np.ndarray:
+    def Γ(self, history: History, h: np.float = 1) -> np.ndarray:
         '''The Gamma matrix defined in the text
 
         h: relative increase in penalty as we approach coalescent horizon
@@ -146,9 +146,19 @@ class SFS():
         s = np.diff(history.t)
         u = np.exp(-s / history.y)
         u = np.insert(u, 0, 1)
-        return np.diag(np.sqrt(h - (h - 1) * np.cumprod(u[:-1]))
-                       ) @ (np.diag(np.ones_like(history.y), k=0)
-                             - np.diag(np.ones_like(history.y[:-1]), k=-1))
+
+        # the A_2j product of these terms
+        l = np.array(range(2, self.n + 1))[:, np.newaxis]
+        with np.errstate(divide='ignore'):
+            A2_terms = l * (l-1) / (l * (l-1) - l.T * (l.T-1))
+        np.fill_diagonal(A2_terms, 1)
+        A2 = np.prod(A2_terms, axis=0)
+
+        return (np.diag(np.insert(h * np.ones(history.m - 1), 0, h - 1)) + (1 - h)
+                * np.diagflat(A2[np.newaxis, :]
+                          @ np.cumprod(u[np.newaxis, :-1],
+                                       axis=1) ** self._binom_vec[:, np.newaxis])) \
+               @ (np.eye(history.m, k=0) - np.eye(history.m, k=-1))
 
     def ξ(self, history: History) -> np.ndarray:
[…]

### 2019-08-23 mushi.py
class methods needed for `mushi.History` hashability

@@ -51,6 +51,27 @@ class History():
     def μ(self):
         return History(self.t[1:-1], z=self.z)
 
+    def __hash__(self):
+        '''needed for hashability
+        '''
+        return hash((tuple(self.t) if self.t is not None else self.t,
+                     tuple(self.y) if self.y is not None else self.y,
+                     tuple(self.z) if self.z is not None else self.z))
+
+    def __eq__(self, other) -> bool:
+        '''needed for hashability
+        '''
+        if self.t is not None:
+            if any(self.t != other.t):
+                return False
+        if self.y is not None:
+            if any(self.y != other.y):
+                return False
+        if self.z is not None:
+            if any(self.z != other.z):
+                return False
+        return True
+
 
 class SFS():
     '''The SFS model described in the text
@@ -132,7 +153,8 @@ class SFS():
         u = np.insert(u, 0, 1)
 
         return self._binom_array_recip \
-               @ np.cumprod(u[np.newaxis, :-1], axis=1) ** self._binom_vec[:, np.newaxis] \
+               @ np.cumprod(u[np.newaxis, :-1],
+                            axis=1) ** self._binom_vec[:, np.newaxis] \
                @ (np.eye(history.m, k=0) - np.eye(history.m, k=-1)) \
                @ np.diag(history.y)
 
@@ -196,7 +218,8 @@ class SFS():
         ρ_μ = ((Γ @ history.z)**2).sum()
         return -self.ℓ(history) + λ_η * ρ_η + λ_μ * ρ_μ
 
-    def infer_μ(self, history: History, λ_μ: np.float = 0, h: np.float = 1):
+    def infer_μ(self, history: History, λ_μ: np.float = 0,
+                h: np.float = 1) -> History:
         '''infer the μ history given the simulated sfs and η history
 
         history: initial history
@@ -204,9 +227,8 @@ class SFS():
         h: relative increase in penalty as we approach coalescent horizon
         '''
         def f(z) -> np.float:
-            _history = history
-            _history.z = z
-            return self.L(_history, λ_η=0, λ_μ=λ_μ, h=h)
+            return self.L(History(history.t[1:-1], history.y, z),
+                          λ_η=0, λ_μ=λ_μ, h=h)
 
         result = minimize(f,
                           history.z,
@@ -219,10 +241,8 @@ class SFS():
                           bounds=[(1e-10, None)] * history.m
                           )
         if not result.success:
-            raise ValueError(result.message)
-        else:
[…]

### 2019-08-23 mushi.py
initialize \mu fitting with MLE constant history

@@ -77,22 +77,22 @@ class SFS():
     '''The SFS model described in the text
     '''
 
-    def __init__(self, n: int = None, sfs: np.ndarray = None):
+    def __init__(self, n: int = None, x: np.ndarray = None):
         '''
         pass one of these arguments
         n: number of sampled haplotypes
         sfs: observed sfs vector
         '''
-        if sfs is not None:
-            self.sfs = sfs
-            self.n = len(sfs) + 1
-            assert n is None, 'pass only one of n or sfs'
+        if x is not None:
+            self.x = x
+            self.n = len(x) + 1
+            assert n is None, 'pass only one of n or x'
         elif n is not None:
-            self.sfs = None
+            self.x = None
             self.n = n
-            assert sfs is None, 'pass only one of n or sfs'
+            assert x is None, 'pass only one of n or x'
         else:
-            raise ValueError('must pass either n or sfs')
+            raise ValueError('must pass either n or x')
         if self.n < 2:
             raise ValueError('n must be larger than 1')
         self._binom_vec = binom(np.arange(2, self.n + 1), 2)
@@ -146,7 +146,12 @@ class SFS():
     @lru_cache(maxsize=1)
     def M(self, history: History) -> np.ndarray:
         '''The M matrix defined in the text
+
+        history: η history
         '''
+        if history.z is not None:
+            raise ValueError('history must be only η, not μ. Consider using '
+                             'the History.η() method on a joint history')
         # epoch durations
         s = np.diff(history.t)
         u = np.exp(-s / history.y)
@@ -162,8 +167,12 @@ class SFS():
     def Γ(self, history: History, h: np.float = 1) -> np.ndarray:
         '''The Gamma matrix defined in the text
 
+        history: η history
         h: relative increase in penalty as we approach coalescent horizon
         '''
+        if history.z is not None:
+            raise ValueError('history must be only η, not μ. Consider using '
+                             'the History.η() method on a joint history')
         # epoch durations
         s = np.diff(history.t)
         u = np.exp(-s / history.y)
@@ -190,22 +199,36 @@ class SFS():
     def simulate(self, history: History) -> None:
         '''simulate a SFS under the Poisson random field model (no linkage)
         '''
-        self.sfs = poisson.rvs(self.ξ(history))
+        self.x = poisson.rvs(self.ξ(history))
 
     def ℓ(self, history: History) -> np.float:
         '''Poisson random field log-likelihood of history
         '''
[…]

### 2019-08-23 mushi.py
refactoring `mushi.py`, including a slick plotting interface

@@ -1,6 +1,7 @@
 #! /usr/bin/env python
 # -*- coding: utf-8 -*-
 
+from dataclasses import dataclass
 import numpy as np
 from scipy.special import binom
 from functools import lru_cache
@@ -10,67 +11,99 @@ from matplotlib import pyplot as plt
 from scipy.optimize import minimize
 
 
-class History():
-    '''piecewise constant history of population size η and mutation rate μ
+@dataclass(frozen=True)
+class PiecewiseConstantHistory():
+    '''The first epoch starts at zero, and the last epoch extends to infinity.
+    Can be used for η or μ
+
+    change_points: epoch change points (times)
+    vals: vector of constant values for each epoch
     '''
+    change_points: np.array
+    vals: np.ndarray
 
-    def __init__(self, t: np.array,
-                 y: np.ndarray = None, z: np.ndarray = None):
+    def __post_init__(self):
+        if any(np.diff(self.change_points) <= 0) or any(
+           np.isinf(self.change_points)) or any(self.change_points <= 0):
+            raise ValueError('change_points must be increasing, finite, and '
+                             'positive')
+        if len(self.vals) != len(self.change_points) + 1:
+            raise ValueError(f'len(change_points) = {len(self.change_points)} implies '
+                             f'{len(self.change_points) + 1} epochs, but len(vals) = '
+                             f'{len(self.vals)}')
+        if any(self.vals <= 0) or np.isinf(self.vals).sum():
+            raise ValueError('elements of vals must be finite and positive')
+
+    def m(self):
+        '''number of epochs
         '''
-        t: The time axis, given as epoch change points. The first epoch starts
-           at zero, and the last epoch extends to infinity
-        y: vector of constant η(t) pieces
-        z: vector of constant μ(t) pieces
-        '''
-        if any(np.diff(t) <= 0) or np.isinf(t).sum() or any(t <= 0):
-            raise ValueError('t must be increasing, finite, and positive')
-        self.t = np.concatenate(([0], t, [np.inf]))
-        self.m = len(self.t) - 1
-        if y is not None:
-            if len(y) != len(t) + 1:
-                raise ValueError(f'len(t) = {len(t)} implies {len(t) + 1} '
-                                 f'epochs, but len(y) = {len(y)}')
-            if any(y <= 0) or np.isinf(y).sum():
[…]

### 2019-08-26 manuscript/main.tex
he's into it
a schematic figure of our model

@@
 \usepackage{bbm}
-
+\usepackage{graphicx}
 \usepackage{hyperref}
@@
 \author{
-W. DeWitt$^{1}$, K. Harris (if he's into it?)$^{2}$, and K. Harris$^{1}$\\
+W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
 \small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}
@@
 We will retrace the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
-We will make use of important results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} which facilitate numerical computation under this model, and formulate the expected 3-SFS under given demographic and mutation spectrum histories following the notation of \cite{Rosen2018-bb} for the expected SFS with a constant mutation process.
+We will make use of important results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} which facilitate numerical computation under this model, and formulate the expected 3-SFS under piecewise constant demographic and mutation spectrum histories building on the notation of \cite{Rosen2018-bb} for the expected SFS with a piecewise constant demographic history and constant mutation process.
 
@@
 The following argument can be generalized to a nonuniform distribution of mutation intensity over the genome without changing the results.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/model}
+\caption{Schematic of the marked Poisson process model with $n=4$.
+We condition on coalescent times $T_4=t_4,T_3=t_3,T_2=t_2$ and consider mutation intensity function $\mu(t)$.
+Red dots indicate mutation events placed by time $t$, genomic position $x$, and coalescent line (which are depicted as extruded in the genomic coordinate axis, grey rectangles).
+The probability that a differential element $dxdt$ on a given sheet contains a mutation is proportional to the instantaneous mutation intensity $\mu(t)$.
+}
+\label{fig:model}
+\end{figure}
 
@@
 \item repeat 1KG analysis like in \cite{Harris2017-fw}, see if we recapitulate Kelley's simulation-based pulse results
-\item cluster triplet time series to see if we pull in minor components. Dynamic time warping?
+\item cluster triplet time series to see if we pull in minor components.
 \end{itemize}

### 2019-08-26 mushi.py
analytical grad for \mu, closes #3 (can reopen if we want \eta gradient)

@@ -6,7 +6,6 @@ import numpy as np
 from scipy.special import binom
 from functools import lru_cache
 from scipy.stats import poisson
-from scipy.special import gammaln
 from matplotlib import pyplot as plt
 from scipy.optimize import minimize
 
@@ -28,9 +27,9 @@ class PiecewiseConstantHistory():
             raise ValueError('change_points must be increasing, finite, and '
                              'positive')
         if len(self.vals) != len(self.change_points) + 1:
-            raise ValueError(f'len(change_points) = {len(self.change_points)} implies '
-                             f'{len(self.change_points) + 1} epochs, but len(vals) = '
-                             f'{len(self.vals)}')
+            raise ValueError(f'len(change_points) = {len(self.change_points)}'
+                             f' implies {len(self.change_points) + 1} epochs,'
+                             f' but len(vals) = {len(self.vals)}')
         if any(self.vals <= 0) or np.isinf(self.vals).sum():
             raise ValueError('elements of vals must be finite and positive')
 
@@ -89,19 +88,17 @@ class JointHistory():
         if fig:
             axes = fig.axes
         else:
-            fig, axes = plt.subplots(2, 1, sharex=True)
+            fig, axes = plt.subplots(2, 1, sharex=True, figsize=(6, 6))
         plt.sca(axes[0])
         self.η.plot(**kwargs)
         plt.ylabel('$η(t)$')
-        plt.ylim([0, None])
-        plt.xscale('symlog')
 
         plt.sca(axes[1])
         self.μ.plot(**kwargs)
         plt.xlabel('$t$')
+        plt.xscale('symlog')
         plt.ylabel('$μ(t)$')
         plt.ylim([0, None])
-        plt.xscale('symlog')
 
         return fig
 
@@ -209,6 +206,7 @@ class SFS():
         u = np.insert(u, 0, 1)
 
         # the A_2j product of these terms
+        # NOTE: using letter  "l" as a variable name to match text
         l = np.array(range(2, self.n + 1))[:, np.newaxis]
         with np.errstate(divide='ignore'):
             A2_terms = l * (l-1) / (l * (l-1) - l.T * (l.T-1))
@@ -217,28 +215,45 @@ class SFS():
 
         return (np.diag(np.insert(h * np.ones(η.m() - 1), 0, h - 1)) + (1 - h)
                 * np.diagflat(A2[np.newaxis, :]
[…]

### 2019-08-26 mushi.py
prox implemented, and it's friggin fast! (h/t @kharris), closes #13

@@ -7,7 +7,7 @@ from scipy.special import binom
 from functools import lru_cache
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
-from scipy.optimize import minimize
+import prox_tv as ptv
 
 
 @dataclass(frozen=True)
@@ -266,91 +266,39 @@ class SFS():
         z0 = (self.x.sum() / (SFS.C(self.n) @ self.M(η)).sum())
         return PiecewiseConstantHistory(η.change_points, z0 * np.ones(η.m()))
 
-    def ρ_η(self, history: JointHistory, h: np.float = 1,
-            grad_η: bool = False):
-        '''η regulizer described in the text
-
-        history: η and μ joint history
-        h: relative increase in penalty as we approach coalescent horizon
-        grad_η: flag to return gradient wrt history.η
-        '''
-        Γ = self.Γ(history.η, h)
-        y = history.η.vals
-        ρ_η = ((Γ @ y)**2).sum()
-        if grad_η:
-            raise NotImplementedError('gradient of ρ_η not implemented')
-        return ρ_η
-
-    def ρ_μ(self, history: JointHistory, h: np.float = 1,
-            grad_μ: bool = False):
-        '''μ regulizer described in the text
-
-        history: η and μ joint history
-        h: relative increase in penalty as we approach coalescent horizon
-        grad_μ: flag to return gradient wrt history.μ
-        '''
-        Γ = self.Γ(history.η, h)
-        z = history.μ.vals
-        ρ_μ = ((Γ @ z)**2).sum()
-        if grad_μ:
-            dρ_μdμ = 2 * Γ.T @ Γ @ z
-            return ρ_μ, dρ_μdμ
-        return ρ_μ
-
-    def L(self, history: JointHistory,
-          λ_η: np.float = 0, λ_μ: np.float = 0,
-          h: np.float = 1, grad_μ: bool = False) -> np.float:
-        '''loss: negative log likelihood (Poisson random field) and
-        regularization as described in the text
-
-        history: η and μ joint history
-        λ_η: regularization strength on η
-        λ_μ: regularization strength on μ
-        h: relative increase in penalty as we approach coalescent horizon
-        grad_μ: flag to return gradient wrt history.μ
-        '''
-        ρ_η = self.ρ_η(history, h)
-        if grad_μ:
[…]

### 2019-08-27 manuscript/main.tex
fig tweak

@@
 We condition on coalescent times $T_4=t_4,T_3=t_3,T_2=t_2$ and consider mutation intensity function $\mu(t)$.
-Red dots indicate mutation events placed by time $t$, genomic position $x$, and coalescent line (which are depicted as extruded in the genomic coordinate axis, grey rectangles).
+Red dots indicate mutation events placed by time $t$, genomic position $x$, and coalescent line (which are depicted as extruded in the genomic coordinate axis, grey sheets).
 The probability that a differential element $dxdt$ on a given sheet contains a mutation is proportional to the instantaneous mutation intensity $\mu(t)$.

### 2019-08-27 manuscript/main.tex
prototype pulse fitting figure
integrating KDH comments

@@
 This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
+%WD do we actually do joint inference? If not, maybe save for discussion
+
+\section{Model}\label{sec:model}
 
@@
 
-\section{Model}\label{sec:model}
-
 \subsection{The expected SFS given demographic and mutation rate histories}\label{sec:model:xi}
 
-Suppose $n$ haplotypes are sampled in the present, and let rvs $T = T_2,\dots,T_n$ denote the coalescent times measured retrospectively from the present (i.e.
+Suppose $n$ haplotypes are sampled in the present, and let rvs $(T = T_2,\dots,T_n)$ denote the coalescent times measured retrospectively from the present (i.e.
 $T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
 As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
-This is tantamount to an (uncountably) infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
-Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) \ge 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
+This is tantamount to the infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
+%WD Kam suggests replacing below with $\mu^* \ge \mu(t) \ge 0$, but I don't get why
+Further suppose that mutation rate is not constant, but a specified function of time $0\le \mu(t)<\infty$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
 A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
-The following argument can be generalized to a nonuniform distribution of mutation intensity over the genome without changing the results.
+The following argument can be generalized to allow the labelling distribution to be nonuniform over the unit interval, but the same results follow.
 
@@
 
-Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
+Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation present in $b\in\{1, 2, \dots, n-1\}$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (the probability of this event is independent of $x$, given the uniform labeling distribution).
 Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
@@
 \mathbbm{P}(\mathcal{E}_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(\mathcal{E}_{\delta x, b, k}\mid T)\\
-&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O(\delta x)\right),
+&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O((\delta x)^2)\right),
 \end{align*}
@@
 \end{equation}
-is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parenthesis is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
+is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parentheses is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
 Marginalizing $T$ gives
 \begin{align*}
-\mathbbm{P}(\mathcal{E}_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right].
+\mathbbm{P}(\mathcal{E}_{\delta x, b}) &= \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right] + O((\delta x)^2).
 \end{align*}
@@
 \[
-\xi_b = \int d\mathbbm{P}(\mathcal{E}_{dx, b}) = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
+\xi_b = \int_0^1 \mathbbm{P}(\mathcal{E}_{dx, b})dx = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
 \]
-Using the explicit bounds of the intercoalescent intervals gives
+Using the explicit bounds of the intercoalescent intervals, $I_{k} = (T_{k+1}, T_k)$, gives
 \begin{align}
 \label{eqn:xi}
+%WD Kam noted an "off by -1 error" below, but I think it's correct as is
 \xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
@@
 \end{equation}
-corresponding to density $q_j(t)$.
+corresponding to density $q_j(t)$ (that this is the appropriate survival function can be seen by considering the limit of a sequence of Bernoulli failures).
 
@@
 \xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt\nonumber\\
-&= \sum_{j=2}^n \left(\sum_{k=2}^j B_{b,k} A_{k,j}\right) \int_0^\infty r_j(t)\mu(t)dt.
+&= \sum_{j=2}^n \left(\sum_{k=2}^j B_{b,k} A_{k,j}\right) \int_0^\infty r_j(t)\mu(t)dt,
 \end{align}
+where we've exchanged summation order in the last line.
 
-We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
+We then have a linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
+%WD Kam suggests a linear operator notation that combines the integral on t with the summation on j. Note this operator would be (nonlinearly) dependent on \eta
 \begin{equation}
@@
 \label{eqn:d}
-d_j \equiv \int_0^\infty r_j(t)\mu(t)dt
+d_j \equiv \int_0^\infty r_j(t)\mu(t)dt = \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt
 \end{equation}
-has linear (nonlinear) dependence on $\mu$ ($\eta$).
+is a linear functional of $\mu$ and a nonlinear functional of $\eta$.
 
@@
 
-We next develop an efficient recursive procedure for computing $C$.
-Using \ref{eqn:B}
+%WD Kam suggests just doing C = B A, but computing the elements of B and A would seem computationally infeasible
+We next develop an efficient recursive procedure for computing the $(n-1)\times(n-1)$ matrix $C$.
+Using \eqref{eqn:B}
 \begin{align*}
@@
 \]
-so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \ref{eqn:W1} and \ref{eqn:W2} are definite sums over geometric terms.
-Zeilberger's algorithm \citep{petkovvsek1996b, paule1995mathematica} can thus be used to yield the following procedurally generated second-order recursions in $j$:
+so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \eqref{eqn:W1} and \eqref{eqn:W2} are definite sums over geometric terms.
+Zeilberger's algorithm, which finds polynomial recurrences for definite sums of hypergeometric terms \citep{petkovvsek1996b, paule1995mathematica}, can thus be used to yield the following procedurally generated second-order recursions in $j$:
+%WD need to wrap this eqn, as it runs will beyond the margin. Also make an appendix at some point for these
 \begin{align*}
@@
 
-Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
+%WD Kam describes this discretization as ad hoc, but I'm pretty sure this is your textbook Riemann sum with rectangular partitions.
+We could do fancier quadrature, but then we'll be less able to lean on previous work (i.e.
+Rosen et al.)
+
+Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
 \begin{equation}
 \label{eqn:d2}
-d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
+d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)d\tau,
 \end{equation}
@@
 Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
+%WD Kam hates the latin loooooooool. It really does follow by dropping in the y vector and proceeding as in their proof
 With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
@@
 
+%WD Kam suggests calling the data vector \xi instead of x, but I disagree because we're already using \xi for the expected SFS.
+one could quibble that we already used x in the marked Poisson process at the beginning, but I think our readers have forgotten that by now and this won't confuse
+
 In linkage equilibrium the log likelihood function of the history $(\boldsymbol y, \boldsymbol z)$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ is the Poisson random field \citep{}
@@
 
+%WD does Kam not hate this latin? lulz
 \subsection{Tempora incognita: regularization toward the coalescent horizon}\label{sec:model:loss}
 
+%WD more inverse problem intro
 The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
-To penalize oscillatory solutions we introduce regularizers on the history derivatives.
+To penalize oscillatory histories we regularize on their derivatives.
 Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
 Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
+%WD This is ad hoc, can we motivate it with a Bayesian argument?
 To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
@@
 At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
-From \eqref{eqn:pi} and \eqref{eqn:r} the CDF is
+From \eqref{eqn:pi} and \eqref{eqn:r} the CDF of $T_2$ is
 \begin{align}
@@
 
-We define $L_p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
+We define weighted $L_p^p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
 \begin{align}
-\rho_\eta[\eta(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\eta(t)}{dt}\right|^p dt,\\
-\rho_\mu[\eta(t), \mu(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\mu(t)}{dt}\right|^p dt.
+\rho_\eta[\eta(t); h, p] &\equiv \int_0^\infty\left(1+(h-1)F_2(t)\right)\left|\frac{d\eta(t)}{dt}\right|^p dt,\\
+\rho_\mu[\eta(t), \mu(t); h, p] &\equiv \int_0^\infty\left(1+(h-1)F_2(t)\right)\left|\frac{d\mu(t)}{dt}\right|^p dt.
 \end{align}
@@
 \begin{align}
-\rho_\eta(\boldsymbol y, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p,\\
-\rho_\mu(\boldsymbol y, \boldsymbol z, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p.
+\rho_\eta(\boldsymbol y, h, p) &= \sum_2^m w_i|y_i-y_{i-1}|^p,\\
+\rho_\mu(\boldsymbol y, \boldsymbol z, h, p) &= \sum_2^m w_i|z_i-z_{i-1}|^p.
 \end{align}
-where
+with weights
 \begin{align}
-\label{eqn:Gamma}
-\Gamma(\boldsymbol y, h) \equiv &\left(
-\begin{bmatrix}
-h-1 &   &   &        & \\
-    & h &   &        & \\
-    &   & h &        & \\
-    &   &   & \ddots & \\
-    &   &   &        & h
-\end{bmatrix}
-+ (1-h)\times
-\text{diag}\left(
+\label{eqn:w}
+\boldsymbol w(\boldsymbol y, h) \equiv
+h \boldsymbol 1 + (1-h)\times
+\left(
 \begin{bmatrix}
@@
 \begin{bmatrix}
-1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
-1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
-\vdots  & \vdots             & \ddots & \vdots                                \\
-1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
-\end{bmatrix}\right)\right)\\
-&\times
-\begin{bmatrix}
-1  &      &        &             &       \\
--1 & 1  &        &             &       \\
-     & -1 & 1    &             &       \\
-     &      & \ddots & \ddots      &       \\
-     &      &        & -1 & 1
-\end{bmatrix}.
+u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
+u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
+\vdots             & \ddots & \vdots                                \\
+u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
+\end{bmatrix}\right)^\intercal.
 \end{align}
@@
 
-Our regularized loss function is then
+Our fused LASSO \citep{tibs} loss function is then
 \begin{align}
-L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y, h, p) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z, h, p)\\
-&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p
+L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\rho_\eta(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\eta(\boldsymbol y, h, 2)\right) + \lambda_\mu \left(\alpha\rho_\mu(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\mu(\boldsymbol y, h, 2)\right),
 \end{align}
-where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.
-
-\subsection{$p=1$}
-
-To optimize with $p=1$ we use the alternating direction method of multipliers (ADMM) of \cite{Boyd2011-dm}, which allows us to handle the non-differentiable penalty term.
+where $\lambda_\eta$ and $\lambda_\mu$ tune the regularization strength on $\eta$ and $\mu$, respectively, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
 
 \section{Results}\label{sec:results}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{figures/fit_teaser}
+  \caption{Simulating and inverting a pulse.}
+  \label{}
+\end{figure}
 

### 2019-08-28 manuscript/main.tex
forgot y, and no more admm I guess
adding identifiability refs

@@
 %WD do we actually do joint inference? If not, maybe save for discussion
+
+Some references for nonidentifiability of demographic history from the SFS: \cite{Baharian2018-np, Bhaskar2014-fu, Terhorst2015-xt, Myers2008-jp}.
+It might be good to consider identifiability of mutation rate history from the SFS conditioned on demographic history.
 
@@
 \begin{align}
-L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\rho_\eta(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\eta(\boldsymbol y, h, 2)\right) + \lambda_\mu \left(\alpha\rho_\mu(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\mu(\boldsymbol y, h, 2)\right),
+L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\rho_\eta(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\eta(\boldsymbol y, h, 2)\right) + \lambda_\mu \left(\alpha\rho_\mu(\boldsymbol y, \boldsymbol z, h, 1) + \frac{(1-\alpha)}{2}\rho_\mu(\boldsymbol y, \boldsymbol z, h, 2)\right),
 \end{align}
@@
 \section{Acknowledgements}\label{sec:ack}
-Jean Feng suggested ADMM (if we end up using that).
 

### 2019-08-29 mushi.py
automatic differentiation with jax, and proper weighted norms

@@ -1,12 +1,20 @@
 #! /usr/bin/env python
 # -*- coding: utf-8 -*-
 
+from jax.config import config
+config.update("jax_enable_x64", True)
+
 from dataclasses import dataclass
-import numpy as np
+import jax.numpy as np
+import numpy as onp # original numpy
+from jax import jit, grad, value_and_grad, jacfwd
+from jax.lax import clamp
 from scipy.special import binom
 from functools import lru_cache
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
+from matplotlib import figure
+from scipy.optimize import minimize
 import prox_tv as ptv
 
 
@@ -30,8 +38,8 @@ class PiecewiseConstantHistory():
             raise ValueError(f'len(change_points) = {len(self.change_points)}'
                              f' implies {len(self.change_points) + 1} epochs,'
                              f' but len(vals) = {len(self.vals)}')
-        if any(self.vals <= 0) or np.isinf(self.vals).sum():
-            raise ValueError('elements of vals must be finite and positive')
+        # if any(self.vals <= 0) or np.sum(np.isinf(self.vals)):
+        #     raise ValueError('elements of vals must be finite and positive')
 
     def m(self):
         '''number of epochs
@@ -57,11 +65,16 @@ class PiecewiseConstantHistory():
 
         kwargs: key word arguments passed to plt.step
         '''
-        t = np.insert(self.change_points, 0, 0)
+        t = np.concatenate((np.array([0]), self.change_points))
         plt.step(t, self.vals, where='post', **kwargs)
         if 'label' in kwargs:
             plt.legend()
 
+    def arrays(self):
+        t = np.concatenate((np.array([0]),
+                            self.change_points,
+                            np.array([np.inf])))
+        return t, self.vals
 
 @dataclass()
 class JointHistory():
@@ -78,13 +91,23 @@ class JointHistory():
         if any(self.η.change_points != self.μ.change_points):
             raise ValueError('η and μ histories must use the same time grid')
 
-    def plot(self, **kwargs) -> np.ndarray:
+    def __hash__(self):
+        '''needed for hashability
+        '''
[…]

### 2019-08-29 dement.py
dement :gun:

@@ -1,293 +1,0 @@
-#! /usr/bin/env python
-# -*- coding: utf-8 -*-
-
-import numpy as np
-from scipy.special import binom
-from scipy.stats import poisson
-from matplotlib import pyplot as plt
-from scipy.optimize import minimize
-
-
-class DemEnt():
-    '''A class that implements the model of Rosen et al., but adds a Poisson
-    random field for generating the SFS from ξ
-    '''
-
-    def __init__(self, n: int, t: np.ndarray, y: np.ndarray, r: float = 1,
-                 infinite: bool = True):
-        '''
-        n: number of sampled haplotypes
-        t: The time axis. The last epoch in t extends to infinity in Rosen, but
-           we truncate if infinite=False
-        y: vector of eta pieces
-        r: mutation rate is per genome per generation (controls SFS noise)
-        infinite: extend to infinity with constant size (otherwise truncate)
-        '''
-        if t[0] != 0:
-            raise ValueError('the first element of t must be 0')
-        if any(np.diff(t) <= 0) or np.isinf(t).sum():
-            raise ValueError('t must be monotonically increasing and finite')
-        if any(y <= 0) or np.isinf(t).sum():
-            raise ValueError('elements of y must be finite and positive')
-        if n < 2:
-            raise ValueError('n must be larger than 1')
-        self.infinite = infinite
-        self.n = n
-        self.t = t
-        self.y_true = y
-        self.r = r
-        self._binom_array = binom(np.arange(2, n + 1), 2)[:, np.newaxis]
-        self.A = DemEnt.A(n)
-        self._s = np.diff(self.t)
-        self.simulate_sfs()  # this sets self.sfs
-
-    @staticmethod
-    def A(n: int) -> np.ndarray:
-        '''The A_n matrix of Polanski and Kimmel (2003) (equations 13–15)
-        Using notation of Rosen et al. (2018)
-
-        n: number of sampled haplotypes
-        '''
-        A = np.zeros((n - 1, n - 1))
-        b = np.arange(1, n - 1 + 1)
-        A[:, 0] = 6 / (n + 1)
-        A[:, 1] = 30 * (n - 2 * b) / (n + 1) / (n + 2)
-        for col in range(n - 3):
-            j = col + 2
-            c1 = - (1 + j) * (3 + 2 * j) * (n - j) / j / (2 * j - 1) / (n+j+1)
[…]

### 2019-08-31 mushi.py
no more weight ramp

@@ -1,22 +1,21 @@
 #! /usr/bin/env python
 # -*- coding: utf-8 -*-
 
-from jax.config import config
-config.update("jax_enable_x64", True)
-
 from dataclasses import dataclass
 import jax.numpy as np
-import numpy as onp # original numpy
-from jax import jit, grad, value_and_grad, jacfwd
-from jax.lax import clamp
+import numpy as onp  # original numpy
+from jax import jit, grad
+from jax.experimental.optimizers import adagrad
 from scipy.special import binom
 from functools import lru_cache
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
 from matplotlib import figure
-from scipy.optimize import minimize
+from scipy.optimize import minimize_scalar
 import prox_tv as ptv
 
+from jax.config import config
+config.update("jax_enable_x64", True)
 
 @dataclass(frozen=True)
 class PiecewiseConstantHistory():
@@ -119,7 +118,6 @@ class JointHistory():
         plt.sca(axes[1])
         self.μ.plot(**kwargs)
         plt.xlabel('$t$')
-        plt.xscale('symlog')
         plt.ylabel('$μ(t)$')
 
         return fig
@@ -229,12 +227,10 @@ class SFS():
                @ (np.eye(len(y), k=0) - np.eye(len(y), k=-1)) \
                @ np.diag(y)
 
-    @lru_cache(maxsize=1)
-    def w(self, η, h: np.float64 = 1) -> np.ndarray:
-        '''The weight vector defined in the text
+    def tmrca_cdf(self, η: PiecewiseConstantHistory) -> np.ndarray:
+        '''The cdf of the TMRCA of the sample at each change point
 
         η: η history
-        h: relative increase in penalty as we approach coalescent horizon
         '''
         t, y = η.arrays()
         # epoch durations
@@ -250,7 +246,7 @@ class SFS():
         onp.fill_diagonal(A2_terms, 1)
         A2 = np.prod(A2_terms, axis=0)
 
-        return h + (1 - h) * (A2[np.newaxis, :] @ SFS.cumprod(u[np.newaxis, 1:-1]) ** self.binom_vec[:, np.newaxis]).T
+        return 1 - (A2[np.newaxis, :] @ SFS.cumprod(u[np.newaxis, 1:-1]) ** self.binom_vec[:, np.newaxis]).T
 
     def ξ(self, history, use_cache=True) -> np.ndarray:
[…]

### 2019-08-31 mushi.py
removing jax and eta fitting

@@ -2,10 +2,7 @@
 # -*- coding: utf-8 -*-
 
 from dataclasses import dataclass
-import jax.numpy as np
-import numpy as onp  # original numpy
-from jax import jit, grad
-from jax.experimental.optimizers import adagrad
+import numpy as np
 from scipy.special import binom
 from functools import lru_cache
 from scipy.stats import poisson
@@ -14,8 +11,6 @@ from matplotlib import figure
 from scipy.optimize import minimize_scalar
 import prox_tv as ptv
 
-from jax.config import config
-config.update("jax_enable_x64", True)
 
 @dataclass(frozen=True)
 class PiecewiseConstantHistory():
@@ -163,8 +158,8 @@ class SFS():
 
         n: number of sampled haplotypes
         '''
-        W1 = onp.zeros((n - 1, n - 1))
-        b = onp.arange(1, n - 1 + 1)
+        W1 = np.zeros((n - 1, n - 1))
+        b = np.arange(1, n - 1 + 1)
         # j = 2
         W1[:, 0] = 6 / (n + 1)
         # j = 3
@@ -181,8 +176,8 @@ class SFS():
 
         n: number of sampled haplotypes
         '''
-        W2 = onp.zeros((n - 1, n - 1))
-        b = onp.arange(1, n - 1 + 1)
+        W2 = np.zeros((n - 1, n - 1))
+        b = np.arange(1, n - 1 + 1)
         # j = 2
         W2[:, 0] = 0
         # j = 3
@@ -201,14 +196,6 @@ class SFS():
         '''
         return SFS.W1(n) - SFS.W2(n)
 
-    @staticmethod
-    def cumprod(X: np.ndarray) -> np.ndarray:
-        '''Alas, jax doesn't differentiate np.cumprod
-
-        X: 2D array
-        '''
-        return np.exp(np.cumsum(np.log(X)))
-
     @lru_cache(maxsize=1)
     def M(self, η) -> np.ndarray:
         '''The M matrix defined in the text
@@ -222,8 +209,8 @@ class SFS():
         u = np.concatenate((np.array([1]), u))
 
         return self.binom_array_recip \
-               @ SFS.cumprod(u[np.newaxis, :-1],
-                         ) ** self.binom_vec[:, np.newaxis] \
+               @ np.cumprod(u[np.newaxis, :-1],
+                            axis=1) ** self.binom_vec[:, np.newaxis] \
[…]

### 2019-08-31 manuscript/main.tex

@@
 \begin{align}
-\ell(\boldsymbol y, \boldsymbol z) &= \sum_{b=1}^{n-1}\left(x_b\log(\xi_b) - \xi_b - \log(x_b!)\right)\\
+\ell(\boldsymbol y, \boldsymbol z; \boldsymbol x) &= \sum_{b=1}^{n-1}\left(x_b\log(\xi_b) - \xi_b - \log(x_b!)\right)\\
 &= \boldsymbol x^\intercal\log(\boldsymbol\xi) - \boldsymbol 1^\intercal\left(\boldsymbol\xi + \lngamma(\boldsymbol x + \boldsymbol 1)\right)\\
@@
 
-%WD does Kam not hate this latin? lulz
-\subsection{Tempora incognita: regularization toward the coalescent horizon}\label{sec:model:loss}
+\subsection{Regularization}\label{sec:model:reg}
 
@@
 To penalize oscillatory histories we regularize on their derivatives.
+We define weighted $L_p^p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$:
+\[
+\int_0^\infty\left|\frac{d\eta(t)}{dt}\right|^p dt, \int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt.
+\]
+With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and these become
+\[
+\left|\Delta\boldsymbol y\right|_p^p, \left|\Delta\boldsymbol z\right|_p^p.
+\]
+where $\Delta$ denotes the first difference operator.
+
+Our fused LASSO \citep{tibs} loss function is then
+\begin{align}
+L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\left|\Delta\boldsymbol y\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol y\right|_2^2\right) + \lambda_\mu \left(\alpha\left|\Delta\boldsymbol z\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol z\right|_2^2\right),
+\end{align}
+where $\lambda_\eta$ and $\lambda_\mu$ tune the regularization strength on $\eta$ and $\mu$, respectively, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
+
+\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
+
+Todo: SVD on $CM(\boldsymbol y)$
+
 Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
 Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
-%WD This is ad hoc, can we motivate it with a Bayesian argument?
-To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
-At times for which the CDF is near unity, the TMRCA has probably occurred, so we should penalize heavily because the data are not informative about the history at this time.
-At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
+
 From \eqref{eqn:pi} and \eqref{eqn:r} the CDF of $T_2$ is
@@
 \end{align}
-
-We define weighted $L_p^p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
-\begin{align}
-\rho_\eta[\eta(t); h, p] &\equiv \int_0^\infty\left(1+(h-1)F_2(t)\right)\left|\frac{d\eta(t)}{dt}\right|^p dt,\\
-\rho_\mu[\eta(t), \mu(t); h, p] &\equiv \int_0^\infty\left(1+(h-1)F_2(t)\right)\left|\frac{d\mu(t)}{dt}\right|^p dt.
-\end{align}
-With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and it is straightforward to show (using notation as in section \ref{sec:model:pcsws}) that these become
-\begin{align}
-\rho_\eta(\boldsymbol y, h, p) &= \sum_2^m w_i|y_i-y_{i-1}|^p,\\
-\rho_\mu(\boldsymbol y, \boldsymbol z, h, p) &= \sum_2^m w_i|z_i-z_{i-1}|^p.
-\end{align}
-with weights
-\begin{align}
-\label{eqn:w}
-\boldsymbol w(\boldsymbol y, h) \equiv
-h \boldsymbol 1 + (1-h)\times
-\left(
-\begin{bmatrix}
-A_{2,2} & A_{2,3} & \hdots & A_{2,n}
-\end{bmatrix}
-\times
-\begin{bmatrix}
-u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
-u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
-\vdots             & \ddots & \vdots                                \\
-u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
-\end{bmatrix}\right)^\intercal.
-\end{align}
-% Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.
-
-Our fused LASSO \citep{tibs} loss function is then
-\begin{align}
-L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\rho_\eta(\boldsymbol y, h, 1) + \frac{(1-\alpha)}{2}\rho_\eta(\boldsymbol y, h, 2)\right) + \lambda_\mu \left(\alpha\rho_\mu(\boldsymbol y, \boldsymbol z, h, 1) + \frac{(1-\alpha)}{2}\rho_\mu(\boldsymbol y, \boldsymbol z, h, 2)\right),
-\end{align}
-where $\lambda_\eta$ and $\lambda_\mu$ tune the regularization strength on $\eta$ and $\mu$, respectively, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
 

### 2019-09-02 mushi.py
simplifying

@@ -149,44 +149,8 @@ class SFS():
         if self.n < 2:
             raise ValueError('n must be larger than 1')
         self.C = SFS.C(self.n)
-        self.binom_vec = binom(np.arange(2, self.n + 1), 2)
-        self.binom_array_recip = np.diag(1 / self.binom_vec)
-
-    @staticmethod
-    def W1(n: int) -> np.ndarray:
-        '''The W1 matrix defined in the text
-
-        n: number of sampled haplotypes
-        '''
-        W1 = np.zeros((n - 1, n - 1))
-        b = np.arange(1, n - 1 + 1)
-        # j = 2
-        W1[:, 0] = 6 / (n + 1)
-        # j = 3
-        W1[:, 1] = 10 * (5 * n - 6 * b - 4) / (n + 2) / (n + 1)
-        for j in range(2, n - 1):
-            col = j - 2
-            # procedurally generated by Zeilberger's algorithm in Mathematica
-            W1[:, col + 2] = -((-((-1 + j)*(1 + j)**2*(3 + 2*j)*(j - n)*(4 + 2*j - 2*b*j + j**2 - b*j**2 + 4*n + 2*j*n + j**2*n)*W1[:, col]) - (-1 + 2*j)*(3 + 2*j)*(-4*j - 12*b*j - 4*b**2*j - 6*j**2 - 12*b*j**2 - 2*b**2*j**2 - 4*j**3 + 4*b**2*j**3 - 2*j**4 + 2*b**2*j**4 + 4*n + 2*j*n - 6*b*j*n + j**2*n - 9*b*j**2*n - 2*j**3*n - 6*b*j**3*n - j**4*n - 3*b*j**4*n + 4*n**2 + 6*j*n**2 + 7*j**2*n**2 + 2*j**3*n**2 + j**4*n**2)*W1[:, col + 1])/(j**2*(2 + j)*(-1 + 2*j)*(1 + j + n)*(3 + b + j**2 - b*j**2 + 3*n + j**2*n)))
-        return np.array(W1)
-
-    @staticmethod
-    def W2(n: int) -> np.ndarray:
-        '''The W2 matrix defined in the text
-
-        n: number of sampled haplotypes
-        '''
-        W2 = np.zeros((n - 1, n - 1))
-        b = np.arange(1, n - 1 + 1)
-        # j = 2
-        W2[:, 0] = 0
-        # j = 3
-        W2[:, 1] = (20 * (n - 2)) / (n+1) / (n+2)
[…]

### 2019-09-03 manuscript/main.tex
most math moved to Methods, and some better operator notation

@@
 \cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
-We develop coalescent theory for the dependence of the expected 3-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation rates as a function of time into the past).
-This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized, and extendable to joint inference of demographic and mutation spectrum histories.
-%WD do we actually do joint inference? If not, maybe save for discussion
+We develop coalescent theory for the dependence of the expected 3-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation intensities as a function of time into the past).
+This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized.
 
 Some references for nonidentifiability of demographic history from the SFS: \cite{Baharian2018-np, Bhaskar2014-fu, Terhorst2015-xt, Myers2008-jp}.
-It might be good to consider identifiability of mutation rate history from the SFS conditioned on demographic history.
+It might be good to consider identifiability of mutation intensity history from the SFS conditioned on demographic history.
+
+%WD todo: inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
+
+%WD todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
 
@@
 The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
-We will retrace the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
-We will make use of important results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} which facilitate numerical computation under this model, and formulate the expected 3-SFS under piecewise constant demographic and mutation spectrum histories building on the notation of \cite{Rosen2018-bb} for the expected SFS with a piecewise constant demographic history and constant mutation process.
+In the Methods we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
+We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
+Complete derivation of formulae are found in the Methods.
 
-\subsection{The expected SFS given demographic and mutation rate histories}\label{sec:model:xi}
+Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Methods section \ref{sec:methods:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
+\begin{equation}
+  \label{eqn:transform}
+\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
+\end{equation}
+where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
+It is dependent on $n$ and $\eta$, which we take to be fixed.
 
-Suppose $n$ haplotypes are sampled in the present, and let rvs $(T = T_2,\dots,T_n)$ denote the coalescent times measured retrospectively from the present (i.e.
-$T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
-As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
-This is tantamount to the infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
-%WD Kam suggests replacing below with $\mu^* \ge \mu(t) \ge 0$, but I don't get why
-Further suppose that mutation rate is not constant, but a specified function of time $0\le \mu(t)<\infty$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
-A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
-The following argument can be generalized to allow the labelling distribution to be nonuniform over the unit interval, but the same results follow.
+% Need to talk about this fig \ref{fig:model}.
 
@@
 \end{figure}
+
+For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
+We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
+Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
+In Methods section \ref{sec:methods:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
+\begin{equation}
+\label{eqn:transform_discrete}
+\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
+\end{equation}
+where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.
+
+In linkage equilibrium the log likelihood function of the mutation intensity history $\boldsymbol z$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
+\[
+\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
+\]
+where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $\boldsymbol z$.
+
+%WD more inverse problem intro
+The inverse problem \eqref{eqn:transform} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
+To penalize oscillatory histories we regularize on the derivative of $\mu$.
+We define $L_p^p$ derivative regularizers for history $\mu(t)$:
+\[
+\int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt.
+\]
+In the finite-dimensional approximation \eqref{eqn:transform_discrete} the derivatives become Dirac delta functions at the change points, and this become
+\[
+\left|\Delta\boldsymbol z\right|_p^p.
+\]
+where $\Delta$ denotes the first difference operator.
+Our penalized log-likelihood is then
+\begin{equation}
+\label{eqn:penalized}
+\tilde\ell(\boldsymbol z) = \ell(\boldsymbol z) - \lambda \left(\alpha\left|\Delta\boldsymbol z\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol z\right|_2^2\right),
+\end{equation}
+where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
+
+\section{Results}\label{sec:results}
+
+% Todo:
+% \begin{itemize}
+% \item repeat 1KG analysis like in \cite{Harris2017-fw}, see if we recapitulate Kelley's simulation-based pulse results
+% \item cluster triplet time series to see if we pull in minor components.
+% \end{itemize}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{figures/fit_teaser}
+  \caption{Simulating and inverting a pulse.}
+  \label{}
+\end{figure}
+
+\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
+
+Todo: SVD on $L_{n, \boldsymbol y}$
+
+Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
+Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
+
+From \eqref{eqn:pi} and \eqref{eqn:r} in the Methods, the CDF of $T_2$ is
+\begin{align}
+F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
+&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
+\end{align}
+
+\section{Discussion}\label{sec:discussion}
+
+Joint inference of $\eta$ and $\mu$.
+
+\section{Methods}\label{sec:methods}
+
+\subsection{The expected SFS given demographic and mutation intensity histories}\label{sec:methods:xi}
+
+Suppose $n$ haplotypes are sampled in the present, and let rvs $(T = T_2,\dots,T_n)$ denote the coalescent times measured retrospectively from the present (i.e.
+$T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
+As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
+This is tantamount to the infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
+%WD Kam suggests replacing below with $\mu^* \ge \mu(t) \ge 0$, but I don't get why
+Further suppose that mutation intensity is not constant, but a specified function of time $0\le \mu(t)<\infty$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
+A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
+The following argument can be generalized to allow the labelling distribution to be nonuniform over the unit interval, but the same results follow.
 
@@
 \label{eqn:xi}
-%WD Kam noted an "off by -1 error" below, but I think it's correct as is
 \xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
@@
 
-We then have a linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
-%WD Kam suggests a linear operator notation that combines the integral on t with the summation on j. Note this operator would be (nonlinearly) dependent on \eta
+We then have a linear expression for the expected SFS as a function of the mutation intensity history $\mu(t)$:
 \begin{equation}
@@
 is a linear functional of $\mu$ and a nonlinear functional of $\eta$.
+We recover \ref{eqn:transform} by defining the operator $\mathcal{L}_{n,\eta}$ such that $\mathcal{L}_{n,\eta}\mu \equiv C \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt$.
 
-\subsection{Computing the elements of $C$}\label{sec:model:C}
+\subsection{Computing the elements of $C$}\label{sec:methods:C}
 
@@
 \end{align*}
-This completes the framing of mutation rate inference as a linear inverse problem.
+This completes the framing of mutation intensity inference as a linear inverse problem.
 
-\subsection{Finite dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:model:pcsws}
+\subsection{Finite-dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:methods:pcsws}
 
@@
 where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
-We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
-We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite dimensional histories.
-Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
+We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:methods:xi}).
+We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
+Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
 Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
@@
 \begin{equation}
-\label{eqn:xivec2}
-\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z.
+\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z = L_{n, \boldsymbol y} \boldsymbol z,
 \end{equation}
+where $L\equiv C M(\boldsymbol y)$.
 
-\subsection{Poisson random field log likelihood}\label{sec:model:ell}
+\subsection{Implementation and pipeline}\label{sec:methods:tool}
 
-%WD Kam suggests calling the data vector \xi instead of x, but I disagree because we're already using \xi for the expected SFS.
-one could quibble that we already used x in the marked Poisson process at the beginning, but I think our readers have forgotten that by now and this won't confuse
-
-In linkage equilibrium the log likelihood function of the history $(\boldsymbol y, \boldsymbol z)$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ is the Poisson random field \citep{}
-\begin{align}
-\ell(\boldsymbol y, \boldsymbol z; \boldsymbol x) &= \sum_{b=1}^{n-1}\left(x_b\log(\xi_b) - \xi_b - \log(x_b!)\right)\\
-&= \boldsymbol x^\intercal\log(\boldsymbol\xi) - \boldsymbol 1^\intercal\left(\boldsymbol\xi + \lngamma(\boldsymbol x + \boldsymbol 1)\right)\\
-&= \boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) - \boldsymbol 1^\intercal\left(C M(\boldsymbol y) \boldsymbol z + \lngamma(\boldsymbol x + \boldsymbol 1)\right),
-\end{align}
-where $\log(\cdot)$ and the log-gamma function $\lngamma(\cdot)$ operate on vectors elementwise.
-
-\subsection{Regularization}\label{sec:model:reg}
-
-%WD more inverse problem intro
-The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
-To penalize oscillatory histories we regularize on their derivatives.
-We define weighted $L_p^p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$:
-\[
-\int_0^\infty\left|\frac{d\eta(t)}{dt}\right|^p dt, \int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt.
-\]
-With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and these become
-\[
-\left|\Delta\boldsymbol y\right|_p^p, \left|\Delta\boldsymbol z\right|_p^p.
-\]
-where $\Delta$ denotes the first difference operator.
-
-Our fused LASSO \citep{tibs} loss function is then
-\begin{align}
-L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \left(\alpha\left|\Delta\boldsymbol y\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol y\right|_2^2\right) + \lambda_\mu \left(\alpha\left|\Delta\boldsymbol z\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol z\right|_2^2\right),
-\end{align}
-where $\lambda_\eta$ and $\lambda_\mu$ tune the regularization strength on $\eta$ and $\mu$, respectively, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
-
-\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
-
-Todo: SVD on $CM(\boldsymbol y)$
-
-Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
-Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
-
-From \eqref{eqn:pi} and \eqref{eqn:r} the CDF of $T_2$ is
-\begin{align}
-F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
-&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
-\end{align}
-
-\section{Results}\label{sec:results}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{figures/fit_teaser}
-  \caption{Simulating and inverting a pulse.}
-  \label{}
-\end{figure}
-
-Todo:
-\begin{itemize}
-\item repeat 1KG analysis like in \cite{Harris2017-fw}, see if we recapitulate Kelley's simulation-based pulse results
-\item cluster triplet time series to see if we pull in minor components.
-\end{itemize}
-
-\section{Discussion}\label{sec:discussion}
+The \texttt{mushi} software package is available at \url{https://github.com/harrispopgen/mushi}.
+%WD cite whatever packages we use, e.g. proxtv
 

### 2019-09-03 mushi.py
most math moved to Methods, and some better operator notation

@@ -223,10 +223,10 @@ class SFS():
         jac: flag to return jacobian wrt μ
         '''
         z = history.μ.vals
-        CM = self.C @ self.M(history.η)
+        L = self.C @ self.M(history.η)
         if jac:
-            return CM @ z, CM
-        return CM @ z
+            return L @ z, L
+        return L @ z
 
     def simulate(self, history: JointHistory, seed: int = None) -> None:
         '''simulate a SFS under the Poisson random field model (no linkage)

### 2019-09-03 manuscript/appendix.tex
better yet, an appendix

--- 2019-09-03 manuscript/main.tex
@@
-\documentclass[11pt]{article}
+\section{Appendix}\label{sec:appendix}
 
-\usepackage{fullpage}
-\usepackage[round,semicolon]{natbib}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{bbm}
-\usepackage{graphicx}
-\usepackage{hyperref}
-
-\DeclareMathOperator{\erf}{erf}
-\DeclareMathOperator{\lngamma}{\ln\!\Gamma}
-
-% NOTE
-% one sentence per line for nice git diffs
-% WD: I'm a fan of initialed source comments like this
-
-\title{Coalescent inference of mutation spectrum histories from sample frequency spectra}
-
-\author{
-W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
-\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}
-% \small{$^\ast$ Equal contribution}
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-Single nucleotide variant (SNV) frequencies partitioned according to triplet nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
-The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
-We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the triplet-resolved sample frequency spectrum (3-SFS).
-We formulate inference of mutation spectrum histories from the 3-SFS as a linear inverse problem.
-
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-\cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
-We develop coalescent theory for the dependence of the expected 3-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation intensities as a function of time into the past).
-This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized.
-
-Some references for nonidentifiability of demographic history from the SFS: \cite{Baharian2018-np, Bhaskar2014-fu, Terhorst2015-xt, Myers2008-jp}.
-It might be good to consider identifiability of mutation intensity history from the SFS conditioned on demographic history.
-
-%WD todo: inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
-
-%WD todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
-
-\section{Model}\label{sec:model}
-
-The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
-In the Methods we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
-We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
-Complete derivation of formulae are found in the Methods.
-
-Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Methods section \ref{sec:methods:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
-\begin{equation}
-  \label{eqn:transform}
-\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
-\end{equation}
-where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
-It is dependent on $n$ and $\eta$, which we take to be fixed.
-
-% Need to talk about this fig \ref{fig:model}.
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{figures/model}
-\caption{Schematic of the marked Poisson process model with $n=4$.
-We condition on coalescent times $T_4=t_4,T_3=t_3,T_2=t_2$ and consider mutation intensity function $\mu(t)$.
-Red dots indicate mutation events placed by time $t$, genomic position $x$, and coalescent line (which are depicted as extruded in the genomic coordinate axis, grey sheets).
-The probability that a differential element $dxdt$ on a given sheet contains a mutation is proportional to the instantaneous mutation intensity $\mu(t)$.
-}
-\label{fig:model}
-\end{figure}
-
-For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
-We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
-Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
-In Methods section \ref{sec:methods:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
-\begin{equation}
-\label{eqn:transform_discrete}
-\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
-\end{equation}
-where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.
-
-In linkage equilibrium the log likelihood function of the mutation intensity history $\boldsymbol z$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
-\[
-\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
-\]
-where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $\boldsymbol z$.
-
-%WD more inverse problem intro
-The inverse problem \eqref{eqn:transform} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
-To penalize oscillatory histories we regularize on the derivative of $\mu$.
-We define $L_p^p$ derivative regularizers for history $\mu(t)$:
-\[
-\int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt.
-\]
-In the finite-dimensional approximation \eqref{eqn:transform_discrete} the derivatives become Dirac delta functions at the change points, and this become
-\[
-\left|\Delta\boldsymbol z\right|_p^p.
-\]
-where $\Delta$ denotes the first difference operator.
-Our penalized log-likelihood is then
-\begin{equation}
-\label{eqn:penalized}
-\tilde\ell(\boldsymbol z) = \ell(\boldsymbol z) - \lambda \left(\alpha\left|\Delta\boldsymbol z\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol z\right|_2^2\right),
-\end{equation}
-where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
-
-\section{Results}\label{sec:results}
-
-% Todo:
-% \begin{itemize}
-% \item repeat 1KG analysis like in \cite{Harris2017-fw}, see if we recapitulate Kelley's simulation-based pulse results
-% \item cluster triplet time series to see if we pull in minor components.
-% \end{itemize}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{figures/fit_teaser}
-  \caption{Simulating and inverting a pulse.}
-  \label{}
-\end{figure}
-
-\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
-
-Todo: SVD on $L_{n, \boldsymbol y}$
-
-Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
-Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
-
-From \eqref{eqn:pi} and \eqref{eqn:r} in the Methods, the CDF of $T_2$ is
-\begin{align}
-F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
-&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
-\end{align}
-
-\section{Discussion}\label{sec:discussion}
-
-Joint inference of $\eta$ and $\mu$.
-
-\section{Methods}\label{sec:methods}
-
-\subsection{The expected SFS given demographic and mutation intensity histories}\label{sec:methods:xi}
+\subsection{The expected SFS given demographic and mutation intensity histories}\label{sec:appendix:xi}
 
@@
 
-\subsection{Computing the elements of $C$}\label{sec:methods:C}
+\subsection{Computing the elements of $C$}\label{sec:appendix:C}
 
@@
 
-\subsection{Finite-dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:methods:pcsws}
+\subsection{Finite-dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:appendix:pcsws}
 
@@
 where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
-We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:methods:xi}).
+We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:appendix:xi}).
 We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
@@
 \end{equation}
-where $L\equiv C M(\boldsymbol y)$.
-
-\subsection{Implementation and pipeline}\label{sec:methods:tool}
-
-The \texttt{mushi} software package is available at \url{https://github.com/harrispopgen/mushi}.
-%WD cite whatever packages we use, e.g. proxtv
-
-\section{Acknowledgements}\label{sec:ack}
-
-\bibliographystyle{plainnat}
-\bibliography{refs}
-
-\end{document}
+where $L_{n, \boldsymbol y}\equiv C M(\boldsymbol y)$.

### 2019-09-03 manuscript/main.tex
better yet, an appendix

@@
 
-\section{Introduction}\label{sec:intro}
+\section*{Introduction}\label{sec:intro}
 
@@
 
-\section{Model}\label{sec:model}
+\section*{Model}\label{sec:model}
 
@@
 We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
-Complete derivation of formulae are found in the Methods.
+Complete derivation of formulae are found in the Appendix.
 
-Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Methods section \ref{sec:methods:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
+Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Appendix \ref{sec:appendix:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
 \begin{equation}
@@
 Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
-In Methods section \ref{sec:methods:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
+In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
 \begin{equation}
@@
 
-\section{Results}\label{sec:results}
+\section*{Results}\label{sec:results}
 
@@
 
-\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
+\subsection*{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}
 
@@
 \begin{align}
-F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
+F_2(t) &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
 &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
@@
 
-\section{Discussion}\label{sec:discussion}
+\section*{Discussion}\label{sec:discussion}
 
@@
 
-\section{Methods}\label{sec:methods}
+\section*{Methods}\label{sec:methods}
 
-\subsection{The expected SFS given demographic and mutation intensity histories}\label{sec:methods:xi}
-
-Suppose $n$ haplotypes are sampled in the present, and let rvs $(T = T_2,\dots,T_n)$ denote the coalescent times measured retrospectively from the present (i.e.
-$T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
-As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
-This is tantamount to the infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
-%WD Kam suggests replacing below with $\mu^* \ge \mu(t) \ge 0$, but I don't get why
-Further suppose that mutation intensity is not constant, but a specified function of time $0\le \mu(t)<\infty$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
-A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
-The following argument can be generalized to allow the labelling distribution to be nonuniform over the unit interval, but the same results follow.
-
-Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation present in $b\in\{1, 2, \dots, n-1\}$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (the probability of this event is independent of $x$, given the uniform labeling distribution).
-Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
-Let rv $\mathcal{E}_{\delta x, b, k}$ denote the event that the mutation $\mathcal{E}_{\delta x, b}$ occurred during $I_k$.
-For small $\delta x$ and finite $\mu(t)$ we have
-\begin{align*}
-\mathbbm{P}(\mathcal{E}_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(\mathcal{E}_{\delta x, b, k}\mid T)\\
-&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O((\delta x)^2)\right),
-\end{align*}
-where
-\begin{equation}
-\label{eqn:p}
-p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
-\end{equation}
-is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parentheses is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
-Marginalizing $T$ gives
-\begin{align*}
-\mathbbm{P}(\mathcal{E}_{\delta x, b}) &= \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right] + O((\delta x)^2).
-\end{align*}
-For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
-\[
-\xi_b = \int_0^1 \mathbbm{P}(\mathcal{E}_{dx, b})dx = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
-\]
-Using the explicit bounds of the intercoalescent intervals, $I_{k} = (T_{k+1}, T_k)$, gives
-\begin{align}
-\label{eqn:xi}
-\xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
-&= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=3}^{n} (k-1) p_{n,k-1}(b) \mathbbm{E}_{T_{k}}\left[\int_0^{T_k}\mu(t)dt\right]\nonumber\\
-&= \sum_{k=2}^n B_{b,k} \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right],
-\end{align}
-where
-\begin{equation}
-\label{eqn:B}
-B_{b,k}\equiv
-\begin{cases}
-k p_{n,k}(b),& k=2\\
-k p_{n,k}(b) - (k-1) p_{n,k-1}(b),& k > 2
-\end{cases}
-\end{equation}
-
-\cite{Polanski2003-kg} (eqns 5-8) give the marginal density for the coalescent time $T_k$ as
-\begin{equation}
-\label{eqn:pi}
-\pi_k(t_k) = \sum_{j=k}^n A_{k,j} q_j(t_k)
-\end{equation}
-where
-\begin{align*}
-A_{k,j} &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
-A_{n,n} &\equiv 1,\\
-q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
-\end{align*}
-and $\eta(t)$ is the haploid effective population size history.
-Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present, with inhomogeneous Poisson intensity function $\binom{j}{2}/\eta(t)$.
-
-The expectations in \eqref{eqn:xi} can be expressed using \eqref{eqn:pi} as
-\begin{align}
-\label{eqn:exp}
-\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] &= \int_0^\infty\pi_k(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
-&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
-&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^\infty 1_{[0<t<t_k]}(t)\mu(t)dt dt_k\nonumber\\
-&= \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt
-\end{align}
-where in the last line we've exchanged integration order and defined the inhomogeneous Poisson survival function
-\begin{equation}
-\label{eqn:r}
-r_j(t) \equiv \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]
-\end{equation}
-corresponding to density $q_j(t)$ (that this is the appropriate survival function can be seen by considering the limit of a sequence of Bernoulli failures).
-
-Using \eqref{eqn:exp} in \eqref{eqn:xi} gives
-\begin{align}
-\label{eqn:xi2}
-\xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt\nonumber\\
-&= \sum_{j=2}^n \left(\sum_{k=2}^j B_{b,k} A_{k,j}\right) \int_0^\infty r_j(t)\mu(t)dt,
-\end{align}
-where we've exchanged summation order in the last line.
-
-We then have a linear expression for the expected SFS as a function of the mutation intensity history $\mu(t)$:
-\begin{equation}
-\label{eqn:xivec}
-\boldsymbol\xi = C \boldsymbol d,
-\end{equation}
-where the $(n-1)\times(n-1)$ matrix
-\[
-C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
-\]
-is constant wrt $\mu$ \emph{and} $\eta$, and
-\begin{equation}
-\label{eqn:d}
-d_j \equiv \int_0^\infty r_j(t)\mu(t)dt = \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt
-\end{equation}
-is a linear functional of $\mu$ and a nonlinear functional of $\eta$.
-We recover \ref{eqn:transform} by defining the operator $\mathcal{L}_{n,\eta}$ such that $\mathcal{L}_{n,\eta}\mu \equiv C \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt$.
-
-\subsection{Computing the elements of $C$}\label{sec:methods:C}
-
-%WD Kam suggests just doing C = B A, but computing the elements of B and A would seem computationally infeasible
-We next develop an efficient recursive procedure for computing the $(n-1)\times(n-1)$ matrix $C$.
-Using \eqref{eqn:B}
-\begin{align*}
-C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
-&= W_{b,j}^{(1)} - W_{b,j}^{(2)},
-\end{align*}
-where
-\begin{align}
-\label{eqn:W1}
-W_{b,j}^{(1)} &\equiv \sum_{k=2}^j k p_{n,k}(b) A_{k,j}\\
-\label{eqn:W2}
-W_{b,j}^{(2)} &\equiv \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
-\end{align}
-\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
-\[
-A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
-\]
-so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \eqref{eqn:W1} and \eqref{eqn:W2} are definite sums over geometric terms.
-Zeilberger's algorithm, which finds polynomial recurrences for definite sums of hypergeometric terms \citep{petkovvsek1996b, paule1995mathematica}, can thus be used to yield the following procedurally generated second-order recursions in $j$:
-%WD need to wrap this eqn, as it runs will beyond the margin. Also make an appendix at some point for these
-\begin{align*}
-&W_{b,2}^{(1)} = \frac{6}{(n+1)}\\
-&W_{b,3}^{(1)} = \frac{10(5n-6b-4)}{(n+2)(n+1)}\\
-&W_{b,j+2}^{(1)} = -\frac{(2 j+3) \left(-(2 j-1) W_{b,j+1}^{(1)}  \left(2 j (j+1) \left(b^2 \left(j^2+j-2\right)-6 b-j (j+1)-2\right)-j (j+1) n \left(3 b \left(j^2+j+2\right)+j^2+j-2\right)+\left(j (j+1) \left(j^2+j+6\right)+4\right) n^2+4 n\right)-(j-1) (j+1)^2 (j-n) W_{b,j}^{(1)}  (4 (n+1)-j (j+2) (b-n-1))\right)}{j^2 (j+2) (2 j-1) (j+n+1) \left(-b j^2+b+\left(j^2+3\right) (n+1)\right)}
-\end{align*}
-and
-\begin{align*}
-&W_{b,2}^{(2)} = 0\\
-&W_{b,3}^{(2)} = \frac{20 (n-2)}{(n+1)(n+2)}\\
-&W_{b,j+2}^{(2)} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W_{b,j}^{(2)}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W_{b,j+1}^{(2)}\right)
-\end{align*}
-This completes the framing of mutation intensity inference as a linear inverse problem.
-
-\subsection{Finite-dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:methods:pcsws}
-
-%WD Kam describes this discretization as ad hoc, but I'm pretty sure this is your textbook Riemann sum with rectangular partitions.
-We could do fancier quadrature, but then we'll be less able to lean on previous work (i.e.
-Rosen et al.)
-
-Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
-\begin{equation}
-\label{eqn:d2}
-d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)d\tau,
-\end{equation}
-where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
-We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:methods:xi}).
-We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
-Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
-Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
-%WD Kam hates the latin loooooooool. It really does follow by dropping in the y vector and proceeding as in their proof
-With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
-\begin{equation}
-\label{eqn:d3}
-\boldsymbol d = M(\boldsymbol y) \boldsymbol z
-\end{equation}
-where
-\begin{equation}
-\label{eqn:M}
-M(\boldsymbol y) \equiv
-\begin{bmatrix}
-1 &             &        &                       \\
-  & \frac{1}{3} &        &                       \\
-  &             & \ddots &                       \\
-  &             &        & \frac{1}{\binom{n}{2}}
-\end{bmatrix}
-\begin{bmatrix}
-1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
-1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
-\vdots  & \vdots             & \ddots & \vdots                                \\
-1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
-\end{bmatrix}
-\begin{bmatrix}
-1  &      &        &             &       \\
--1 & 1  &        &             &       \\
-     & -1 & 1    &             &       \\
-     &      & \ddots & \ddots      &       \\
-     &      &        & -1 & 1
-\end{bmatrix}
-\begin{bmatrix}
-y_1 &     &      &             &       \\
-    & y_2 &      &             &       \\
-     &     & y_3 &             &       \\
-     &      &    & \ddots      &       \\
-     &      &        &  & y_m
-\end{bmatrix}.
-\end{equation}
-Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.
-
-Combining \ref{eqn:d3} with \ref{eqn:xivec} gives the discretized inverse problem
-\begin{equation}
-\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z = L_{n, \boldsymbol y} \boldsymbol z,
-\end{equation}
-where $L\equiv C M(\boldsymbol y)$.
-
-\subsection{Implementation and pipeline}\label{sec:methods:tool}
+\subsection*{Implementation and pipeline}\label{sec:methods:tool}
 
@@
 
-\section{Acknowledgements}\label{sec:ack}
+\section*{Acknowledgements}\label{sec:ack}
 
@@
 
+\appendix
+\input{appendix}
+
 \end{document}

### 2019-09-03 manuscript/main.tex
changed writing that explains the cost function

@@
 \usepackage{amsmath}
+\usepackage{bm}
 \usepackage{bbm}
 \usepackage{graphicx}
+\usepackage{color}
 \usepackage{hyperref}
@@
 \author{
-W. DeWitt$^{1}$, K. Harris$^{2}$, and K. Harris$^{1}$\\
-\small{Departments of $^1$Genome Sciences and $^2$Biology, University of Washington, Seattle, USA}
+W. DeWitt$^{1}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1}$\\
+{\small
+$^1$Genome Sciences,
+$^2$Paul G.\ Allen School of Computer Science \& Engineering, 
+and $^3$Biology}\\
+{\small University of Washington, Seattle, USA}
 % \small{$^\ast$ Equal contribution}
@@
 The inverse problem \eqref{eqn:transform} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
-To penalize oscillatory histories we regularize on the derivative of $\mu$.
-We define $L_p^p$ derivative regularizers for history $\mu(t)$:
+In order to recover smooth histories, 
+we penalize the $L^p$ norms
+of the derivative of $\mu$ 
 \[
-\int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt.
+\left\| \frac{d \mu}{d t} \right\|_p^p 
+= \int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt,
 \]
-In the finite-dimensional approximation \eqref{eqn:transform_discrete} the derivatives become Dirac delta functions at the change points, and this become
+for $p=1$ and 2.
+With time discretized \eqref{eqn:transform_discrete},
+this becomes
 \[
-\left|\Delta\boldsymbol z\right|_p^p.
+\left\|\Delta\boldsymbol z \right\|_p^p.
 \]
-where $\Delta$ denotes the first difference operator.
-Our penalized log-likelihood is then
+where $\Delta$ denotes the first difference matrix.
+Our cost function (penalized log-likelihood) is then
 \begin{equation}
 \label{eqn:penalized}
-\tilde\ell(\boldsymbol z) = \ell(\boldsymbol z) - \lambda \left(\alpha\left|\Delta\boldsymbol z\right|_1 + \frac{(1-\alpha)}{2}\left|\Delta\boldsymbol z\right|_2^2\right),
+C(\bm z)
+%\tilde\ell(\boldsymbol z) 
+= -\ell(\boldsymbol z) + \lambda \left(\alpha\left\|\Delta\boldsymbol z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta\boldsymbol z\right\|_2^2\right),
 \end{equation}
 where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
+\textcolor{blue}{Kam note: probably we need a ``Ridge'' (Tikhonov) 
+term like the following to deal with small singular values:}
+\[
+\lambda' \| \bm z \|_2^2
+\]
 

### 2019-09-03 manuscript/main.tex
more verbose conditioning

@@
 \author{
-W. DeWitt$^{1}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1}$\\
+W.S. DeWitt$^{1,4}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1}$\\
 {\small
-$^1$Genome Sciences,
-$^2$Paul G.\ Allen School of Computer Science \& Engineering, 
-and $^3$Biology}\\
-{\small University of Washington, Seattle, USA}
+$^1$Department of Genome Sciences,
+$^2$Paul G.\ Allen School of Computer Science \& Engineering,}\\
+{\small and $^3$Department of Biology, University of Washington, Seattle, WA}\\
+{\small $^4$Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
 % \small{$^\ast$ Equal contribution}
@@
 \[
-\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
+\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y, n) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
 \]
@@
 The \texttt{mushi} software package is available at \url{https://github.com/harrispopgen/mushi}.
-%WD cite whatever packages we use, e.g. proxtv
+%WD cite whatever packages we use, e.g. proxtv, msprime, stdpopsim, etc.
 

### 2019-09-04 mushi.py
placing msprime mutations accordind to mushi \mu(t)

@@ -70,6 +70,22 @@ class PiecewiseConstantHistory():
                             np.array([np.inf])))
         return t, self.vals
 
+    def epochs(self):
+        '''generator yielding epochs in history as tuples:
+        (start_time, end_time, value)
+        '''
+        for i in range(self.m()):
+            if i == 0:
+                start_time = 0
+            else:
+                start_time = self.change_points[i - 1]
+            if i == self.m() - 1:
+                end_time = np.inf
+            else:
+                end_time = self.change_points[i]
+            value = self.vals[i]
+            yield (start_time, end_time, value)
+
 @dataclass()
 class JointHistory():
     '''Piecewise constant history of population size η and mutation rate μ.

### 2019-09-05 mushi.py
tennesen model in notebook

@@ -112,24 +112,6 @@ class JointHistory():
         '''
         return self.η == other.η and self.μ == other.μ
 
-    def plot(self, **kwargs) -> figure.Figure:
-        '''plot the history
-
-        fig: add to another figure
-        kwargs: keyword arguments passed to plotting calls
-        '''
-        fig, axes = plt.subplots(2, 1, sharex=True, figsize=(6, 6))
-        plt.sca(axes[0])
-        self.η.plot(**kwargs)
-        plt.ylabel('$η(t)$')
-
-        plt.sca(axes[1])
-        self.μ.plot(**kwargs)
-        plt.xlabel('$t$')
-        plt.ylabel('$μ(t)$')
-
-        return fig
-
     def arrays(self):
         t = np.concatenate((np.array([0]),
                             self.μ.change_points,
@@ -251,7 +233,6 @@ class SFS():
 
     def ℓ(self, history: JointHistory, grad: bool = False) -> np.float:
         '''Poisson random field log-likelihood of history
-        ignores constant term
 
         history: η and μ joint history
         grad: flag to return gradient wrt μ
@@ -260,11 +241,45 @@ class SFS():
             raise ValueError('use simulate() to generate data first')
         if grad:
             ξ, J_μξ = self.ξ(history, jac=True)
-            ℓ = poisson.logpmf(self.x, ξ).sum()
-            dℓdμ = (((self.x / ξ)[:, np.newaxis] - 1) * J_μξ).sum(axis=0)
-            return ℓ, dℓdμ
+            dℓdμ = J_μξ.T @ (self.x / ξ - 1)
+            return dℓdμ
         else:
             return poisson.logpmf(self.x, self.ξ(history)).sum()
+
+    def d_kl(self, history: JointHistory, grad: bool = False) -> float:
+        '''Kullback-Liebler divergence between normalized SFS and its
+        expectation under history
+        ignores constant term
+
+        history: η and μ joint history
+        grad: flag to return gradient wrt μ
+        '''
+        if self.x is None:
+            raise ValueError('use simulate() to generate data first')
+        x_normalized = self.x / self.x.sum()
+        if grad:
+            ξ, J_μξ = self.ξ(history, jac=True)
+            ξ_normalized = ξ / ξ.sum()
+            return -J_μξ.T @ ((x_normalized / ξ) * (1 - ξ_normalized))
+        else:
+            ξ = self.ξ(history)
[…]

### 2019-09-05 manuscript/main.tex
Methods->Appendix

@@
 The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
-In the Methods we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
+In Appendix \ref{sec:appendix} we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
 We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
@@
 
-From \eqref{eqn:pi} and \eqref{eqn:r} in the Methods, the CDF of $T_2$ is
+From \eqref{eqn:pi} and \eqref{eqn:r} in the Appendix \ref{sec:appendix}, the CDF of $T_2$ is
 \begin{align}

### 2019-09-09 manuscript/main.tex

@@
 The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
-We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the triplet-resolved sample frequency spectrum (3-SFS).
-We formulate inference of mutation spectrum histories from the 3-SFS as a linear inverse problem.
+We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the sample frequency spectrum resolved by $k$-mer nucleotide context ($k$-SFS).
+We formulate inference of mutation spectrum histories from the $k$-SFS as a linear inverse problem.
 

### 2019-09-10 mushi.py
Frobenius

@@ -45,6 +45,7 @@ class History():
         else:
             vals = self.vals
         plt.step(t, vals, where='post', **kwargs)
+        plt.xlabel('$t$')
         if 'label' in kwargs:
             plt.legend()
 
@@ -233,12 +234,14 @@ class kSFS():
                        z0[np.newaxis, :] * np.ones((self.η.m, 1)))
 
     def infer_μ(self, λ: np.float64 = 0, α: np.float64 = .99,
+                λ_F: np.float64 = 0,
                 s: np.float64 = .01, steps: int = 100, fit='prf',
                 bins: np.ndarray = None) -> History:
         '''return inferred μ history given the sfs and η history
 
-        λ: regularization strength
-        α: relative penalty on L1 vs L2
+        λ: LASSO regularization strength
+        α: relative penalty on L1 vs L2 in LASSO
+        λ_F: Frobenius norm regularization strength
         s: step size parameter for proximal gradient descent
         steps: number of proximal gradient descent steps
         fit: fit function, 'prf' for Poisson random field, 'kl' for
@@ -281,7 +284,7 @@ class kSFS():
         μ = self.constant_μ_MLE()
         for _ in range(steps):
             Z = μ.vals
-            G = misfit_func(μ, grad=True) + λ * (1 - α) * D2 @ Z
+            G = misfit_func(μ, grad=True) + λ * (1 - α) * D2 @ Z + λ_F * Z
             if not np.all(np.isfinite(G)):
                 raise RuntimeError(f'invalid gradient: {G}')
             Z = Z - s * G

### 2019-09-10 manuscript/main.tex
promote vectors to spectrum matrices in TeX

@@
 \DeclareMathOperator{\lngamma}{\ln\!\Gamma}
+\DeclareMathOperator{\Tr}{Tr}
 
@@
   \label{eqn:transform}
-\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
+\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu(t),
 \end{equation}
@@
 
-For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
+In previous work \citep{Kelley, cancer, etc}, mutation spectra have been parameterized by grouping variants into mutation types according to local $k$-mer nucleotide context, and the number of variants called in each mutation type is taken as a proxy integrating an individual's ancestral mutational input in each type.
+With $k$ odd and the variant registered in the central position there are $\kappa_k=2\times3\times4^{k-1}$ mutation types after collapsing by strand symmetry.
+For example, there are $96$ triplet mutation types ($k=3$).
+Here, we use the same $k$-mer mutation type parameterization to promote the $(n-1)$-element expected SFS vector $\xi$ to the $(n-1)\times\kappa_k$ expected $k$-SFS matrix $\Xi$.
+Similarly, the mutation intensity history function $\mu(t)$ is promoted to the $\kappa_k$-element mutation spectrum history $\boldsymbol\mu(t)$, with each element giving the mutation intensity history function for one mutation type.
+Equation \eqref{eqn:transform} becomes
+\begin{equation}
+  \label{eqn:transform_matrix}
+\Xi = \mathcal{L}_{n,\eta}\boldsymbol\mu^\intercal(t),
+\end{equation}
+
+For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\boldsymbol\mu(t)$ as piecewise constant functions of time on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
 We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
-Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
-In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
+Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let the $m\times\kappa_k$ matrix $Z$ denote the constant mutation intensity during each epoch (rows) for each mutation type (columns).
+In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform_matrix} reduces to the matrix equation
 \begin{equation}
 \label{eqn:transform_discrete}
-\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
+\Xi = L_{n, \boldsymbol y} Z,
 \end{equation}
@@
 
-In linkage equilibrium the log likelihood function of the mutation intensity history $\boldsymbol z$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
+In linkage equilibrium the log likelihood function of the mutation spectrum history $Z$ given an observed $k$-SFS matrix $X$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
 \[
-\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y, n) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
+\ell(Z; X, \boldsymbol y, n) = \Tr(X^\intercal\log(L_{n, \boldsymbol y} Z)) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} Z\boldsymbol 1,
 \]
-where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $\boldsymbol z$.
+where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $Z$.
 
 %WD more inverse problem intro
-The inverse problem \eqref{eqn:transform} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
+The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
 In order to recover smooth histories,
 we penalize the $L^p$ norms
-of the derivative of $\mu$
+of the time derivative of $\boldsymbol\mu(t)$
 \[
-\left\| \frac{d \mu}{d t} \right\|_p^p
-= \int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt,
+\sum_{i=1}^{\kappa_k}\left\| \frac{d \mu_i(t)}{d t} \right\|_p^p
+= \sum_{i=1}^{\kappa_k}\int_0^\infty\left|\frac{d\mu_i(t)}{dt}\right|^p dt,
 \]
-for $p=1$ and 2.
-With time discretized \eqref{eqn:transform_discrete},
-this becomes
+for $p=1,2$.
+With piecewise constant histories as in \eqref{eqn:transform_discrete} this becomes
 \[
-\left\|\Delta\boldsymbol z \right\|_p^p.
+\left\|\Delta Z \right\|_p^p.
 \]
@@
 \label{eqn:penalized}
-C(\bm z)
+C(Z)
 %\tilde\ell(\boldsymbol z)
-= -\ell(\boldsymbol z) + \lambda \left(\alpha\left\|\Delta\boldsymbol z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta\boldsymbol z\right\|_2^2\right),
+= -\ell(Z) + \lambda \left(\alpha\left\|\Delta Z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta Z\right\|_2^2\right),
 \end{equation}
@@
 \[
-\lambda' \| \bm z \|_2^2
+\lambda' \| Z \|_2^2
 \]

### 2019-09-11 mushi.py
spectral regularization and Armijo line search

@@ -8,7 +8,6 @@ from scipy.stats import poisson
 from matplotlib import pyplot as plt
 import prox_tv as ptv
 
-
 @dataclass
 class History():
     '''Piecewise constant history. The first epoch starts at zero, and the last
@@ -177,52 +176,57 @@ class kSFS():
         np.random.seed(seed)
         self.X = poisson.rvs(self.L @ μ.vals)
 
-    def ℓ(self, μ: History, grad: bool = False) -> np.float:
+    def ℓ(self, Z: np.ndarray, grad: bool = False) -> np.float:
         '''Poisson random field log-likelihood of history
 
-        μ: mutation spectrum history
-        grad: flag to return gradient wrt μ
+        Z: mutation spectrum history matrix (μ.vals)
+        grad: flag to also return gradient wrt Z
         '''
         if self.X is None:
             raise ValueError('use simulate() to generate data first')
-        Ξ = self.L @ μ.vals
+        Ξ = self.L @ Z
+        ℓ = poisson.logpmf(self.X, Ξ).sum()
         if grad:
-            dℓdμ = self.L.T @ (self.X / Ξ - 1)
-            return dℓdμ
+            dℓdZ = self.L.T @ (self.X / Ξ - 1)
+            return np.array([ℓ, dℓdZ])
         else:
-            return poisson.logpmf(self.X, Ξ).sum(axis=0)
+            return ℓ
 
-    def d_kl(self, μ: History, grad: bool = False) -> float:
+    def d_kl(self, Z: np.ndarray, grad: bool = False) -> float:
         '''Kullback-Liebler divergence between normalized SFS and its
         expectation under history
         ignores constant term
 
-        μ: mutation spectrum history
-        grad: flag to return gradient wrt μ
+        Z: mutation spectrum history matrix (μ.vals)
+        grad: flag to also return gradient wrt Z
         '''
         if self.X is None:
             raise ValueError('use simulate() to generate data first')
         X_normalized = self.X / self.X.sum(axis=0)
-        Ξ = self.L @ μ.vals
+        Ξ = self.L @ Z
         Ξ_normalized = Ξ / Ξ.sum(axis=1)
+        d_kl = (-X_normalized * np.log(Ξ_normalized)).sum()
         if grad:
-            return -self.L.T @ ((X_normalized / Ξ) * (1 - Ξ_normalized))
[…]

### 2019-09-12 mushi.py
some :bug: :gun: and adding l2 penalty on singular values

@@ -8,6 +8,7 @@ from scipy.stats import poisson
 from matplotlib import pyplot as plt
 import prox_tv as ptv
 
+
 @dataclass
 class History():
     '''Piecewise constant history. The first epoch starts at zero, and the last
@@ -237,23 +238,26 @@ class kSFS():
         return History(self.η.change_points,
                        z0[np.newaxis, :] * np.ones((self.η.m, 1)))
 
-    def infer_μ(self, λ: np.float64 = 0, α: np.float64 = .99,
-                β: np.float64 = 0, γ: np.float64 = 0.8,
-                steps: int = 1000, tol: np.float64 = 1e-4, fit='prf',
-                bins: np.ndarray = None) -> History:
+    def infer_μ(self, λ_tv: np.float64 = 0, α_tv: np.float64 = 0,
+                λ_r: np.float64 = 0, α_r: np.float64 = 0,
+                γ: np.float64 = 0.8, steps: int = 1000, tol: np.float64 = 1e-4,
+                fit='prf', bins: np.ndarray = None) -> History:
         '''return inferred μ history given the sfs and η history
 
-        λ: fused LASSO regularization strength
-        α: relative penalty on L1 vs L2 in fused LASSO
-        β: spectral regularization strength
+        λ_tv: fused LASSO regularization strength
+        α_tv: relative penalty on L1 vs L2 in fused LASSO
+        λ_r: spectral (rank) regularization strength
+        α_r: relative penalty on L1 vs L2 in spectral regularization
         γ: step size shrinkage rate for line search
         steps: number of proximal gradient descent steps
         tol: relative tolerance in objective function
         fit: fit function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
         '''
-        assert λ >= 0, 'λ must be nonnegative'
-        assert 0 <= α <= 1, 'α must be in the unit interval'
+        assert λ_tv >= 0, 'λ_tv must be nonnegative'
+        assert λ_r >= 0, 'λ_r must be nonnegative'
+        assert 0 <= α_tv <= 1, 'α_tv must be in the unit interval'
+        assert 0 <= α_r <= 1, 'α_r must be in the unit interval'
         self.bins = bins
         if bins is not None:
             bin_idxs = np.digitize(np.arange(self.n - 1), bins=bins)
@@ -281,23 +285,21 @@ class kSFS():
             misfit_func = self.lsq
         else:
             raise ValueError(f'unrecognized fit argument {fit}')
[…]

### 2019-09-12 mushi.py
classier inheritance, and heatmaps for matrix vizualization

@@ -6,13 +6,14 @@ import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson
 from matplotlib import pyplot as plt
+from matplotlib.colors import SymLogNorm
 import prox_tv as ptv
 
 
 @dataclass
 class History():
-    '''Piecewise constant history. The first epoch starts at zero, and the last
-    epoch extends to infinity. Can be used for η or μ.
+    '''base class piecewise constant history. The first epoch starts at zero,
+    and the last epoch extends to infinity
 
     change_points: epoch change points (times)
     vals: constant values for each epoch (rows)
@@ -30,28 +31,12 @@ class History():
                              f' implies {len(self.change_points) + 1} epochs,'
                              f' but len(vals) = {len(self.vals)}')
         if np.any(self.vals <= 0) or np.sum(np.isinf(self.vals)):
-            raise ValueError('elements of vals must be finite and positive')
+            raise ValueError(f'elements of vals must be finite and '
+                             'positive')
         self.m = len(self.vals)
 
-    def plot(self, i=None, **kwargs) -> None:
-        '''plot the history
-
-        i: value column to plot (optional)
-        kwargs: key word arguments passed to plt.step
-        '''
-        t = np.concatenate((np.array([0]), self.change_points))
-        if i is not None:
-            vals = self.vals[:, i]
-        else:
-            vals = self.vals
-        plt.step(t, vals, where='post', **kwargs)
-        plt.xlabel('$t$')
-        if 'label' in kwargs:
-            plt.legend()
-
     def arrays(self):
-        '''return time grid and values in each cell
-        '''
+        '''return time grid and values in each cell'''
         t = np.concatenate((np.array([0]),
                             self.change_points,
                             np.array([np.inf])))
@@ -73,17 +58,103 @@ class History():
             yield (start_time, end_time, value)
 
     def check_grid(self, other: int):
+        '''test if time grid is the same as another instance'''
         if any(self.change_points != other.change_points):
             return False
         else:
             return True
 
+    def plot(self, idxs=None, **kwargs) -> None:
+        '''plot the history
+
[…]

### 2019-09-13 mushi.py
chi^2 are more legit than z-scores for assessing prf matrix fit

@@ -4,7 +4,7 @@
 from dataclasses import dataclass
 import numpy as np
 from scipy.special import binom
-from scipy.stats import poisson
+from scipy.stats import poisson, chi2
 from matplotlib import pyplot as plt
 from matplotlib.colors import SymLogNorm
 import prox_tv as ptv
@@ -309,7 +309,7 @@ class kSFS():
 
     def infer_μ(self, λ_tv: np.float64 = 0, α_tv: np.float64 = 0,
                 λ_r: np.float64 = 0, α_r: np.float64 = 0,
-                γ: np.float64 = 0.8, steps: int = 1000, tol: np.float64 = 1e-4,
+                γ: np.float64 = 0.8, max_iter: int = 1000, tol: np.float64 = 1e-4,
                 fit='prf', bins: np.ndarray = None) -> μ:
         '''return inferred μ history given the sfs and η history
 
@@ -318,7 +318,7 @@ class kSFS():
         λ_r: spectral (rank) regularization strength
         α_r: relative penalty on L1 vs L2 in spectral regularization
         γ: step size shrinkage rate for line search
-        steps: number of proximal gradient descent steps
+        max_iter: maximum number of proximal gradient descent iterations
         tol: relative tolerance in objective function
         fit: fit function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
@@ -417,7 +417,7 @@ class kSFS():
         f_old = f(Z)
         # initial step size
         s = 1
-        for k in range(1, steps + 1):
+        for k in range(1, max_iter + 1):
             # g(Q) and ∇g(Q)
             g1, grad_g1 = g(Q, grad=True)
             # Armijo line search
@@ -445,12 +445,12 @@ class kSFS():
             rel_change = np.abs((f_new - f_old) / f_old)
             if rel_change < tol:
                 print(f'relative change in loss function {rel_change:.2g} '
-                      f'is within tolerance {tol} after {k} steps')
+                      f'is within tolerance {tol} after {k} iterations')
                 break
             else:
                 f_old = f_new
-            if k == steps:
-                print(f'step size limit {steps} reached with relative '
+            if k == max_iter:
+                print(f'maximum iteration {max_iter} reached with relative '
[…]

### 2019-09-15 ancestral_states.py
1kg data scons pipeline

@@ -1,0 +1,111 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from Bio import SeqIO
+from Bio.Seq import Seq
+from Bio.SeqRecord import SeqRecord
+from Bio.Alphabet import generic_dna
+import re, sys
+import pandas as pd
+
+class AncestralStates():
+    def __init__(self, aln: SeqRecord, outgroup_aln: SeqRecord=None, df_seg: pd.DataFrame=None,
+                 verbose=False):
+        """
+        ancestral states of a chromosome
+        two constructors: load ancestral aln directly, or compute from reference, outgroup, and snps
+        @param aln: single record for a chr of ancestral sequence OR reference genome, e.g. hg19.1.fa if the other args are provided
+        @param outgroup_aln: outgroup (i.e. chimp) Bio.SeqRecord alignment for same chr
+                                    (single record, e.g. hg19ancNoChr.1.fa)
+        @param df_seg: pandas dataframe of snps indexed by (chr, pos) with columns [chr, pos, ref, alt, ac, an]
+        """
+        self.verbose = verbose
+        fields = aln.id.split(':')
+        self.chr = int(fields[0])
+        if len(fields) > 1:
+            self.start = int(fields[1].split('-')[0])
+        else:
+            self.start = 1
+        if outgroup_aln is None and df_seg is None:
+            # assert set(self.ancestral_alignment) == set('ACGT')
+            self.ancestral_alignment = aln.upper()
+        else:
+            assert aln.id == outgroup_aln.id
+            # initalize ascertainment to the reference, all uppercase
+            ascertained = aln.seq.upper().tomutable()
+            # iterate over seg sites and possibly change to chimp or fixed nonref
+            n_seg_sites = df_seg.shape[0]
+            for ct, (chr2, site) in enumerate(df_seg.index, 1):
+                if self.verbose:
+                    print('computing ancestral states for chr {} pos {}\r'.format(chr2, site), end='', flush=True)
+                assert chr2 == self.chr
+                ref_char = aln.seq[site - self.start].upper()
+                anc_char = outgroup_aln.seq[site - self.start].upper()
+                assert ref_char == df_seg.ref[(self.chr, site)]
+                assert anc_char in 'ACGTN'
+                # if the site is segregating, use the outgroup for ancestral state ascertainment
+                # fixed
+                if df_seg.ac[(self.chr, site)] == df_seg.an[(self.chr, site)]:
[…]

### 2019-09-15 qsub_scons.sh
1kg data scons pipeline

@@ -1,0 +1,29 @@
+#!usr/bin/bash
+
+cd /net/harris/vol1/project/europulse
+source activate europulse
+
+
+n_jobs=`nproc --all`
+ref_dir="/net/harris/vol1/data/phase3_1000genomes/bcfs"
+sample_pops="/net/harris/vol1/data/phase3_1000genomes/integrated_call_samples_v3.20130502.ALL.panel"
+# pop_flag="--pop=CEU"
+hg19="/net/harris/vol1/project/europulse/data/hg19NoChr.fa"
+hg19anc="/net/harris/vol1/project/europulse/data/hg19ancNoChr.fa"
+metadata="/net/gs/vol1/home/wsdewitt/kelley_home/project/europulse/data/kelley_bamInfo.xlsx"
+bam_dir="/net/harris/vol1/project/europulse/data/Martin_2018.06.04/"
+phastcons="/net/harris/vol1/project/europulse/data/phastConsElements100way.txt.gz"
+repeats="/net/harris/vol1/project/europulse/data/nestedRepeats.txt.gz"
+
+# outdir="/net/harris/vol1/project/europulse/scons_output"
+outdir="/net/harris/vol1/project/europulse/scons_output_debug"
+debug="--debug_mode"
+
+# no_exec="--no-exec"
+
+# set this if you want genotype likelihood fanciness
+gl="--gl"
+
+cmd="scons --metadata=${metadata} --bam_dir=${bam_dir} --ref=${ref_dir} ${pop_flag} --hg19=${hg19} --hg19_ancestral=${hg19anc} --phastcons=${phastcons} --repeats=${repeats} --jobs=${n_jobs} --outdir=${outdir} ${debug} ${no_exec} --sample_pops=${sample_pops} ${gl}"
+echo $cmd
+$cmd

### 2019-09-15 mushi.py
plot mutation type enrichment (like in Harris and Pritchard) as heatmap

@@ -6,7 +6,6 @@ import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson, chi2
 from matplotlib import pyplot as plt
-from matplotlib.colors import SymLogNorm
 import prox_tv as ptv
 
 
@@ -309,8 +308,9 @@ class kSFS():
 
     def infer_μ(self, λ_tv: np.float64 = 0, α_tv: np.float64 = 0,
                 λ_r: np.float64 = 0, α_r: np.float64 = 0,
-                γ: np.float64 = 0.8, max_iter: int = 1000, tol: np.float64 = 1e-4,
-                fit='prf', bins: np.ndarray = None) -> μ:
+                γ: np.float64 = 0.8, max_iter: int = 1000,
+                tol: np.float64 = 1e-4, fit='prf',
+                bins: np.ndarray = None) -> μ:
         '''return inferred μ history given the sfs and η history
 
         λ_tv: fused LASSO regularization strength
@@ -499,17 +499,16 @@ class kSFS():
         '''
         Y, X = np.meshgrid(range(1, self.n + 1), range(1, self.X.shape[1] + 2))
         if μ is None:
-            Z = self.X.T
-            cbar_label = 'number of variants'
-            c = plt.pcolormesh(X, Y, Z, norm=SymLogNorm(linthresh))
+            Z = self.X.T / self.X.sum(axis=1, keepdims=True).T
+            cbar_label = 'mutation type enrichment'
+            c = plt.pcolormesh(X, Y, Z)
             cbar = plt.colorbar(c)
             cbar.set_label(cbar_label, rotation=90)
         else:
             Ξ = self.L @ μ.Z
             Z = (self.X.T - Ξ.T) ** 2 / Ξ.T
             cbar_label = '$\\chi^2$'
-            c = plt.pcolormesh(X, Y, Z, vmin=0, cmap='Reds',
-                               norm=SymLogNorm(linthresh))
+            c = plt.pcolormesh(X, Y, Z, vmin=0, cmap='Reds')
             cbar = plt.colorbar(c)
             cbar.set_label(cbar_label, rotation=0)
             χ2_total = Z.sum()
@@ -517,7 +516,7 @@ class kSFS():
             print(f'χ\N{SUPERSCRIPT TWO} goodness of fit {χ2_total}, '
                   f'p = {p}')
         for line in range(1, self.X.shape[1] + 2):
-            plt.axvline(line, c='k', lw=0.5)
[…]

### 2019-09-16 mushi.py
hard sv thresholding

@@ -310,7 +310,8 @@ class kSFS():
                 λ_r: np.float64 = 0, α_r: np.float64 = 0,
                 γ: np.float64 = 0.8, max_iter: int = 1000,
                 tol: np.float64 = 1e-4, fit='prf',
-                bins: np.ndarray = None) -> μ:
+                bins: np.ndarray = None,
+                hard=False) -> μ:
         '''return inferred μ history given the sfs and η history
 
         λ_tv: fused LASSO regularization strength
@@ -322,6 +323,8 @@ class kSFS():
         tol: relative tolerance in objective function
         fit: fit function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
+        bins: SFS frequency bins
+        hard: hard Vs soft singular value thresholding
         '''
         assert λ_tv >= 0, 'λ_tv must be nonnegative'
         assert λ_r >= 0, 'λ_r must be nonnegative'
@@ -364,12 +367,21 @@ class kSFS():
                 '''
                 return ptv.tvgen(Z, [s * λ_tv * α_tv], [1], [1])
         elif λ_r * α_r > 0:
-            def prox_update(Z, s):
-                '''spectral regularization (nuclear norm penalty)
-                '''
-                U, σ, Vt = np.linalg.svd(Z, full_matrices=False)
-                Σ = np.diag(np.maximum(0, σ - s * λ_r * α_r))
-                return U @ Σ @ Vt
+            if hard:
+                def prox_update(Z, s):
+                    '''l0 norm on singular values (hard-thresholding)
+                    '''
+                    U, σ, Vt = np.linalg.svd(Z, full_matrices=False)
+                    σ[σ <= s * λ_r * α_r] = 0.0
+                    Σ = np.diag(σ)
+                    return U @ Σ @ Vt
+            else:
+                def prox_update(Z, s):
+                    '''l1 norm on singular values (soft-thresholding)
+                    '''
+                    U, σ, Vt = np.linalg.svd(Z, full_matrices=False)
+                    Σ = np.diag(np.maximum(0, σ - s * λ_r * α_r))
+                    return U @ Σ @ Vt
         else:
             def prox_update(Z, s):
                 return Z
@@ -380,9 +392,9 @@ class kSFS():
         # some matrices we'll need for the first difference penalties
[…]

### 2019-09-16 mushi.py
hard sv thresholding

@@ -6,7 +6,6 @@ import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson, chi2
 from matplotlib import pyplot as plt
-from matplotlib.colors import SymLogNorm
 import prox_tv as ptv
 
 
@@ -309,8 +308,10 @@ class kSFS():
 
     def infer_μ(self, λ_tv: np.float64 = 0, α_tv: np.float64 = 0,
                 λ_r: np.float64 = 0, α_r: np.float64 = 0,
-                γ: np.float64 = 0.8, max_iter: int = 1000, tol: np.float64 = 1e-4,
-                fit='prf', bins: np.ndarray = None) -> μ:
+                γ: np.float64 = 0.8, max_iter: int = 1000,
+                tol: np.float64 = 1e-4, fit='prf',
+                bins: np.ndarray = None,
+                hard=False) -> μ:
         '''return inferred μ history given the sfs and η history
 
         λ_tv: fused LASSO regularization strength
@@ -322,6 +323,8 @@ class kSFS():
         tol: relative tolerance in objective function
         fit: fit function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
+        bins: SFS frequency bins
+        hard: hard Vs soft singular value thresholding
         '''
         assert λ_tv >= 0, 'λ_tv must be nonnegative'
         assert λ_r >= 0, 'λ_r must be nonnegative'
@@ -364,12 +367,21 @@ class kSFS():
                 '''
                 return ptv.tvgen(Z, [s * λ_tv * α_tv], [1], [1])
         elif λ_r * α_r > 0:
-            def prox_update(Z, s):
-                '''spectral regularization (nuclear norm penalty)
-                '''
-                U, σ, Vt = np.linalg.svd(Z, full_matrices=False)
-                Σ = np.diag(np.maximum(0, σ - s * λ_r * α_r))
-                return U @ Σ @ Vt
+            if hard:
+                def prox_update(Z, s):
+                    '''l0 norm on singular values (hard-thresholding)
+                    '''
+                    U, σ, Vt = np.linalg.svd(Z, full_matrices=False)
+                    σ[σ <= s * λ_r * α_r] = 0.0
+                    Σ = np.diag(σ)
+                    return U @ Σ @ Vt
+            else:
+                def prox_update(Z, s):
[…]

### 2019-09-23 mushi.py
1kg 3-sfs and mushing

@@ -2,11 +2,16 @@
 # -*- coding: utf-8 -*-
 
 from dataclasses import dataclass
+from typing import List
+
 import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson, chi2
-from matplotlib import pyplot as plt
 import prox_tv as ptv
+
+from matplotlib import pyplot as plt
+import pandas as pd
+import seaborn as sns
 
 
 @dataclass
@@ -108,13 +113,17 @@ class η(History):
         plt.tight_layout()
 
 
+@dataclass
 class μ(History):
     '''mutation spectrum history
 
     change_points: epoch change points (times)
     Z: matrix of constant values for each epoch (rows) in each mutation type
        (columns)
+    mutation_types: list of mutation type names (default integer names)
     '''
+    mutation_types: List[str] = None
+
     @property
     def Z(self):
         '''read-only alias to vals attribute in base class'''
@@ -127,6 +136,11 @@ class μ(History):
     def __post_init__(self):
         super().__post_init__()
         assert len(self.Z.shape) == 2, self.Z.shape
+        if self.mutation_types is None:
+            self.mutation_types = range(1, self.Z.shape[1] + 1)
+        assert len(self.mutation_types) == self.Z.shape[1]
+        self.mutation_types = pd.Index(self.mutation_types,
+                                       name='mutation type')
 
     def plot(self, idxs=None, **kwargs) -> None:
         super().plot(idxs=idxs, **kwargs)
@@ -135,35 +149,43 @@ class μ(History):
         plt.xscale('symlog')
         plt.tight_layout()
 
-    def heatmap(self):
+    def clustermap(self, **kwargs):
+        '''clustermap of k-SFS
+
+        mutation_types: list of column names
+        kwargs: additional keyword arguments passed to pd.clustermap
+        '''
         t = np.concatenate((np.array([0]), self.change_points))
-        y, x = np.meshgrid(t, range(1, self.Z.shape[1] + 2))
-        c = plt.pcolormesh(x, y, self.Z.T)
-        for line in range(1, self.Z.shape[1] + 2):
-            plt.axvline(line, c='k', lw=0.5)
-        plt.gca().invert_yaxis()
-        plt.xlabel('mutation type')
-        plt.yscale('symlog')
-        plt.ylabel('$t$', rotation=0)
-        cbar = plt.colorbar(c)
-        cbar.set_label('$\\mu(t)$', rotation=0)
-        plt.tight_layout()
[…]

### 2019-09-23 qsub_scons.sh

@@ -1,29 +1,25 @@
 #!usr/bin/bash
 
-cd /net/harris/vol1/project/europulse
-source activate europulse
+# cd /net/harris/vol1/project/mushi
+# source activate mushi
 
+# linux
+# n_jobs=`nproc --all`
+# mac
+n_jobs=`sysctl -n hw.physicalcpu`
 
-n_jobs=`nproc --all`
-ref_dir="/net/harris/vol1/data/phase3_1000genomes/bcfs"
-sample_pops="/net/harris/vol1/data/phase3_1000genomes/integrated_call_samples_v3.20130502.ALL.panel"
-# pop_flag="--pop=CEU"
-hg19="/net/harris/vol1/project/europulse/data/hg19NoChr.fa"
-hg19anc="/net/harris/vol1/project/europulse/data/hg19ancNoChr.fa"
-metadata="/net/gs/vol1/home/wsdewitt/kelley_home/project/europulse/data/kelley_bamInfo.xlsx"
-bam_dir="/net/harris/vol1/project/europulse/data/Martin_2018.06.04/"
-phastcons="/net/harris/vol1/project/europulse/data/phastConsElements100way.txt.gz"
-repeats="/net/harris/vol1/project/europulse/data/nestedRepeats.txt.gz"
+bcfs="data.nosync/bcf_files"
+ancs="data.nosync/human_ancestor_GRCh37_e59"
+sample_pops="data.nosync/integrated_call_samples_v3.20130502.ALL.panel"
+pop_flag="--pop=CEU"
+k="3"
+phastcons="data.nosync/phastConsElements100way.txt.gz"
+repeats="data.nosync/nestedRepeats.txt.gz"
 
-# outdir="/net/harris/vol1/project/europulse/scons_output"
-outdir="/net/harris/vol1/project/europulse/scons_output_debug"
-debug="--debug_mode"
+outdir="scons_output"
 
-# no_exec="--no-exec"
+no_exec="--no-exec"
 
-# set this if you want genotype likelihood fanciness
-gl="--gl"
-
[…]

### 2019-09-23 ancestor.py
1kg 3-sfs and mushing

@@ -1,0 +1,207 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from Bio import SeqIO
+from Bio.Seq import reverse_complement
+from typing import Tuple
+import numpy as np
+import re
+import sys
+import pandas as pd
+from pandas.api.types import CategoricalDtype
+
+
+class Ancestor():
+    nts = ('A', 'C', 'G', 'T')
+
+    def __init__(self, anc_aln_file: str):
+        '''ancestral state of a chromosome
+
+        anc_aln_file: path to ancestral sequence fasta
+        '''
+
+        # parse fasta to Bio.SeqRecord
+        self.anc_aln = SeqIO.read(anc_aln_file, 'fasta')
+        self.chr = self.anc_aln.id.split(':')[2]
+        if int(self.chr) not in range(1, 23):
+            raise ValueError(f'chromosome {self.chr} not in 1-22')
+        self.anc_aln_len = len(self.anc_aln)
+
+    def get_state(self, pos: int) -> str:
+        '''return ancestral state at site pos (1-based)'''
+        if pos > self.anc_aln_len:
+            print(f'position {pos} is beyond chromosome length '
+                  f'{self.anc_aln_len}')
+            return None
+        return self.anc_aln.seq[pos - 1]
+
+    def write(self, file=sys.stdout):
+        '''print fasta record to file (default stdout)'''
+        SeqIO.write(self.anc_aln, file, 'fasta')
+
+    def motif_sites(self, motif: str = 'TCC', target: int = None) -> pd.Index:
+        '''return pandas frame of all sites that have a given motif context
+
+        motif: sequence, can be regex, (default "TCC")
+        target: which position of the site within the motif (default middle)
+        '''
+        if target is None:
+            assert len(motif) % 2 == 1
+            target = len(motif) // 2
+        else:
+            target = target
+        if motif[target] not in 'ACGT':
+            raise ValueError(f'motif {motif} at target position {target} must'
+                             ' be a nucleotide')
+        # add the reverse complement of the motif to make a regular expression
+        motif_complement = reverse_complement(motif)
+        motif_regex = re.compile(motif + '|' + motif_complement)
+        sites = []
+        for match in motif_regex.finditer(str(self.anc_aln.seq)):
+            # if it's revcomp match, adjust target accordingly
+            if match.group(0) == motif:
+                this_target = target
+            elif match.group(0) == motif_complement:
+                this_target = len(motif) - target - 1
+            else:
[…]

### 2019-09-23 ancestral_states.py

@@ -1,111 +1,0 @@
-#! /usr/bin/env python
-# -*- coding: utf-8 -*-
-
-from Bio import SeqIO
-from Bio.Seq import Seq
-from Bio.SeqRecord import SeqRecord
-from Bio.Alphabet import generic_dna
-import re, sys
-import pandas as pd
-
-class AncestralStates():
-    def __init__(self, aln: SeqRecord, outgroup_aln: SeqRecord=None, df_seg: pd.DataFrame=None,
-                 verbose=False):
-        """
-        ancestral states of a chromosome
-        two constructors: load ancestral aln directly, or compute from reference, outgroup, and snps
-        @param aln: single record for a chr of ancestral sequence OR reference genome, e.g. hg19.1.fa if the other args are provided
-        @param outgroup_aln: outgroup (i.e. chimp) Bio.SeqRecord alignment for same chr
-                                    (single record, e.g. hg19ancNoChr.1.fa)
-        @param df_seg: pandas dataframe of snps indexed by (chr, pos) with columns [chr, pos, ref, alt, ac, an]
-        """
-        self.verbose = verbose
-        fields = aln.id.split(':')
-        self.chr = int(fields[0])
-        if len(fields) > 1:
-            self.start = int(fields[1].split('-')[0])
-        else:
-            self.start = 1
-        if outgroup_aln is None and df_seg is None:
-            # assert set(self.ancestral_alignment) == set('ACGT')
-            self.ancestral_alignment = aln.upper()
-        else:
-            assert aln.id == outgroup_aln.id
-            # initalize ascertainment to the reference, all uppercase
-            ascertained = aln.seq.upper().tomutable()
-            # iterate over seg sites and possibly change to chimp or fixed nonref
-            n_seg_sites = df_seg.shape[0]
-            for ct, (chr2, site) in enumerate(df_seg.index, 1):
-                if self.verbose:
-                    print('computing ancestral states for chr {} pos {}\r'.format(chr2, site), end='', flush=True)
-                assert chr2 == self.chr
-                ref_char = aln.seq[site - self.start].upper()
-                anc_char = outgroup_aln.seq[site - self.start].upper()
-                assert ref_char == df_seg.ref[(self.chr, site)]
-                assert anc_char in 'ACGTN'
-                # if the site is segregating, use the outgroup for ancestral state ascertainment
-                # fixed
-                if df_seg.ac[(self.chr, site)] == df_seg.an[(self.chr, site)]:
[…]

### 2019-09-23 1KG/ancestor.py
tidying

@@ -118,7 +118,7 @@ class Ancestor():
         # categorical dtype for nucleotides
         nuc_dtype = CategoricalDtype(list('ACGTN'))
         snps = pd.read_csv(snps_file, sep='\t', header=None,
-                           names=('chr', 'pos', 'ref', 'alt', 'aa', 'ac', 'an'),
+                           names=('chr', 'pos', 'ref', 'alt', 'ac', 'an'),
                            dtype={'chr': np.uint8, 'pos': np.uint32,
                                   'ref': nuc_dtype, 'alt': nuc_dtype,
                                   'ac': np.uint16, 'an': np.uint16},

### 2019-09-23 1KG/qsub_scons.sh

@@ -8,13 +8,13 @@
 # mac
 n_jobs=`sysctl -n hw.physicalcpu`
 
-bcfs="data.nosync/bcf_files"
-ancs="data.nosync/human_ancestor_GRCh37_e59"
-sample_pops="data.nosync/integrated_call_samples_v3.20130502.ALL.panel"
+bcfs="../data.nosync/bcf_files"
+ancs="../data.nosync/human_ancestor_GRCh37_e59"
+sample_pops="../data.nosync/integrated_call_samples_v3.20130502.ALL.panel"
 pop_flag="--pop=CEU"
 k="3"
-phastcons="data.nosync/phastConsElements100way.txt.gz"
-repeats="data.nosync/nestedRepeats.txt.gz"
+phastcons="../data.nosync/phastConsElements100way.txt.gz"
+repeats="../data.nosync/nestedRepeats.txt.gz"
 
 outdir="scons_output"
 

### 2019-09-23 mushi.py

@@ -162,7 +162,7 @@ class μ(History):
                           columns=self.mutation_types)
         g = sns.clustermap(df, row_cluster=False, center=1,
                            metric='correlation',
-                           cbar_kws={'label': '$relative mutation intensity$'},
+                           cbar_kws={'label': 'relative mutation intensity'},
                            **kwargs)
         g.ax_heatmap.set_yscale('symlog')
         return g

### 2019-09-23 histories.py
refactoring a few classes/fcts to their own modules to tidy `mushi.py`

@@ -1,0 +1,165 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from dataclasses import dataclass
+from typing import List
+
+import numpy as np
+
+from matplotlib import pyplot as plt
+import pandas as pd
+import seaborn as sns
+
+
+@dataclass
+class History():
+    '''base class piecewise constant history. The first epoch starts at zero,
+    and the last epoch extends to infinity
+
+    change_points: epoch change points (times)
+    vals: constant values for each epoch (rows)
+    '''
+    change_points: np.array
+    vals: np.ndarray
+
+    def __post_init__(self):
+        if any(np.diff(self.change_points) <= 0) or any(
+           np.isinf(self.change_points)) or any(self.change_points <= 0):
+            raise ValueError('change_points must be increasing, finite, and '
+                             'positive')
+        if len(self.vals) != len(self.change_points) + 1:
+            raise ValueError(f'len(change_points) = {len(self.change_points)}'
+                             f' implies {len(self.change_points) + 1} epochs,'
+                             f' but len(vals) = {len(self.vals)}')
+        if np.any(self.vals <= 0) or np.sum(np.isinf(self.vals)):
+            raise ValueError(f'elements of vals must be finite and '
+                             'positive')
+        self.m = len(self.vals)
+
+    def arrays(self):
+        '''return time grid and values in each cell'''
+        t = np.concatenate((np.array([0]),
+                            self.change_points,
+                            np.array([np.inf])))
+        return t, self.vals
+
+    def epochs(self):
+        '''generator yielding epochs as tuples: (start_time, end_time, value)
+        '''
+        for i in range(self.m):
+            if i == 0:
+                start_time = 0
+            else:
+                start_time = self.change_points[i - 1]
+            if i == self.m - 1:
+                end_time = np.inf
+            else:
+                end_time = self.change_points[i]
+            value = self.vals[i]
+            yield (start_time, end_time, value)
+
+    def check_grid(self, other: int):
+        '''test if time grid is the same as another instance'''
+        if any(self.change_points != other.change_points):
+            return False
+        else:
+            return True
+
+    def plot(self, idxs=None, **kwargs) -> None:
+        '''plot the history
+
+        idxs: indices of value column(s) to plot (optional)
[…]

### 2019-09-23 utils.py

@@ -1,0 +1,112 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import numpy as np
+from scipy.special import binom
+
+import histories
+
+from matplotlib import pyplot as plt
+import pandas as pd
+import seaborn as sns
+
+def C(n: int) -> np.ndarray:
+    '''The C matrix defined in the text
+
+    n: number of sampled haplotypes
+    '''
+    W1 = np.zeros((n - 1, n - 1))
+    W2 = np.zeros((n - 1, n - 1))
+    b = np.arange(1, n - 1 + 1)
+    # j = 2
+    W1[:, 0] = 6 / (n + 1)
+    W2[:, 0] = 0
+    # j = 3
+    W1[:, 1] = 10 * (5 * n - 6 * b - 4) / (n + 2) / (n + 1)
+    W2[:, 1] = (20 * (n - 2)) / (n+1) / (n+2)
+    for j in range(2, n - 1):
+        col = j - 2
+        # procedurally generated by Zeilberger's algorithm in Mathematica
+        W1[:, col + 2] = -((-((-1 + j)*(1 + j)**2*(3 + 2*j)*(j - n)*(4 + 2*j - 2*b*j + j**2 - b*j**2 + 4*n + 2*j*n + j**2*n)*W1[:, col]) - (-1 + 2*j)*(3 + 2*j)*(-4*j - 12*b*j - 4*b**2*j - 6*j**2 - 12*b*j**2 - 2*b**2*j**2 - 4*j**3 + 4*b**2*j**3 - 2*j**4 + 2*b**2*j**4 + 4*n + 2*j*n - 6*b*j*n + j**2*n - 9*b*j**2*n - 2*j**3*n - 6*b*j**3*n - j**4*n - 3*b*j**4*n + 4*n**2 + 6*j*n**2 + 7*j**2*n**2 + 2*j**3*n**2 + j**4*n**2)*W1[:, col + 1])/(j**2*(2 + j)*(-1 + 2*j)*(1 + j + n)*(3 + b + j**2 - b*j**2 + 3*n + j**2*n)))
[…]

### 2019-09-23 mushi.py
refactoring a few classes/fcts to their own modules to tidy `mushi.py`

@@ -1,177 +1,24 @@
 #! /usr/bin/env python
 # -*- coding: utf-8 -*-
 
-from dataclasses import dataclass
 from typing import List
-
 import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson, chi2
 import prox_tv as ptv
+
+import histories
+import utils
 
 from matplotlib import pyplot as plt
 import pandas as pd
 import seaborn as sns
 
 
-@dataclass
-class History():
-    '''base class piecewise constant history. The first epoch starts at zero,
-    and the last epoch extends to infinity
-
-    change_points: epoch change points (times)
-    vals: constant values for each epoch (rows)
-    '''
-    change_points: np.array
-    vals: np.ndarray
-
-    def __post_init__(self):
-        if any(np.diff(self.change_points) <= 0) or any(
-           np.isinf(self.change_points)) or any(self.change_points <= 0):
-            raise ValueError('change_points must be increasing, finite, and '
-                             'positive')
-        if len(self.vals) != len(self.change_points) + 1:
-            raise ValueError(f'len(change_points) = {len(self.change_points)}'
-                             f' implies {len(self.change_points) + 1} epochs,'
-                             f' but len(vals) = {len(self.vals)}')
-        if np.any(self.vals <= 0) or np.sum(np.isinf(self.vals)):
-            raise ValueError(f'elements of vals must be finite and '
-                             'positive')
-        self.m = len(self.vals)
-
-    def arrays(self):
-        '''return time grid and values in each cell'''
-        t = np.concatenate((np.array([0]),
-                            self.change_points,
-                            np.array([np.inf])))
-        return t, self.vals
-
-    def epochs(self):
-        '''generator yielding epochs as tuples: (start_time, end_time, value)
-        '''
-        for i in range(self.m):
-            if i == 0:
-                start_time = 0
-            else:
-                start_time = self.change_points[i - 1]
-            if i == self.m - 1:
-                end_time = np.inf
-            else:
-                end_time = self.change_points[i]
-            value = self.vals[i]
[…]

### 2019-09-23 mushi.py
changed "misfit" to "loss" for better readability

@@ -321,7 +321,7 @@ class kSFS():
         γ: step size shrinkage rate for line search
         max_iter: maximum number of proximal gradient descent iterations
         tol: relative tolerance in objective function
-        fit: fit function, 'prf' for Poisson random field, 'kl' for
+        fit: loss function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
         bins: SFS frequency bins
         hard: hard Vs soft singular value thresholding
@@ -349,12 +349,12 @@ class kSFS():
             self.L = L_binned
         # badness of fit
         if fit == 'prf':
-            def misfit_func(*args, **kwargs):
+            def loss_func(*args, **kwargs):
                 return -self.ℓ(*args, **kwargs)
         elif fit == 'kl':
-            misfit_func = self.d_kl
+            loss_func = self.d_kl
         elif fit == 'lsq':
-            misfit_func = self.lsq
+            loss_func = self.lsq
         else:
             raise ValueError(f'unrecognized fit argument {fit}')
         if λ_tv * α_tv > 0 and λ_r * α_r > 0:
@@ -393,20 +393,20 @@ class kSFS():
         D = (np.eye(self.η.m, k=0) - np.eye(self.η.m, k=-1))
         W = np.eye(self.η.m)
         W[0, 0] = 0  # W matrix deals with boundary condition
-        D1 = W @ D
-        D2 = D1.T @ D1
+        D1 = W @ D  # 1st difference matrix
+        D2 = D1.T @ D1 # square of 1st difference matrix (Laplacian)
 
         def g(Z, grad=False):
             '''differentiable piece of loss'''
             if grad:
-                misfit, grad_misfit = misfit_func(Z, grad=True)
+                loss, grad_loss = loss_func(Z, grad=True)
             else:
-                misfit = misfit_func(Z)
-            g = misfit \
+                loss = loss_func(Z)
+            g = loss \
                 + (λ_tv / 2) * (1 - α_tv) * ((D1 @ Z) ** 2).sum() \
                 + (λ_r / 2) * (1 - α_r) * (Z ** 2).sum()
             if grad:
-                grad_g = grad_misfit + λ_tv * (1 - α_tv) * D2 @ Z \
+                grad_g = grad_loss + λ_tv * (1 - α_tv) * D2 @ Z \
                                      + λ_r * (1 - α_r) * Z
                 return g, grad_g
             return g

### 2019-09-23 manuscript/main.tex
added various losses to manuscript

@@
 
+Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
+\begin{enumerate}
+\item the traditional sample frequency \emph{spectrum}
+\item mutation \emph{spectrum} as in the nucleotide context of each mutation
+\item \emph{spectral} regularization for encouraging low rank solutions
+\end{enumerate}
+
 \section*{Model}\label{sec:model}
@@
 %WD more inverse problem intro
-The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
+The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, 
+so many very different histories can be equally consistent with the data 
+\citep{oscillation paper? Yun's other papers?}.
+We deal with this problem using regularization, to enforce well-behaved histories.
+Our cost function to minimize, a penalized log-likelihood, is
+\begin{equation}
+\label{eqn:penalized}
+C(Z)
+%\tilde\ell(\boldsymbol z)
+= -\ell(Z) + R(Z) .
+\end{equation}
+The function $R(Z)$ incorporates our regularization.
 In order to recover smooth histories,
@@
 where $\Delta$ denotes the first difference matrix.
-Our cost function (penalized log-likelihood) is then
+We have implemented three different types of regularization.
+
+The first option we call ``soft'' rank regularization, with 
 \begin{equation}
-\label{eqn:penalized}
-C(Z)
-%\tilde\ell(\boldsymbol z)
-= -\ell(Z) + \lambda \left(\alpha\left\|\Delta Z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta Z\right\|_2^2\right),
+  \label{eq:regularization_soft}
+  R_\mathrm{soft}(Z) 
+  = 
+  \lambda_\mathrm{spline} 
+  \frac{1}{2} \left\|\Delta Z\right\|_2^2
+  +
+  \lambda_\mathrm{rank} \| Z \|_*.
 \end{equation}
-where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
-\textcolor{blue}{Kam note: probably we need a ``Ridge'' (Tikhonov)
-term like the following to deal with small singular values:}
-\[
-\lambda' \| Z \|_2^2
-\]
+The spline term controls the smoothness of solutions, 
+and the rank term allows us to recover low-rank solutions.
+The rank penalty is the nuclear norm $\| Z \|_*$, 
+which is equivalent to the $\ell^1$-norm on the singular values.
+If $Z = U S V^\intercal$ is the SVD of $Z$, with $S = \mathrm{diag}(\sigma)$ then $\| Z \|_* = \| \sigma \|_1$.
+The parameters $\lambda_\mathrm{spline}$ and $\lambda_\mathrm{rank}$ control the overall strength
+of smoothing and rank regularization.
+The function $R_\mathrm{soft}$ is convex; therefore we have a convex optimization problem 
+in this scenario.
+
+The second option is ``hard'' rank regularization, where
+\begin{equation}
+  \label{eq:regularization_soft}
+  R_\mathrm{hard}(Z) 
+  = 
+  \lambda_\mathrm{spline} 
+  \frac{1}{2} \left\|\Delta Z\right\|_2^2
+  +
+  \lambda_\mathrm{rank} \, \mathrm{rank}(Z).
+\end{equation}
+Here, the rank penalty is the actual rank of $Z$,
+and note that $\mathrm{rank}(Z) = \| \sigma \|_0$.
+In this case the problem is {\em not} convex.
+However, we may still compute a proximal map for the non-differentiable rank term
+and optimize with prox-gradients.
+
+The third option allows more general smoothing but ignores all rank penalization:
+\begin{equation}
+  \label{eq:regularization_soft}
+  R_\mathrm{indep}(Z) 
+  = 
+  \lambda_\mathrm{TV} 
+  \left(
+    \alpha_\mathrm{TV}\left\|\Delta Z\right\|_1 + \frac{(1-\alpha_\mathrm{TV})}{2}\left\|\Delta Z\right\|_2^2
+  \right) .
+\end{equation}
+The TV terms control the smoothness of solutions, 
+and the rank term allows us to recover low-rank solutions.
+The function $R_\mathrm{indep}$ is convex; 
+therefore we have a convex optimization problem.
+Note also that the columns of $Z$ are uncoupled, 
+so that we may solve this case as $\kappa_k$ different subproblems.
 

### 2019-09-23 manuscript/main.tex

--- 2019-09-10 manuscript/main.tex
@@
 
-Single nucleotide variant (SNV) frequencies partitioned according to triplet nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
+Single nucleotide variant (SNV) frequencies partitioned according to local nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
 The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
@@
 \cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
-We develop coalescent theory for the dependence of the expected 3-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} triplet mutation spectrum history (triplet-specific mutation intensities as a function of time into the past).
+We develop coalescent theory for the dependence of the expected $k$-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} mutation spectrum history (mutation intensities as a function of time into the past for each mutation type).
 This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized.
@@
 %WD todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
+
+Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
+\begin{enumerate}
+\item the traditional sample frequency \emph{spectrum}
+\item mutation \emph{spectrum} as in the nucleotide context of each mutation
+\item \emph{spectral} regularization for encouraging low rank solutions
+\end{enumerate}
 
@@
   \label{eqn:transform}
-\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu(t),
+\boldsymbol \xi = \mathcal{L}_{n,\eta(t)}\mu(t),
 \end{equation}
-where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
+where $\mathcal{L}_{n,\eta(t)}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
 It is dependent on $n$ and $\eta$, which we take to be fixed.
@@
   \label{eqn:transform_matrix}
-\Xi = \mathcal{L}_{n,\eta}\boldsymbol\mu^\intercal(t),
+\Xi = \mathcal{L}_{n,\eta(t)}\boldsymbol\mu^\intercal(t),
 \end{equation}
@@
 \[
-\ell(Z; X, \boldsymbol y, n) = \Tr(X^\intercal\log(L_{n, \boldsymbol y} Z)) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} Z\boldsymbol 1,
+\ell(Z; X, \boldsymbol y, n) = \Tr(X^\intercal\log(L_{n, \boldsymbol y} Z)) - \|L_{n, \boldsymbol y} Z\|_1,
 \]

### 2019-09-23 manuscript/main.tex
defined TV & spline

--- 2019-09-23 manuscript/main.tex
@@
 where $\Delta$ denotes the first difference matrix.
+When $p = 1$, this is referred to as a fused LASSO or total variation (TV) penalty,
+whereas $p=2$ is called a spline penalty.
 We have implemented three different types of regularization.

### 2019-09-23 mushi.py
small edits & loss -> cost

@@ -184,7 +184,7 @@ class kSFS():
         D2 = D1.T @ D1 # square of 1st difference matrix (Laplacian)
 
         def g(logZ, grad=False):
-            '''differentiable piece of loss'''
+            '''differentiable piece of cost'''
             Z = np.exp(logZ)
             if grad:
                 loss, grad_loss = loss_func(Z, grad=True)
@@ -200,18 +200,19 @@ class kSFS():
             return g
 
         def h(logZ):
-            '''nondifferentiable piece of loss'''
+            '''nondifferentiable piece of cost'''
             Z = np.exp(logZ)
             σ = np.linalg.svd(Z, compute_uv=False)
             if hard:
                 rank_penalty = np.linalg.norm(σ, 0)
             else:
                 rank_penalty = np.linalg.norm(σ, 1)
-            return λ_tv * α_tv * np.abs(D1 @ Z).sum() \
+
+                return λ_tv * α_tv * np.abs(D1 @ Z).sum() \
                 + λ_r * α_r * rank_penalty
 
         def f(logZ):
-            '''loss'''
+            '''cost'''
             return g(logZ) + h(logZ)
 
         # initialize using constant μ history MLE

### 2019-09-23 mushi.py
added comments

@@ -172,7 +172,7 @@ class kSFS():
             def prox_update(Z, s):
                 return Z
 
-        # Accelerated proximal gradient descent: our loss function decomposes
+        # Accelerated proximal gradient descent: our cost function decomposes
         # as f = g + h, where g is differentiable and h is not.
         # We transform logZ to restrict to positive solutions
         # https://people.eecs.berkeley.edu/~elghaoui/Teaching/EE227A/lecture18.pdf
@@ -196,7 +196,8 @@ class kSFS():
             if grad:
                 grad_g = grad_loss + λ_tv * (1 - α_tv) * D2 @ Z \
                                      + λ_r * (1 - α_r) * Z
-                return g, grad_g * Z
+                grad_g_log = grad_g * Z # change of variables
+                return g, grad_g_log 
             return g
 
         def h(logZ):
@@ -207,8 +208,8 @@ class kSFS():
                 rank_penalty = np.linalg.norm(σ, 0)
             else:
                 rank_penalty = np.linalg.norm(σ, 1)
-
-                return λ_tv * α_tv * np.abs(D1 @ Z).sum() \
+            # now add in TV term... you will never have both
+            return λ_tv * α_tv * np.abs(D1 @ Z).sum() \
                 + λ_r * α_r * rank_penalty
 
         def f(logZ):
@@ -217,18 +218,17 @@ class kSFS():
 
         # initialize using constant μ history MLE
         μ = self.constant_μ_MLE()
-        logZ = np.log(μ.Z)
-        # our auxiliary iterates for acceleration
-        logQ = np.log(μ.Z)
+        logZ = np.log(μ.Z) # current iterate
+        logQ = np.log(μ.Z) # momentum iterate
         # initial loss
         f_trajectory = [f(logZ)]
         # initialize step size
-        s0 = 1
-        s = s0
+        s0 = 1 # max step size
+        s = s0 # current step size
         # max number of Armijo step size reductions
         max_line_iter = 100
         for k in range(1, max_iter + 1):
-            # g(logQ) and ∇g(logQ)
+            # evaluate smooth part of loss at momentum point
             g1, grad_g1 = g(logQ, grad=True)
             # Armijo line search
             for line_iter in range(max_line_iter):
@@ -238,9 +238,12 @@ class kSFS():
                 G = (1 / s) * (logQ - prox_update(logQ - s * grad_g1, s))
                 # test g(logQ - sG_s(logQ)) for line search
                 if g(logQ - s * G) <= (g1 - s * (grad_g1 * G).sum()
[…]

### 2019-09-24 mushi.py
fixes to nesterov

@@ -230,13 +230,17 @@ class kSFS():
         for k in range(1, max_iter + 1):
             # evaluate smooth part of loss at momentum point
             g1, grad_g1 = g(logQ, grad=True)
+            # store old iterate
+            logZ_old = logZ
             # Armijo line search
             for line_iter in range(max_line_iter):
                 if not np.all(np.isfinite(grad_g1)):
                     raise RuntimeError(f'invalid gradient: {grad_g1}')
+                # new point via prox-gradient of momentum point
+                logZ = prox_update(logQ - s * grad_g1, s)
                 # G_s(Q) as in the notes linked above
-                G = (1 / s) * (logQ - prox_update(logQ - s * grad_g1, s))
-                # test g(logQ - sG_s(logQ)) for line search
+                G = (1 / s) * (logQ - logZ)
+                # test g(logQ - sG_s(logQ)) for sufficient decrease
                 if g(logQ - s * G) <= (g1 - s * (grad_g1 * G).sum()
                                     + (s / 2) * (G ** 2).sum()):
                     # Armijo satisfied
@@ -244,10 +248,8 @@ class kSFS():
                 else:
                     # Armijo not satisfied
                     s *= γ # shrink step size
-            # accelerated gradient step
-            logZ_old = logZ
-            logZ = logQ - s * G
-            logQ = logZ + (k / (k + 3)) * (logZ - logZ_old)
+            # update momentum term
+            logQ = logZ + ((k - 1) / (k + 2)) * (logZ - logZ_old)
             if line_iter == max_line_iter - 1:
                 print('warning: line search failed')
                 s = s0

### 2019-09-24 histories.py
handle mu vector instead of matrix

@@ -132,6 +132,9 @@ class μ(History):
 
     def __post_init__(self):
         super().__post_init__()
+        # if mutation rate vector instead of matrix, promote to matrix
+        if len(self.Z.shape) == 1:
+            self.Z = self.Z[:, np.newaxis]
         assert len(self.Z.shape) == 2, self.Z.shape
         if self.mutation_types is None:
             self.mutation_types = range(1, self.Z.shape[1] + 1)

### 2019-09-24 mushi.py
minor edits for PEP 8 compliance

@@ -187,7 +187,7 @@ class kSFS():
         W = np.eye(self.η.m)
         W[0, 0] = 0  # W matrix deals with boundary condition
         D1 = W @ D  # 1st difference matrix
-        D2 = D1.T @ D1 # square of 1st difference matrix (Laplacian)
+        D2 = D1.T @ D1  # square of 1st difference matrix (Laplacian)
 
         def g(logZ, grad=False):
             '''differentiable piece of cost'''
@@ -202,8 +202,8 @@ class kSFS():
             if grad:
                 grad_g = grad_loss + λ_tv * (1 - α_tv) * D2 @ Z \
                                      + λ_r * (1 - α_r) * Z
-                grad_g_log = grad_g * Z # change of variables
-                return g, grad_g_log 
+                grad_g_log = grad_g * Z  # change of variables
+                return g, grad_g_log
             return g
 
         def h(logZ):
@@ -224,13 +224,13 @@ class kSFS():
 
         # initialize using constant μ history MLE
         μ = self.constant_μ_MLE()
-        logZ = np.log(μ.Z) # current iterate
-        logQ = np.log(μ.Z) # momentum iterate
+        logZ = np.log(μ.Z)  # current iterate
+        logQ = np.log(μ.Z)  # momentum iterate
         # initial loss
         f_trajectory = [f(logZ)]
         # initialize step size
-        s0 = 1 # max step size
-        s = s0 # current step size
+        s0 = 1  # max step size
+        s = s0  # current step size
         # max number of Armijo step size reductions
         max_line_iter = 100
         for k in range(1, max_iter + 1):
@@ -248,12 +248,12 @@ class kSFS():
                 G = (1 / s) * (logQ - logZ)
                 # test g(logQ - sG_s(logQ)) for sufficient decrease
                 if g(logQ - s * G) <= (g1 - s * (grad_g1 * G).sum()
-                                    + (s / 2) * (G ** 2).sum()):
+                                       + (s / 2) * (G ** 2).sum()):
                     # Armijo satisfied
                     break
                 else:
                     # Armijo not satisfied
-                    s *= γ # shrink step size
+                    s *= γ  # shrink step size
             # update momentum term
             logQ = logZ + ((k - 1) / (k + 2)) * (logZ - logZ_old)
             if line_iter == max_line_iter - 1:

### 2019-09-25 mushi.py
`\ell->prf`

@@ -140,7 +140,7 @@ class kSFS():
             X = self.X
         if fit == 'prf':
             def loss_func(Z, **kwargs):
-                return -utils.ℓ(Z, X, L, **kwargs)
+                return -utils.prf(Z, X, L, **kwargs)
         elif fit == 'kl':
             def loss_func(Z, **kwargs):
                 return utils.d_kl(Z, X, L, **kwargs)

### 2019-09-25 utils.py
`\ell->prf`

@@ -51,7 +51,7 @@ def M(n: int, η: histories.η) -> np.ndarray:
         @ np.diag(y)
 
 
-def ℓ(Z: np.ndarray, X: np.ndarray, L: np.ndarray,
+def prf(Z: np.ndarray, X: np.ndarray, L: np.ndarray,
       grad: bool = False) -> np.float:
     '''Poisson random field log-likelihood of history
 

### 2019-09-25 mushi.py
better gradient error message

@@ -241,7 +241,8 @@ class kSFS():
             # Armijo line search
             for line_iter in range(max_line_iter):
                 if not np.all(np.isfinite(grad_g1)):
-                    raise RuntimeError(f'invalid gradient: {grad_g1}')
+                    raise RuntimeError(f'invalid gradient at step {k}, line '
+                                       f'search step {line_iter}: {grad_g1}')
                 # new point via prox-gradient of momentum point
                 logZ = prox_update(logQ - s * grad_g1, s)
                 # G_s(Q) as in the notes linked above

### 2019-09-26 histories.py
plotting with optional `normed` flag

@@ -5,7 +5,7 @@ from dataclasses import dataclass
 from typing import List
 
 import numpy as np
-
+import matplotlib as mpl
 from matplotlib import pyplot as plt
 import pandas as pd
 import seaborn as sns
@@ -65,7 +65,7 @@ class History():
         else:
             return True
 
-    def plot(self, idxs=None, **kwargs) -> None:
+    def plot(self, idxs=None, **kwargs) -> List[mpl.lines.Line2D]:
         '''plot the history
 
         idxs: indices of value column(s) to plot (optional)
@@ -76,10 +76,13 @@ class History():
             vals = self.vals[:, idxs]
         else:
             vals = self.vals
-        plt.step(t, vals, where='post', **kwargs)
+        lines = plt.step(t, vals, where='post', **kwargs)
         plt.xlabel('$t$')
+        plt.xscale('symlog')
         if 'label' in kwargs:
             plt.legend()
+        plt.tight_layout()
+        return lines
 
 
 class η(History):
@@ -101,13 +104,12 @@ class η(History):
         super().__post_init__()
         assert len(self.y.shape) == 1, self.y.shape
 
-    def plot(self, **kwargs) -> None:
-        super().plot(**kwargs)
-        plt.xlabel('$t$')
+    def plot(self, **kwargs) -> List[mpl.lines.Line2D]:
+        lines = super().plot(**kwargs)
         plt.ylabel('$η(t)$')
-        plt.xscale('symlog')
         plt.yscale('log')
         plt.tight_layout()
+        return lines
 
 
 @dataclass
@@ -142,12 +144,29 @@ class μ(History):
         self.mutation_types = pd.Index(self.mutation_types,
                                        name='mutation type')
 
-    def plot(self, idxs=None, **kwargs) -> None:
-        super().plot(idxs=idxs, **kwargs)
-        plt.xlabel('$t$')
-        plt.ylabel('$\\mu(t)$')
-        plt.xscale('symlog')
+    def plot(self, idxs=None, normed=False,
+             **kwargs) -> List[mpl.lines.Line2D]:
+        '''
+        normed: flag to normalize to relative mutation intensity
+        '''
+        lines = super().plot(idxs=idxs, **kwargs)
+        if normed:
+            Z = self.Z / self.Z.sum(1, keepdims=True)
+            if idxs is None:
+                idxs = range(len(lines))
+            for idx, line in zip(idxs, lines):
+                line.set_ydata(Z[:, idx])
+            # recompute the ax.dataLim
[…]

### 2019-09-26 mushi.py

@@ -42,6 +42,7 @@ class kSFS():
         else:
             self.n = n
         self.L = utils.C(self.n) @ utils.M(self.n, self.η)
+        self.bins = None
 
     def tmrca_cdf(self) -> np.ndarray:
         '''The cdf of the TMRCA of at each change point'''
@@ -280,36 +281,66 @@ class kSFS():
         μ.Z = np.exp(logZ)
         return μ, f_trajectory
 
-    def plot(self, i: int = None, μ: histories.μ = None, prf_quantiles=False):
-        '''plot the SFS data and optionally the expected SFS given μ
+    def plot1(self, type: int, μ: histories.μ = None, prf_quantiles=False):
+        '''plot the SFS data of one type and optionally its expectation and
+        confidence bands
 
-        i: component i of kSFS (default first)
+        type: mutation type to plot (i.e. 'TCC>TTC')
         μ: mutation intensity history (optional)
         prf_quantiles: if True show 95% marginal intervals using the Poisson
                        random field
         '''
-        if i is None:
-            i = 0
+        i = self.mutation_types.get_loc(type)
         if μ is not None:
             z = μ.Z[:, i]
             ξ = self.L @ z
             if self.bins is not None:
                 for bin in self.bins:
-                    plt.axvline(bin, c='k', ls=':', alpha=0.2)
-            plt.plot(range(1, self.n), ξ, 'r--', label=r'$\xi$')
+                    plt.axvline(bin, c='C3', ls=':', alpha=0.2)
+            plt.plot(range(1, self.n), ξ, c='C1', ls='--', label=r'$\xi$')
             if prf_quantiles:
                 ξ_lower = poisson.ppf(.025, ξ)
                 ξ_upper = poisson.ppf(.975, ξ)
                 plt.fill_between(range(1, self.n),
                                  ξ_lower, ξ_upper,
-                                 facecolor='r', alpha=0.25,
+                                 facecolor='C1', alpha=0.25,
                                  label='inner 95%\nquantile')
         plt.plot(range(1, len(self.X) + 1), self.X[:, i],
-                 'k.', alpha=.25, label=r'data')
-        plt.xlabel('$b$')
-        plt.ylabel(r'$ξ_b$')
+                 c='C0', ls='', marker='.', alpha=.25, label=r'data')
+        plt.xlabel('sample frequency')
+        plt.ylabel(r'number of variants')
         plt.xscale('log')
[…]

### 2019-09-26 mushi.py
test

@@ -180,7 +180,7 @@ class kSFS():
 
         # Accelerated proximal gradient descent: our cost function decomposes
         # as f = g + h, where g is differentiable and h is not.
-        # We transform logZ to restrict to positive solutions
+        # We transform Z to restrict to positive solutions
         # https://people.eecs.berkeley.edu/~elghaoui/Teaching/EE227A/lecture18.pdf
         # some matrices we'll need for the first difference penalties
         D = (np.eye(self.η.m, k=0) - np.eye(self.η.m, k=-1))
@@ -189,9 +189,8 @@ class kSFS():
         D1 = W @ D  # 1st difference matrix
         D2 = D1.T @ D1  # square of 1st difference matrix (Laplacian)
 
-        def g(logZ, grad=False):
+        def g(Z, grad=False):
             '''differentiable piece of cost'''
-            Z = np.exp(logZ)
             if grad:
                 loss, grad_loss = loss_func(Z, grad=True)
             else:
@@ -202,13 +201,11 @@ class kSFS():
             if grad:
                 grad_g = grad_loss + λ_tv * (1 - α_tv) * D2 @ Z \
                                      + λ_r * (1 - α_r) * Z
-                grad_g_log = grad_g * Z  # change of variables
-                return g, grad_g_log
+                return g, grad_g
             return g
 
-        def h(logZ):
+        def h(Z):
             '''nondifferentiable piece of cost'''
-            Z = np.exp(logZ)
             σ = np.linalg.svd(Z, compute_uv=False)
             if hard:
                 rank_penalty = np.linalg.norm(σ, 0)
@@ -218,16 +215,16 @@ class kSFS():
             return λ_tv * α_tv * np.abs(D1 @ Z).sum() \
                 + λ_r * α_r * rank_penalty
 
-        def f(logZ):
+        def f(Z):
             '''cost'''
-            return g(logZ) + h(logZ)
+            return g(Z) + h(Z)
 
         # initialize using constant μ history MLE
         μ = self.constant_μ_MLE()
-        logZ = np.log(μ.Z)  # current iterate
-        logQ = np.log(μ.Z)  # momentum iterate
+        Z = μ.Z  # current iterate
+        Q = Z    # momentum iterate
         # initial loss
-        f_trajectory = [f(logZ)]
+        f_trajectory = [f(Z)]
         # initialize step size
         s0 = 1  # max step size
         s = s0  # current step size
[…]

### 2019-09-30 histories.py
plot with mutation type strings instead of column indices

@@ -65,14 +65,14 @@ class History():
         else:
             return True
 
-    def plot(self, idxs=None, **kwargs) -> List[mpl.lines.Line2D]:
+    def plot(self, types=None, **kwargs) -> List[mpl.lines.Line2D]:
         '''plot the history
 
-        idxs: indices of value column(s) to plot (optional)
         kwargs: key word arguments passed to plt.step
         '''
         t = np.concatenate((np.array([0]), self.change_points))
-        if idxs is not None:
+        if types is not None:
+            idxs = [self.mutation_types.get_loc(type) for type in types]
             vals = self.vals[:, idxs]
         else:
             vals = self.vals
@@ -144,16 +144,18 @@ class μ(History):
         self.mutation_types = pd.Index(self.mutation_types,
                                        name='mutation type')
 
-    def plot(self, idxs=None, normed=False,
+    def plot(self, types=None, normed=False,
              **kwargs) -> List[mpl.lines.Line2D]:
         '''
         normed: flag to normalize to relative mutation intensity
         '''
-        lines = super().plot(idxs=idxs, **kwargs)
+        lines = super().plot(types=types, **kwargs)
         if normed:
             Z = self.Z / self.Z.sum(1, keepdims=True)
-            if idxs is None:
+            if types is None:
                 idxs = range(len(lines))
+            else:
+                idxs = [self.mutation_types.get_loc(type) for type in types]
             for idx, line in zip(idxs, lines):
                 line.set_ydata(Z[:, idx])
             # recompute the ax.dataLim

### 2019-09-30 mushi.py

@@ -281,7 +281,7 @@ class kSFS():
         μ.Z = np.exp(logZ)
         return μ, f_trajectory
 
-    def plot1(self, type: int, μ: histories.μ = None, prf_quantiles=False):
+    def plot1(self, type, μ: histories.μ = None, prf_quantiles=False):
         '''plot the SFS data of one type and optionally its expectation and
         confidence bands
 
@@ -313,7 +313,7 @@ class kSFS():
         plt.yscale('symlog')
         plt.tight_layout()
 
-    def plot(self, idxs: List[int] = None, normed: bool = False,
+    def plot(self, type=None, normed: bool = False,
              μ: histories.μ = None, **kwargs) -> None:
         '''
         normed: flag to normalize to relative mutation intensity
@@ -326,15 +326,20 @@ class kSFS():
                 Ξ = Ξ / Ξ.sum(1, keepdims=True)
             plt.ylabel('mutation type fraction')
         else:
-            plt.ylabel('$\\mu(t)$')
-        if idxs is not None:
-            X = X[:, idxs]
+            X = self.X
+            plt.ylabel('number of variants')
+        if type is not None:
+            i = self.mutation_types.get_loc(type)
+            X = X[:, i]
             if μ is not None:
-                Ξ = Ξ[:, idxs]
+                Ξ = Ξ[:, i]
 
         if μ is not None:
             plt.plot(range(1, self.n), X, ls='', marker='.', **kwargs)
-            plt.plot(range(1, self.n), Ξ, **kwargs)
+            line_kwargs = kwargs
+            if 'label' in line_kwargs:
+                del line_kwargs['label']
+            plt.plot(range(1, self.n), Ξ, **line_kwargs)
         else:
             plt.plot(range(1, self.n), X, **kwargs)
         plt.xlabel('sample frequency')
@@ -369,5 +374,4 @@ class kSFS():
         g = sns.clustermap(df, row_cluster=False, metric='correlation',
                            cbar_kws={'label': cbar_label}, **kwargs)
         g.ax_heatmap.set_yscale('symlog')
-        plt.tight_layout()
         return g

### 2019-09-30 mushi.py
no more logZ transform, instead `np.clip` and print warning

@@ -77,11 +77,20 @@ class kSFS():
         self.X = poisson.rvs(self.L @ μ.Z)
         self.mutation_types = μ.mutation_types
 
-    def constant_μ_MLE(self) -> histories.μ:
-        '''gives the MLE for a constant μ history'''
+    def constant_μ_MLE(self, mask: np.ndarray = None) -> histories.μ:
+        '''gives the MLE for a constant μ history
+
+        mask: array of bools, with True indicating exclusion of that frequency
+        '''
         if self.X is None:
             raise ValueError('use simulate() to generate data first')
-        z0 = self.X.sum(axis=0) / np.sum(self.L)
+        if mask is not None:
+            L = self.L[~mask, :]
+            X = self.X[~mask, :]
+        else:
+            L = self.L
+            X = self.X
+        z0 = X.sum(axis=0) / np.sum(L)
         return histories.μ(self.η.change_points,
                            z0[np.newaxis, :] * np.ones((self.η.m, 1)),
                            mutation_types=self.mutation_types.values)
@@ -93,7 +102,7 @@ class kSFS():
                 tol: np.float64 = 1e-4, fit='prf',
                 bins: np.ndarray = None,
                 hard=False,
-                exclude_singletons: bool = False) -> histories.μ:
+                mask: np.ndarray = None) -> histories.μ:
         '''return inferred μ history given the sfs and η history
 
         λ_tv: fused LASSO regularization strength
@@ -107,7 +116,7 @@ class kSFS():
              Kullback-Leibler divergence, 'lsq' for least-squares
         bins: SFS frequency bins
         hard: hard Vs soft singular value thresholding
-        exclude_singletons: flag to exclude singleton frequency variants
+        mask: array of bools, with True indicating exclusion of that frequency
         '''
         if self.X is None:
             raise ValueError('use simulate() to generate data first')
@@ -115,6 +124,8 @@ class kSFS():
         assert λ_r >= 0, 'λ_r must be nonnegative'
         assert 0 <= α_tv <= 1, 'α_tv must be in the unit interval'
         assert 0 <= α_r <= 1, 'α_r must be in the unit interval'
+        if mask is not None:
+            assert len(mask) == self.X.shape[0], 'mask must have n-1 elements'
         self.bins = bins
         if bins is not None:
             bin_idxs = np.digitize(np.arange(self.n - 1), bins=bins)
@@ -133,9 +144,9 @@ class kSFS():
             self.X = X_binned
[…]

### 2019-09-30 manuscript/appendix.tex
minor eqn fix

@@
 is a linear functional of $\mu$ and a nonlinear functional of $\eta$.
-We recover \ref{eqn:transform} by defining the operator $\mathcal{L}_{n,\eta}$ such that $\mathcal{L}_{n,\eta}\mu \equiv C \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt$.
+We recover \ref{eqn:transform} by defining the operator $\mathcal{L}_{n,\eta}$ such that $\left(\mathcal{L}_{n,\eta}\mu\right)_j \equiv C \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt$.
 

### 2019-10-19 utils.py
`np.longdouble` to avoid loss-of-significance for larger n

@@ -12,17 +12,17 @@ def C(n: int) -> np.ndarray:
 
     n: number of sampled haplotypes
     '''
-    W1 = np.zeros((n - 1, n - 1))
-    W2 = np.zeros((n - 1, n - 1))
-    b = np.arange(1, n - 1 + 1)
+    W1 = np.zeros((n - 1, n - 1), dtype=np.longdouble)
+    W2 = np.zeros((n - 1, n - 1), dtype=np.longdouble)
+    b = np.arange(1, n - 1 + 1, dtype=np.longdouble)
     # j = 2
     W1[:, 0] = 6 / (n + 1)
     W2[:, 0] = 0
     # j = 3
     W1[:, 1] = 10 * (5 * n - 6 * b - 4) / (n + 2) / (n + 1)
     W2[:, 1] = (20 * (n - 2)) / (n+1) / (n+2)
-    for j in range(2, n - 1):
-        col = j - 2
+    for col in range(n - 3):
+        j = np.longdouble(col + 2)
         # procedurally generated by Zeilberger's algorithm in Mathematica
         W1[:, col + 2] = -((-((-1 + j)*(1 + j)**2*(3 + 2*j)*(j - n)*(4 + 2*j - 2*b*j + j**2 - b*j**2 + 4*n + 2*j*n + j**2*n)*W1[:, col]) - (-1 + 2*j)*(3 + 2*j)*(-4*j - 12*b*j - 4*b**2*j - 6*j**2 - 12*b*j**2 - 2*b**2*j**2 - 4*j**3 + 4*b**2*j**3 - 2*j**4 + 2*b**2*j**4 + 4*n + 2*j*n - 6*b*j*n + j**2*n - 9*b*j**2*n - 2*j**3*n - 6*b*j**3*n - j**4*n - 3*b*j**4*n + 4*n**2 + 6*j*n**2 + 7*j**2*n**2 + 2*j**3*n**2 + j**4*n**2)*W1[:, col + 1])/(j**2*(2 + j)*(-1 + 2*j)*(1 + j + n)*(3 + b + j**2 - b*j**2 + 3*n + j**2*n)))
[…]

### 2019-10-21 utils.py
eta fitting with `mushi.kSFS.infer_η`

@@ -5,7 +5,6 @@ import numpy as np
 from scipy.special import binom
 from scipy.stats import poisson
 
-import histories
 
 def C(n: int) -> np.ndarray:
     '''The C matrix defined in the text
@@ -29,14 +28,12 @@ def C(n: int) -> np.ndarray:
 
     return W1 - W2
 
-
-def M(n: int, η: histories.η) -> np.ndarray:
+def M(n: int, t: np.ndarray, y: np.ndarray) -> np.ndarray:
     '''The M matrix defined in the text
 
-    n: number of sampled haplotypes
-    η: demographic history
+    t: time grid, starting at zero and ending at np.inf
+    y: population size in each epoch
     '''
-    t, y = η.arrays()
     # epoch durations
     s = np.diff(t)
     u = np.exp(-s / y, dtype=np.longdouble)

### 2019-10-22 1KG/masked_size.py
masked genome size

@@ -1,0 +1,36 @@
+#! /usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from ancestor import Ancestor
+from collections import Counter
+
+
+def main():
+    """
+    usage: python masked_size.py -h
+    """
+    import argparse
+
+    parser = argparse.ArgumentParser(description='print masked chromosome '
+                                                 'size')
+    parser.add_argument('anc_aln_file', type=str, default=None,
+                        help='path to ancestral alignment fasta (one '
+                             'chromosome)')
+    parser.add_argument('mask_file', type=str, default=None,
+                        help='path to bed file mask')
+
+    args = parser.parse_args()
+
+    ancestor = Ancestor(args.anc_aln_file)
+    counts = Counter(ancestor.anc_aln)
+
+    for line in open(args.mask_file, 'r'):
+        chr, start, end = line.rstrip().split('\t')
+        assert chr == ancestor.chr
+        for char in ancestor.anc_aln[int(start):int(end)]:
+            counts[char] -= 1
+    print(counts['A'] + counts['C'] + counts['G'] + counts['T'])
+
+
+if __name__ == '__main__':
+    main()

### 2019-10-22 mushi.py
rename reg param for eta

@@ -92,16 +92,16 @@ class kSFS():
         self.mutation_types = self.μ.mutation_types
 
     def infer_η(self, change_points: np.array = None, fit='prf',
-                λ_tv: np.float64 = 0,
-                μ_0: np.float64 = 1,
+                λ: np.float64 = 0,
+                μ_0: np.float64 = None,
                 mask: np.ndarray = None, **kwargs) -> OptimizeResult:
         '''infer η. Either self.μ is not None or change_points is not None. If
         the latter, fit using the total (row-summed) SFS and a unit constant
         total mutation rate history
 
         change_points: epoch change points (times)
-        λ_tv: fused LASSO regularization strength
-        μ_0: total mutation rate
+        λ: fused LASSO regularization strength
+        μ_0: total mutation rate, if self.μ is None
         fit: loss function, 'prf' for Poisson random field, 'kl' for
              Kullback-Leibler divergence, 'lsq' for least-squares
         mask: array of bools, with True indicating exclusion of that frequency
@@ -109,7 +109,7 @@ class kSFS():
         '''
         if self.X is None:
             raise ValueError('use simulate() to generate data first')
-        assert λ_tv >= 0, 'λ_tv must be nonnegative'
+        assert λ >= 0, 'λ must be nonnegative'
         assert (self.μ is None) != (change_points is None)
         if mask is not None:
             assert len(mask) == self.X.shape[0], 'mask must have n-1 elements'
@@ -157,7 +157,7 @@ class kSFS():
 
         def f(logΛ):
             y = 1 / np.exp(logΛ)
-            return loss_func(y) + (λ_tv / 2) * (np.diff(y) ** 2).sum()
+            return loss_func(y) + (λ / 2) * (np.diff(y) ** 2).sum()
 
         result = minimize(f, np.log(Λ),
                           **kwargs)

### 2019-11-14 mushi.py
params to clip low and high

@@ -437,9 +437,11 @@ def main():
                                 npts)
 
     # mask sites
-    clip = config.getint('loss', 'clip', fallback=None)
-    if clip:
-        mask = np.array([False if (0 <= i <= n - clip)
+    clip_high = config.getint('loss', 'clip_high', fallback=None)
+    clip_low = config.getint('loss', 'clip_high', fallback=None)
+    if clip_high or clip_low:
+        assert clip_high and clip_low
+        mask = np.array([False if (clip_low <= i <= n - clip_high)
                          else True
                          for i in range(n - 1)])
     else:

### 2019-11-14 mushi.py
another cfg parse bug, doh

@@ -437,8 +437,8 @@ def main():
                                 npts)
 
     # mask sites
+    clip_low = config.getint('loss', 'clip_low', fallback=None)
     clip_high = config.getint('loss', 'clip_high', fallback=None)
-    clip_low = config.getint('loss', 'clip_high', fallback=None)
     if clip_high or clip_low:
         assert clip_high and clip_low
         mask = np.array([False if (clip_low <= i <= n - clip_high)

### 2019-11-16 mushi.py
oops

@@ -440,7 +440,7 @@ def main():
     clip_low = config.getint('loss', 'clip_low', fallback=None)
     clip_high = config.getint('loss', 'clip_high', fallback=None)
     if clip_high or clip_low:
-        assert clip_high and clip_low
+        assert clip_high is not None and clip_low is not None
         mask = np.array([False if (clip_low <= i <= n - clip_high)
                          else True
                          for i in range(n - 1)])

### 2019-11-16 mushi.py
update `mushi.clustermap()`

@@ -371,18 +371,14 @@ class kSFS():
             plt.legend()
         plt.tight_layout()
 
-    def clustermap(self, linthresh=1, **kwargs):
+    def clustermap(self, fit: bool = False, **kwargs):
         u"""clustermap with mixed linear-log scale color bar
 
-        μ: inferred mutation spectrum history, χ^2 values are shown if not None
-        linthresh: the range within which the plot is linear (when μ = None)
+        fit: if True and self.μ is not None, show χ^2 fit
         kwargs: additional keyword arguments passed to pd.clustermap
         """
-        if self.μ is None:
-            Z = self.X / self.X.sum(axis=1, keepdims=True)
-            Z = Z / Z.mean(0, keepdims=True)
-            cbar_label = 'mutation type\nenrichment'
-        else:
+        if fit:
+            assert self.μ is None
             Ξ = self.L @ self.μ.Z
             Z = (self.X - Ξ) ** 2 / Ξ
             cbar_label = '$\\chi^2$'
@@ -390,10 +386,14 @@ class kSFS():
             p = chi2(np.prod(Z.shape)).sf(χ2_total)
             print(f'χ\N{SUPERSCRIPT TWO} goodness of fit {χ2_total}, '
                   f'p = {p}')
+        else:
+            Z = self.X / self.X.sum(axis=1, keepdims=True)
+            Z = Z / Z.mean(0, keepdims=True)
+            cbar_label = 'mutation type\nenrichment'
         df = pd.DataFrame(data=Z, index=pd.Index(range(1, self.n),
                                                  name='sample frequency'),
                           columns=self.mutation_types)
-        g = sns.clustermap(df, row_cluster=False, metric='correlation',
+        g = sns.clustermap(df, row_cluster=False,
                            cbar_kws={'label': cbar_label}, **kwargs)
         g.ax_heatmap.set_yscale('symlog')
         return g
@@ -521,7 +521,7 @@ def main():
     plt.subplot(223)
     ksfs.plot(normed=True, alpha=0.5)
     plt.subplot(224)
-    ksfs.μ.plot(normed=True, alpha=0.5)
+    ksfs.μ.plot(normed=False, alpha=0.5)
     plt.savefig(f'{args.outbase}.fit.png')
 
     # pickle the final ksfs (which contains all the inferred history info)

### 2019-11-16 manuscript/main.tex
my preferred notation for regularization parameters
acknowledgements
update title and abstract

@@
 
-\title{Coalescent inference of mutation spectrum histories from sample frequency spectra}
+\title{Population genetics with evolving mutation processes: inference of mutation spectrum histories from modern genomic variation}
 
 \author{
-W.S. DeWitt$^{1,4}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1}$\\
+W.S. DeWitt$^{1,4}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1,4}$\\
 {\small
@@
 
-Single nucleotide variant (SNV) frequencies partitioned according to local nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
-The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
-We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the sample frequency spectrum resolved by $k$-mer nucleotide context ($k$-SFS).
-We formulate inference of mutation spectrum histories from the $k$-SFS as a linear inverse problem.
+Models in evolutionary genetics typically assume that mutation rate is constant over time and between populations and closely related species.
+However, recent work casts doubt on this assumption in human and ape populations, and reveals that mutation is a complex and dynamic process.
+Whether arising from variation in replication fidelity, life history, or environmental exposures, mutation rate evolution can be accompanied by changes to the \emph{mutation spectrum}: the mutation rate in different local nucleotide contexts.
+We extend theoretical tools based on Kingman's coalescent to accommodate a richly parameterized mutation process, varying in time and in spectrum.
+We infer human mutation spectrum histories from patterns of modern genomic diversity, allowing us to reconstruct trajectories of divergence between populations, track a transient mutation spectrum perturbation through multiple populations, and characterize how mutation spectrum histories are structured by local nucleotide context.
+Mutation rate and effective population size together determine the strength of genetic drift shaping genomic variation.
+We introduce fast non-parametric joint inference of mutation spectrum history and demographic history.
 
@@
   =
-  \lambda_\mathrm{spline} 
-  \frac{1}{2} \left\|\Delta Z\right\|_2^2
+  \frac{  \lambda_\mathrm{spline} }{2} \left\|\Delta Z\right\|_2^2
   +
-  \lambda_\mathrm{rank} \| Z \|_*.
+  \lambda_\mathrm{rank} \| Z \|_*
+  +
+   \frac{\lambda_\mathrm{ridge}}{2}  \| Z \|_\mathrm{F}^2
+   .
 \end{equation}
@@
   =
-  \lambda_\mathrm{spline} 
-  \frac{1}{2} \left\|\Delta Z\right\|_2^2
+  \frac{\lambda_\mathrm{spline}}{2} \left\|\Delta Z\right\|_2^2
   +
-  \lambda_\mathrm{rank} \, \mathrm{rank}(Z).
+  \lambda_\mathrm{rank} \, \mathrm{rank}(Z)
+  +
+  \frac{\lambda_\mathrm{ridge}}{2} \| Z \|_\mathrm{F}^2
+  .
 \end{equation}
@@
   \lambda_\mathrm{TV}
-  \left(
-    \alpha_\mathrm{TV}\left\|\Delta Z\right\|_1 + \frac{(1-\alpha_\mathrm{TV})}{2}\left\|\Delta Z\right\|_2^2
-  \right) .
+  \left\|\Delta Z\right\|_1
+  +
+  \frac{\lambda_\mathrm{spline}}{2} \left\|\Delta Z\right\|_2^2
+  +
+  \frac{\lambda_\mathrm{ridge}}{2} \| Z \|_\mathrm{F}^2
+  .
 \end{equation}
@@
 and the rank term allows us to recover low-rank solutions.
-The function $R_\mathrm{indep}$ is convex; 
-therefore we have a convex optimization problem.
+The function $R_\mathrm{indep}$ is convex.
 Note also that the columns of $Z$ are uncoupled,
@@
 
+Erick Matsen, Peter Ralph, Andy Kern, Harris lab, Jeff Spence, Aleksandr Aravkin
+
 \bibliographystyle{plainnat}

### 2019-11-21 1KG/qsub_scons.sh
1kg update

@@ -1,26 +1,28 @@
 #!usr/bin/bash
 
-cd /net/harris/vol1/project/mushi/1KG
-source activate mushi
+# cd /net/harris/vol1/project/mushi/1KG
+# source activate mushi
 
 # linux
-n_jobs=`nproc --all`
+# n_jobs=`nproc --all`
 # n_jobs=1
 # mac
-# n_jobs=`sysctl -n hw.physicalcpu`
+n_jobs=`sysctl -n hw.physicalcpu`
 
-bcfs="/net/harris/vol1/data/phase3_1000genomes/bcfs"
-ancs="/net/harris/vol1/data/phase3_1000genomes/human_ancestor_GRCh37_e59"
-sample_pops="/net/harris/vol1/data/phase3_1000genomes/integrated_call_samples_v3.20130502.ALL.panel"
-k="5"
-mutation_rate="1.1e-8"
-phastcons="/net/harris/vol1/data/phase3_1000genomes/phastConsElements100way.txt.gz"
-repeats="/net/harris/vol1/data/phase3_1000genomes/nestedRepeats.txt.gz"
+bcfs="../data/phase3_1000genomes/bcfs"
+ancs="../data/human_ancestor_GRCh37_e59"
+sample_pops="../data/phase3_1000genomes/integrated_call_samples_v3.20130502.ALL.panel"
+k="3"
+phastcons="../data/phastConsElements100way.txt.gz"
+repeats="../data/nestedRepeats.txt.gz"
 
-outdir="/net/harris/vol1/project/mushi/1KG/scons_output_5mer"
+outdir="scons_output_noenhancers"
+
+track='--track=../data/20161129_enhancer_states.txt.4567.bed'
+comp='--track_complement'
 
 # no_exec="--no-exec"
 
-cmd="scons --bcfs=${bcfs} --ancs=${ancs} --mutation_rate=${mutation_rate} --kmer=${k} --phastcons=${phastcons} --repeats=${repeats} --jobs=${n_jobs} --outdir=${outdir} ${debug} ${no_exec} --sample_pops=${sample_pops} ${gl}"
+cmd="scons ${track} ${comp} --bcfs=${bcfs} --ancs=${ancs} --kmer=${k} --phastcons=${phastcons} --repeats=${repeats} --jobs=${n_jobs} --outdir=${outdir} ${debug} ${no_exec} --sample_pops=${sample_pops} ${gl}"
 echo $cmd
 $cmd

### 2019-11-21 histories.py
plot mu relative to ancestral rate

@@ -191,13 +191,12 @@ class μ(History):
         kwargs: additional keyword arguments passed to pd.clustermap
         """
         t = np.concatenate((np.array([0]), self.change_points))
-        Z = self.Z / self.Z.sum(1, keepdims=True)
-        Z = Z / Z.mean(0, keepdims=True)
+        Z = self.Z / self.Z[-1, :]
         df = pd.DataFrame(data=Z, index=pd.Index(t, name='$t$'),
                           columns=self.mutation_types)
         g = sns.clustermap(df, row_cluster=False, center=1,
                            metric='correlation',
-                           cbar_kws={'label': 'relative mutation intensity'},
+                           cbar_kws={'label': 'mutation intensity wrt ancestral rate'},
                            **kwargs)
         # g.ax_heatmap.set_yscale('symlog')
         plt.tight_layout()

### 2019-11-21 mushi.py
plot mu relative to ancestral rate

@@ -316,6 +316,9 @@ class kSFS():
 
     def plot_total(self):
         """plot the total SFS"""
+        x = self.X.sum(1, keepdims=True)
+        plt.plot(range(1, len(x) + 1), x,
+                 c='C0', ls='', marker='.', alpha=.25, label=r'data')
         if self.η is not None:
             if self.μ is not None:
                 z = self.μ.Z.sum(1)
@@ -329,9 +332,6 @@ class kSFS():
                              ξ_lower, ξ_upper,
                              facecolor='C1', alpha=0.25,
                              label='inner 95%\nquantile')
-        x = self.X.sum(1, keepdims=True)
-        plt.plot(range(1, len(x) + 1), x,
-                 c='C0', ls='', marker='.', alpha=.25, label=r'data')
         plt.xlabel('sample frequency')
         plt.ylabel(r'number of variants')
         plt.xscale('log')

### 2019-11-21 manuscript/appendix.tex
committing intro notes in case train crashes

@@
 so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \eqref{eqn:W1} and \eqref{eqn:W2} are definite sums over geometric terms.
-Zeilberger's algorithm, which finds polynomial recurrences for definite sums of hypergeometric terms \citep{petkovvsek1996b, paule1995mathematica}, can thus be used to yield the following procedurally generated second-order recursions in $j$:
+Zeilberger's algorithm, which finds polynomial recurrences for definite sums of hypergeometric terms \cite{petkovvsek1996b, paule1995mathematica}, can thus be used to yield the following procedurally generated second-order recursions in $j$:
 %WD need to wrap this eqn, as it runs will beyond the margin. Also make an appendix at some point for these

### 2019-11-21 manuscript/main.tex
committing intro notes in case train crashes

@@
 \usepackage{fullpage}
-\usepackage[round,semicolon]{natbib}
+\usepackage[numbers]{natbib}
 \usepackage{amssymb}
@@
 
-\title{Population genetics with evolving mutation processes: inference of mutation spectrum histories from modern genomic variation}
+% WD: it's kinda silly to focus on unphased genomes then run on 1KG
+\title{Population genetic reconstruction of mutation spectrum evolution from unphased genomes}
 
@@
 We extend theoretical tools based on Kingman's coalescent to accommodate a richly parameterized mutation process, varying in time and in spectrum.
-We infer human mutation spectrum histories from patterns of modern genomic diversity, allowing us to reconstruct trajectories of divergence between populations, track a transient mutation spectrum perturbation through multiple populations, and characterize how mutation spectrum histories are structured by local nucleotide context.
+% WD: revise below when we settle on what results will be in the paper
+We infer human mutation spectrum histories from patterns of modern genomic diversity, allowing us to reconstruct trajectories of mutation spectrum divergence between populations, track a transient mutation spectrum perturbation through multiple populations, and characterize how mutation spectrum histories are structured by local nucleotide context.
 Mutation rate and effective population size together determine the strength of genetic drift shaping genomic variation.
-We introduce fast non-parametric joint inference of mutation spectrum history and demographic history.
+We introduce fast nonparametric joint inference of mutation spectrum history and demographic history from unphased genomes.
 
@@
 \section*{Introduction}\label{sec:intro}
+
+% WD: should this be less humany?
+Population genetics has proved to be a powerful lens on human history\cite{Harris2017-cd}, revealing a complex past of migration\cite{?}, replacement\cite{?}, and admixture among human groups and extinct homonids\cite{?}, and dramatic population size changes during and after range expansions\cite{?}.
+Germline mutation events in the history of a population give rise to variant alleles, and demographic histories are recorded in patterns of segregation within and between current-day populations.
+Genetic drift is shared when populations mix, and it is private during periods of isolation.
+The strength of drift is universally influenced by the population size history, thus population genetic methods for demographic inference---expressed in the framework of Kimura's diffusion or Kingman's coalescent---attempt to recover the effective population size history $N(t)$.
+
+Mutation drives the evolution of populations and species, and mutation rate evolution has a storied history in the theoretical population genetics literature.
+Haldane developed equilibrium theory for alleles in mutation-selection balance, and used this to provided the first principled estimate of the human mutation rate by studying hemophilia incidence\cite{Haldane2004-ov, Nachman2004-gh}.
+Kimura  \cite{Kimura1967-ku}
+
+Storied history of theoretical treatments:
+- Kimura: \cite{Kimura1967-ku}
+- drift barrier hyp of \cite{Lynch2016-ff}
+
+The process of germline mutation is the writing mechanism that records signatures of demographic events in genomes, so its influence on modern genome variation is similar in importance to the demographic histories themselves.
+Demographic inference methods can model complex population splits, migration, and admixture, in addition to modeling $N(t)$ as a piecewise constant (or exponential) function of time into the past.
+Mutation, however, receives a relatively impoverished treatment: usually a single mutation rate parameter $\mu$ is understood to apply at all loci in all individuals at all times, and may be regarded as a nuisance parameter needed for time calibration of models whose natural dimensionless time is measured in generations.
+
+Growing evidence indicates that germline mutation is a dynamic process on both interspecific and population time scales, and a complex function of polymorphic replication machinery, life history, mutagenic exposures, and genomic content.
+Indeed, mutation rates among great apes have declined along the lineage leading to humans---a phenomenon termed the \emph{homonoid slowdown}\cite{Goodman1985-xc, Scally2012-rb}---, showing that mutation rate evolution between species distorts phylogenetic time calibration.
+Children of older parents receive more germline mutations, especially so for older fathers, and this is thought to arise from replicative errors in spermatogenesis.
+This \emph{parental age effect} \cite{} indicates that life history changes can influence the population mutation process.
+% need an example of mutagenic/spontaneous germline mutation
+
+\begin{enumerate}
+\item genomic regions (agarwal, and other Molly)
+\item \texttt{C>T} at methylated CpGs due to deamination \cite{?}.
+\item gbGC. We might want to talk about isochores here, but I've yet to find a coherent definition.
+\item mutation spectra and pulse \cite{Harris2015-wi, Harris2017-fw}
+\item Jed on context and regions \cite{Carlson2018-rc}
+\item for some of these, it is unknown if driven by replicative Vs spontaneous mutation \cite{Gao2016-qp}, and that baboon paper touches on this (Molly)
+\end{enumerate}
+
+emerging picture of mutation rate accented by spectra
+
+rare variant spectra \cite{Mathieson2017-rc} (maybe discuss normalization issues with this paper)
+
+1KG cell line artifact \cite{Anderson-Trocme2019-fy} (they also use strict mask)
+
+% this has to come after we into triplet spectra
+Although methods exist for inferring locus-wise mutation rates \cite{Bhaskar2015-bd, Nelson2012-fr}, these require very large study sizes, and do not accommodate evolution of the rate or spectrum of the mutational process per se.
+
+New methods that approximate the ancestral recombination graphs...
+
+We have to be very clear about how what we’re offering is different than what can be done with Relate \cite{Speidel2019-ox}.
+\begin{enumerate}
+\item We can deal with unphased genomes.
+\item Our method allows the total mutation rate to be variable whereas Relate fixes it to a constant, and does posthoc counting.
+\item I do not quite understand how they count mutations in a given epoch, since they are not mapped to specific times, only to branches.
+Unless mu(t) is constant, uniform sampling on the branch is not consistent
+\item How does pulse timing compare?
+\end{enumerate}
+
+deserving more rich descriptions in population genetic models
+
+% background:
+% review on demographic inference (diff and coal methods), or popgen theory in general (Kimura's diffusion or Kingman's coal), and how they all assume a constant mu
+
+% review on germline rate var: https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.2000744
+
+% Single nucleotide variant (SNV) frequencies partitioned according to local nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
+
+% moving beyond descriptive comparisons based on standing variation, to put mutation on the same footing as demographic history: "in the light of evolution" Dobzhansky
+
+% models
+
+% there is a family of models, simple panmictic coal, complex diff, psmc
+
+% The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
+
+% gloom and doom perils about N(t), Meyers, Song, other Song, Gravel, etc., nonlinearity, nonconvexity, then contrast this with our discovery of a more tractable problem
+% This has a truly excellent intro/summary of the issues with the SFS: https://www.sciencedirect.com/science/article/pii/S004058091730148X?via%3Dihub
+
+% "Stairway plot" that Kern recommended, is sort non-parametric: X. Liu, Y.-X. Fu, Exploring population size changes using SNP frequency spectra. Nat. Genet. 47, 555–559 (2015).
+
+% We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the sample frequency spectrum resolved by $k$-mer nucleotide context ($k$-SFS).
+% We formulate inference of mutation spectrum histories from the $k$-SFS as a linear inverse problem.
+
+% as the first method like this, we implement in the simpler among models in the family, single population.
+Beginning of fruitful future work to bring model realism to popgen inf, via much more rich descriptions of the mutational process.
+
+% sims with realistic demo hist and various spectrum hists, we can recover
+
+% github
+
+% TODO
+% - ancient african substructure buzz, bantu VS san, admixture signal?
+% - gbGC and isochores: https://www.annualreviews.org/doi/abs/10.1146/annurev-genom-082908-150001?casa_token=EImkIpqbayAAAAAA:Rxc8y0KLVRbZ6UfBTy3VxBRPdt3IZf7CAdVk57DfBUpvpSvuP4uHpDe9XHHCMHFBbYDxcZyFR8t0tA
 
@@
 
-%WD todo: inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
+% WD: todo: inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
 
-%WD todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
+% WD: todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
 
-Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
-\begin{enumerate}
-\item the traditional sample frequency \emph{spectrum}
-\item mutation \emph{spectrum} as in the nucleotide context of each mutation
-\item \emph{spectral} regularization for encouraging low rank solutions
-\end{enumerate}
+% Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
+% \begin{enumerate}
+% \item the traditional sample frequency \emph{spectrum}
+% \item mutation \emph{spectrum} as in the nucleotide context of each mutation
+% \item \emph{spectral} regularization for encouraging low rank solutions
+% \end{enumerate}
 
@@
 
-The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
+The setting for our modeling is Kingman's coalescent \cite{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
 In Appendix \ref{sec:appendix} we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
@@
 
-In previous work \citep{Kelley, cancer, etc}, mutation spectra have been parameterized by grouping variants into mutation types according to local $k$-mer nucleotide context, and the number of variants called in each mutation type is taken as a proxy integrating an individual's ancestral mutational input in each type.
+In previous work \cite{Kelley, cancer, etc}, mutation spectra have been parameterized by grouping variants into mutation types according to local $k$-mer nucleotide context, and the number of variants called in each mutation type is taken as a proxy integrating an individual's ancestral mutational input in each type.
 With $k$ odd and the variant registered in the central position there are $\kappa_k=2\times3\times4^{k-1}$ mutation types after collapsing by strand symmetry.
@@
 
-In linkage equilibrium the log likelihood function of the mutation spectrum history $Z$ given an observed $k$-SFS matrix $X$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
+In linkage equilibrium the log likelihood function of the mutation spectrum history $Z$ given an observed $k$-SFS matrix $X$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \cite{?}
 \[
@@
 
-%WD more inverse problem intro
+% WD: more inverse problem intro
 The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general,
 so many very different histories can be equally consistent with the data
-\citep{oscillation paper? Yun's other papers?}.
+\cite{oscillation paper? Yun's other papers?}.
 We deal with this problem using regularization, to enforce well-behaved histories.
@@
 The \texttt{mushi} software package is available at \url{https://github.com/harrispopgen/mushi}.
-%WD cite whatever packages we use, e.g. proxtv, msprime, stdpopsim, etc.
+% WD: cite whatever packages we use, e.g. jax, proxtv, msprime, stdpopsim, etc.
 

### 2019-12-03 manuscript/main.tex
non-identifiability of N(t) mu(t)

@@
 % \end{enumerate}
+
+\subsection*{Non-identifiability of demographic history and total mutation rate}
+
+As shown in appendix \ref{sec:appendix:forward:pcsws} eqn~\eqref{eqn:d2}, $\eta(t)$ and $\mu(t)$ are non-identifiable at all times.
+This is shown by introducing a change of variables that measures time in expected number of coalescent events since the present.
+This non-identifiability is similarly manifest by a change of variables to measure time in expected mutations \citetext{Peter Ralph, priv.\ comm., 2019}.
 

### 2019-12-03 manuscript/appendix.tex
non-identifiability of eta(t) and mu(t) for all t

@@
 
-%WD Kam describes this discretization as ad hoc, but I'm pretty sure this is your textbook Riemann sum with rectangular partitions.
-We could do fancier quadrature, but then we'll be less able to lean on previous work (i.e.
-Rosen et al.)
-
-Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
+% WD: did they borrow this change of variables from Meyers et al.?
+Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)), we introduce a change of variables that measures time in expected number of coalescent events since the present.
+Let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
 \begin{equation}

### 2019-12-03 manuscript/main.tex
non-identifiability of eta(t) and mu(t) for all t

--- 2019-11-21 manuscript/main.tex
@@
 % \end{enumerate}
+
+\subsection*{Non-identifiability of demographic history and total mutation rate}
+
+As shown in appendix \ref{sec:appendix:forward:pcsws} eqn~\eqref{eqn:d2}, effective population size $\eta(t)$ and the total mutation rate $\mu(t)$ are non-identifiable at all times.
+This is demonstrated by introducing a change of variables that measures time in expected number of coalescent events since the present.
+This non-identifiability is similarly manifest by a change of variables to measure time in expected mutations \citetext{Peter Ralph, priv.\ comm.}.
+This can be understood intuitively: an excess of variants of a given frequency can be explained by an historical bottleneck (which expands coalescent lines in the bottleneck time interval), but equally well by a period of increased total mutation rate with no demographic change.
 

### 2019-12-04 manuscript/main.tex
intro draft

@@
 
-% WD: it's kinda silly to focus on unphased genomes then run on 1KG
-\title{Population genetic reconstruction of mutation spectrum evolution from unphased genomes}
+% TODO
+% - ancient african substructure buzz, bantu VS san, admixture signal?
+% - gbGC and isochores: https://www.annualreviews.org/doi/abs/10.1146/annurev-genom-082908-150001?casa_token=EImkIpqbayAAAAAA:Rxc8y0KLVRbZ6UfBTy3VxBRPdt3IZf7CAdVk57DfBUpvpSvuP4uHpDe9XHHCMHFBbYDxcZyFR8t0tA
+% - thinking on identifiability chat with Stilianos at UO: although we allow the different mutation rates (for the different triplet contexts) to vary in time, it might be reasonable to enforce that the total mutation rate (added over the the triplets) is still rather constant in time.
+Without this, it’s true that we can not identify 1.\ stretching of the coalescent tree in some time interval due to a population crash from 2.\ a period of increased total mutation rate.
+We could enforce this with a strong spline penalty, or strictly, using a prox to a simplex.
+% - inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
+% - tidy up the math with some lemmas and theorems
 
+\title{Population genetic inference of mutation spectrum evolution}
+
+% WD: the affil numbers are a little out of order, but it's nice to concisely bundle all the UW depts
 \author{
@@
 % WD: revise below when we settle on what results will be in the paper
+% WD: may also want to emphasize the hypothesis generating aspect of having the mutation spectrum history
 We infer human mutation spectrum histories from patterns of modern genomic diversity, allowing us to reconstruct trajectories of mutation spectrum divergence between populations, track a transient mutation spectrum perturbation through multiple populations, and characterize how mutation spectrum histories are structured by local nucleotide context.
 Mutation rate and effective population size together determine the strength of genetic drift shaping genomic variation.
-We introduce fast nonparametric joint inference of mutation spectrum history and demographic history from unphased genomes.
+We introduce fast nonparametric inference of mutation spectrum history and demographic history from unphased genomes.
 
@@
 % WD: should this be less humany?
-Population genetics has proved to be a powerful lens on human history\cite{Harris2017-cd}, revealing a complex past of migration\cite{?}, replacement\cite{?}, and admixture among human groups and extinct homonids\cite{?}, and dramatic population size changes during and after range expansions\cite{?}.
+Population genetics has proved to be a powerful lens on human history \cite{Harris2017-cd}, revealing a complex past of migration \cite{?}, replacement \cite{?}, and admixture among human groups and extinct homonids \cite{?}, and dramatic population size changes during and after range expansions \cite{?}.
 Germline mutation events in the history of a population give rise to variant alleles, and demographic histories are recorded in patterns of segregation within and between current-day populations.
 Genetic drift is shared when populations mix, and it is private during periods of isolation.
-The strength of drift is universally influenced by the population size history, thus population genetic methods for demographic inference---expressed in the framework of Kimura's diffusion or Kingman's coalescent---attempt to recover the effective population size history $N(t)$.
+The strength of drift is universally influenced by the population size history, thus population genetic methods for demographic inference---expressed in the framework of Kimura's diffusion or Kingman's coalescent---attempt to recover the effective population size history $N(t)$ (where time is measured in years or generations before the present).
 
+% WD: These examples seem a bit arbitrary, and maybe it's odd to put the more recent Lynch work next to classics like Kimura and Haldane in this paragraph
 Mutation drives the evolution of populations and species, and mutation rate evolution has a storied history in the theoretical population genetics literature.
-Haldane developed equilibrium theory for alleles in mutation-selection balance, and used this to provided the first principled estimate of the human mutation rate by studying hemophilia incidence\cite{Haldane2004-ov, Nachman2004-gh}.
-Kimura  \cite{Kimura1967-ku}
-
-Storied history of theoretical treatments:
-- Kimura: \cite{Kimura1967-ku}
-- drift barrier hyp of \cite{Lynch2016-ff}
+Haldane developed equilibrium theory for alleles in mutation-selection balance, and used this to provided the first principled estimate of the human mutation rate by studying hemophilia incidence \cite{Haldane2004-ov, Nachman2004-gh}.
+Kimura later considered how genetic modifiers of the mutation rate---\emph{mutator alleles}---tune the optimal mutation rate to balance adaptive response to environmental changes, increasing genetic load, and the tendency of mutators to escape their deleterious consequences via recombination \cite{Kimura1967-ku}.
+The \emph{drift-barrier hypothesis} of Lynch et al.\ considers the effect of genetic drift on mutation rate optima, finding that effective population size determines a limit for evolutionary optimization toward high replication fidelity, as the efficiency of selection against mutator alleles increases with $N$ \cite{Lynch2016-ff}.
 
@@
 Demographic inference methods can model complex population splits, migration, and admixture, in addition to modeling $N(t)$ as a piecewise constant (or exponential) function of time into the past.
-Mutation, however, receives a relatively impoverished treatment: usually a single mutation rate parameter $\mu$ is understood to apply at all loci in all individuals at all times, and may be regarded as a nuisance parameter needed for time calibration of models whose natural dimensionless time is measured in generations.
+Mutation, however, receives a relatively impoverished treatment: usually a single mutation rate parameter $\mu$ is understood to apply at all loci in all individuals at all times, and may be regarded as a nuisance parameter needed for time calibration of models whose natural dimensionless time is measured in Wright-Fisher generations.
 
 Growing evidence indicates that germline mutation is a dynamic process on both interspecific and population time scales, and a complex function of polymorphic replication machinery, life history, mutagenic exposures, and genomic content.
-Indeed, mutation rates among great apes have declined along the lineage leading to humans---a phenomenon termed the \emph{homonoid slowdown}\cite{Goodman1985-xc, Scally2012-rb}---, showing that mutation rate evolution between species distorts phylogenetic time calibration.
-Children of older parents receive more germline mutations, especially so for older fathers, and this is thought to arise from replicative errors in spermatogenesis.
-This \emph{parental age effect} \cite{} indicates that life history changes can influence the population mutation process.
-% need an example of mutagenic/spontaneous germline mutation
+Mutation rates among great apes have declined along the lineage leading to humans---a phenomenon termed the \emph{homonoid slowdown} \cite{Goodman1985-xc, Scally2012-rb}---, showing that mutation rate evolution between species distorts phylogenetic time calibration.
+At the level of single generations, children of older parents receive more germline mutations, especially so from older fathers, likely arising from replicative errors in spermatogenesis that add $\sim 1$ additional expected mutation per year of paternal age.
+% WD: parental age or paternal age? Both are found in the lit
+This \emph{parental age effect} \cite{} shows that sex-specific life history traits can influence the population mutation process.
+% WD: need an example of mutagenic/spontaneous germline mutation
+% WD: for some of these, it is unknown if driven by replicative Vs spontaneous mutation \cite{Moorjani2016-fx, Gao2016-qp}, and that baboon paper gets at this (Molly)
 
-\begin{enumerate}
-\item genomic regions (agarwal, and other Molly)
-\item \texttt{C>T} at methylated CpGs due to deamination \cite{?}.
-\item gbGC. We might want to talk about isochores here, but I've yet to find a coherent definition.
-\item mutation spectra and pulse \cite{Harris2015-wi, Harris2017-fw}
-\item Jed on context and regions \cite{Carlson2018-rc}
-\item for some of these, it is unknown if driven by replicative Vs spontaneous mutation \cite{Gao2016-qp}, and that baboon paper touches on this (Molly)
-\end{enumerate}
+A complex and polymorphic mutation process also reveals itself in associations with genomic position and local nucleotide context.
+The rate of \texttt{C$\rightarrow$T} transitions is elevated at methylated CpGs due to spontaneous deamination \cite{Hwang2004-bq, Segurel2014-ss}.
+GC-biased gene conversion (gbGC) refers to the tendency of stronger-binding GC alleles to overwrite AT alleles during homologous recombination \cite{Galtier2007-eb,Duret2009-wn}.
+This biased non-Mendelian segregation pattern is tantamount to selection for \emph{weak to strong} mutations.
+% WD: we might want to talk about isochores here, but I've yet to find a coherent definition
 
-emerging picture of mutation rate accented by spectra
+It is difficult to disentangle past changes in mutation rate from past changes in effective population size.
+However, the myriad mechanisms that drive the mutation process suggest that its evolution may be revealed by studying the \emph{mutation spectrum}: the mutation rate in different local nucleotide contexts.
+Indeed, Harris et al.\ \cite{Harris2015-wi, Harris2017-fw} used a triplet context parameterization borrowed from the somatic mutation signature literature \cite{Alexandrov2013-oe, Helleday2014-xb}, and counted single nucleotide variants in each triplet mutation type as an estimate of the time-averaged mutational input from each individual's history.
+Human triplet spectra distinctly clustered according to continental ancestry group, and evidence of historical pulses in mutation activity (or suppression of repair) was found in the distribution of allele frequencies in certain mutation types.
+Mathiesen et al.\ studied similar mutational signatures in triplet spectra of rare human variants \cite{Mathieson2017-rc}, and clarified alternative non-mutational hypotheses for their origin, including population differences in demography, patterns of selection, recombination, or recombination-associated processes such as gene conversion.
+Rare variants in large human cohorts serve as a proxy for recent de novo mutations, and reveal mutation signatures of replication timing, recombination, and sex differences in repair processes \cite{Carlson2018-rc, Agarwal2019-dn}.
 
-rare variant spectra \cite{Mathieson2017-rc} (maybe discuss normalization issues with this paper)
+Emerging from the literature is a picture of a mutation process evolving within and between populations, anchored to genomic features, and accented by spectra of local nucleotide context.
+If probabilistic models of population genetic processes are to keep pace with phenomenology, mutation deserves much more rich description in these models.
+% WD: remember, "nonparametric" is just a screwy way of saying "the most parameterized"
+In this paper, we build on classical theoretical tools to introduce fast nonparametric inference of population-level \emph{mutation spectrum history} (MUSH)---the relative mutation rate history in different local nucleotide contexts---alongside inference of demographic history from unphased genomes.
+% WD: The thing I came for: the wreck and not the story of the wreck. The thing itself and not the myth.
+In doing so, we promote mutation spectrum evolution from phenomenological statistics on modern variation to an historical object to be inferred, on the same footing as demography.
 
-1KG cell line artifact \cite{Anderson-Trocme2019-fy} (they also use strict mask)
+The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
+Here, we extend a coalescent framework for demographic inference from the SFS to accommodate inference of the MUSH from an SFS that is resolved into different local $k$-mer nucleotide context---a more rich summary statistic that we call the $k$-SFS, where e.g.\ $k=3$ for triplet context.
+We show that inference of the MUSH is a linear inverse problem.
+While challenging to invert due to ill-posedness, this stands in stark contrast to the deep problems of non-identifiability that afflict demographic inference from the SFS \cite{Myers2008-jp, Bhaskar2014-fu, Terhorst2015-xt, Baharian2018-np, Rosen2018-bb} (although an estimate of the latter is also required).
+We develop regularization methods, grounded in recent advances in proximal gradient optimization, to seek non-negative solutions with smoothness (under various norms) and sparsity in mutation signatures.
+We infer both demographic history and MUSH nonparametrically using regularization, rather than selecting a model with a small number of constant or exponential epochs.
+Our open-source software, termed \texttt{mushi} (mutation spectrum history inference), is available at \url{https://github.com/harrispopgen/mushi}.
 
-% this has to come after we into triplet spectra
-Although methods exist for inferring locus-wise mutation rates \cite{Bhaskar2015-bd, Nelson2012-fr}, these require very large study sizes, and do not accommodate evolution of the rate or spectrum of the mutational process per se.
+An inferred MUSH is a richly-parameterized object to which various biological questions about evolution of the mutation process may be addressed, both intrapopulation (patterns within a single MUSH) and interpopulation (comparisons between MUSHs).
+% WD: hold off on summary of results until we settle on what to show
+We use \texttt{mushi} to\dots
 
-New methods that approximate the ancestral recombination graphs...
-
-We have to be very clear about how what we’re offering is different than what can be done with Relate \cite{Speidel2019-ox}.
-\begin{enumerate}
-\item We can deal with unphased genomes.
-\item Our method allows the total mutation rate to be variable whereas Relate fixes it to a constant, and does posthoc counting.
-\item I do not quite understand how they count mutations in a given epoch, since they are not mapped to specific times, only to branches.
-Unless mu(t) is constant, uniform sampling on the branch is not consistent
-\item How does pulse timing compare?
-\end{enumerate}
-
-deserving more rich descriptions in population genetic models
-
-% background:
-% review on demographic inference (diff and coal methods), or popgen theory in general (Kimura's diffusion or Kingman's coal), and how they all assume a constant mu
-
-% review on germline rate var: https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.2000744
-
-% Single nucleotide variant (SNV) frequencies partitioned according to local nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
-
-% moving beyond descriptive comparisons based on standing variation, to put mutation on the same footing as demographic history: "in the light of evolution" Dobzhansky
-
-% models
-
-% there is a family of models, simple panmictic coal, complex diff, psmc
-
-% The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
-
-% gloom and doom perils about N(t), Meyers, Song, other Song, Gravel, etc., nonlinearity, nonconvexity, then contrast this with our discovery of a more tractable problem
-% This has a truly excellent intro/summary of the issues with the SFS: https://www.sciencedirect.com/science/article/pii/S004058091730148X?via%3Dihub
-
-% "Stairway plot" that Kern recommended, is sort non-parametric: X. Liu, Y.-X. Fu, Exploring population size changes using SNP frequency spectra. Nat. Genet. 47, 555–559 (2015).
-
-% We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the sample frequency spectrum resolved by $k$-mer nucleotide context ($k$-SFS).
-% We formulate inference of mutation spectrum histories from the $k$-SFS as a linear inverse problem.
-
-% as the first method like this, we implement in the simpler among models in the family, single population.
-Beginning of fruitful future work to bring model realism to popgen inf, via much more rich descriptions of the mutational process.
-
-% sims with realistic demo hist and various spectrum hists, we can recover
-
-% github
-
-% TODO
-% - ancient african substructure buzz, bantu VS san, admixture signal?
-% - gbGC and isochores: https://www.annualreviews.org/doi/abs/10.1146/annurev-genom-082908-150001?casa_token=EImkIpqbayAAAAAA:Rxc8y0KLVRbZ6UfBTy3VxBRPdt3IZf7CAdVk57DfBUpvpSvuP4uHpDe9XHHCMHFBbYDxcZyFR8t0tA
-
-\cite{Harris2017-fw} showed triplet SNV spectrum differences between human groups and between great apes, and used a coalescent simulation approach to fit the timing of a pulse of \texttt{T\underline{C}C>T\underline{T}C} mutations in Europeans.
-We develop coalescent theory for the dependence of the expected $k$-SFS on both demographic history (effective population size as a function of time into the past) \emph{and} mutation spectrum history (mutation intensities as a function of time into the past for each mutation type).
-This is similar in spirit to Kelley's simulation-based approach (which assumed the European demography of \cite{Tennessen2012-dq}), but is much faster, more flexibly parameterized.
-
-Some references for nonidentifiability of demographic history from the SFS: \cite{Baharian2018-np, Bhaskar2014-fu, Terhorst2015-xt, Myers2008-jp}.
-It might be good to consider identifiability of mutation intensity history from the SFS conditioned on demographic history.
-
-% WD: todo: inverse problem intro, forward Vs inverse problems, and ill-posedness. Point out \eta is nonlinear, but \mu is linear
-
-% WD: todo: the scope of this paper is \mu inference conditioned on \eta. There are plenty of fancy methods one can use to infer \eta
-
-% Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
-% \begin{enumerate}
-% \item the traditional sample frequency \emph{spectrum}
-% \item mutation \emph{spectrum} as in the nucleotide context of each mutation
-% \item \emph{spectral} regularization for encouraging low rank solutions
-% \end{enumerate}
-
-\subsection*{Non-identifiability of demographic history and total mutation rate}
+% WD: need to decide where to put this
+% -------
+\section*{Non-identifiability of demographic history and total mutation rate}
 
@@
 This can be understood intuitively: an excess of variants of a given frequency can be explained by an historical bottleneck (which expands coalescent lines in the bottleneck time interval), but equally well by a period of increased total mutation rate with no demographic change.
+% -------
 
@@
 
-Joint inference of $\eta$ and $\mu$.
+% WD: Peter's references: https://www.bibsonomy.org/user/peter.ralph/mutation_spectrum
+
+1KG cell line artifact \cite{Anderson-Trocme2019-fy} (they also use strict mask)
+
+Although methods exist for inferring locus-wise mutation rates \cite{Bhaskar2015-bd, Nelson2012-fr}, these require very large study sizes, and do not accommodate evolution of the rate or spectrum of the mutation process per se.
+
+New methods that approximate the ancestral recombination graphs...
+
+We have to be very clear about how what we’re offering is different than what can be done with Relate \cite{Speidel2019-ox}.
+\begin{enumerate}
+\item We can deal with unphased genomes.
+\item Our method allows the total mutation rate to be variable whereas Relate fixes it to a constant, and does posthoc counting.
+\item I do not quite understand how they count mutations in a given epoch, since they are not mapped to specific times, only to branches.
+Unless mu(t) is constant, uniform sampling on the branch is not consistent
+\item How does pulse timing compare?
+\end{enumerate}
 
@@
 
-Erick Matsen, Peter Ralph, Andy Kern, Harris lab, Jeff Spence, Aleksandr Aravkin
+Peter Ralph and Andy Kern and the Kern/Ralph colab, Harris lab, Stilianos Louca, Erick Matsen, Aleksandr Aravkin, Jeff Spence
 
